### 2018-02-22 slowtests/slowtests-pima.Rmd
bug fix for gbm

---
title: "Pima  Tests"
author: "Brandon M. Greenwell"
date: "July 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
# Load required packages
library(ggplot2)
library(grid)
library(pdp)

# Set knitr chunk options
knitr::opts_chunk$set(echo = TRUE)

# Function to run tests for a fitted model object using the Pima Indians 
# diabetes data
pimaIndiansTest <- function(object, ...) {

  # Training data
  pima <- na.omit(pima)
  trn <- subset(pima, select = -diabetes)

  # Partial dependence plot for lstat
  pd1 <- partial(object, pred.var = "glucose", train = trn, ...)

  # Partial dependence plot for lstat and rm
  pd2 <- partial(object, pred.var = c("glucose", "age"), chull = TRUE,
                 train = trn, ...)

  # Individual conditional expectation curves for age
  pd3 <- partial(object, pred.var = "glucose", ice = TRUE, train = trn,
                 ...)

  # Centered individual conditional expectation curves for lstat
  pd4 <- partial(object, pred.var = "glucose", ice = TRUE, center = TRUE,
                 train = trn, ...)

  # Display plots
  grid.arrange(
    autoplot(pd1, rug = TRUE, train = pima, main = "PDP for glucose"),
    autoplot(pd2, main = "PDP for glucose and age"),
    autoplot(pd3, rug = TRUE, train = pima,
             main = "ICE curves for glucose", alpha = 0.25),
    autoplot(pd4, rug = TRUE, train = pima,
             main = "c-ICE curves for glucose", alpha = 0.25),
    ncol = 2,
    top = textGrob(deparse(substitute(object)),
                   gp = gpar(fontsize = 20, font = 3))
  )

}
```

# Discriminant analysis

## Package: `MASS`

```{r}
pima.lda <- MASS::lda(diabetes ~ . ^ 2, data = na.omit(pima))
pimaIndiansTest(pima.lda)
pimaIndiansTest(pima.lda, prob = TRUE)
```

```{r}
pima.qda <- MASS::qda(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.qda)
pimaIndiansTest(pima.qda, prob = TRUE)
```

## Package: `mda`

```{r}
pima.mda <- mda::mda(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.mda)
pimaIndiansTest(pima.mda, prob = TRUE)
```

```{r}
pima.fda <- mda::fda(diabetes ~ ., data = na.omit(pima), method = mda::mars,
                     degree = 2)
pimaIndiansTest(pima.fda)
pimaIndiansTest(pima.fda, prob = TRUE)
```

# Decision trees

## Package: `rpart`

```{r}
pima.rpart <- rpart::rpart(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.rpart)
pimaIndiansTest(pima.rpart, prob = TRUE)
```

## Package: `C50`
```{r}
set.seed(101)
pima.C5.0 <- C50::C5.0(diabetes ~ ., data = na.omit(pima), trials = 100)
pimaIndiansTest(pima.C5.0)
pimaIndiansTest(pima.C5.0, prob = TRUE)
```

## Package: `party

```{r}
pima.ctree <- party::ctree(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.ctree)
pimaIndiansTest(pima.ctree, prob = TRUE)
```

## Package: `partykit

```{r}
pima.ctree2 <- partykit::ctree(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.ctree2)
pimaIndiansTest(pima.ctree2, prob = TRUE)
```

# Bagging

## Package: `adabag`

```{r}
set.seed(101)
pima.bagging <- adabag::bagging(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.bagging, quantiles = TRUE)
pimaIndiansTest(pima.bagging, quantiles = TRUE, prob = TRUE)
```

## Package: `ipred

```{r}
set.seed(101)
pima.ipred <- ipred::bagging(diabetes ~ ., data = na.omit(pima), nbagg = 500)
pimaIndiansTest(pima.ipred, quantiles = TRUE)
pimaIndiansTest(pima.ipred, quantiles = TRUE, prob = TRUE)
```

# Random forests

## Package: `randomForest

```{r}
set.seed(101)
pima.rf <- randomForest::randomForest(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.rf)
pimaIndiansTest(pima.rf, prob = TRUE)
```

## Package: `party

```{r}
set.seed(101)
pima.crf <- party::cforest(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.crf, quantiles = TRUE)
pimaIndiansTest(pima.crf, quantiles = TRUE, prob = TRUE)
```

## Package: `partykit`
```{r}
set.seed(101)
pima.crf2 <- partykit::cforest(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.crf2, quantiles = TRUE)
pimaIndiansTest(pima.crf2, quantiles = TRUE, prob = TRUE)
```

## Package: `ranger`

```{r}
set.seed(101)
pima.ranger <- ranger::ranger(diabetes ~ ., data = na.omit(pima), 
                              probability = TRUE)
pimaIndiansTest(pima.ranger)
pimaIndiansTest(pima.ranger, prob = TRUE)
```

# Boosting

## Package: `adabag`

```{r}
set.seed(101)
pima.boosting <- adabag::boosting(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.boosting, quantiles = TRUE)
pimaIndiansTest(pima.boosting, quantiles = TRUE, prob = TRUE)
```

## Package: `gbm`

```{r}
set.seed(101)
pima.gbm <- gbm::gbm(ifelse(diabetes == "neg", 1, 0) ~ ., 
                     data = na.omit(pima),
                     distribution = "bernoulli",
                     n.trees = 5000,
                     interaction.depth = 3,
                     shrinkage = 0.001,
                     # cv.folds = 5,
                     verbose = FALSE)
best.iter <- gbm::gbm.perf(pima.gbm, method = "OOB", plot.it = FALSE)
pimaIndiansTest(pima.gbm, n.trees = best.iter)
pimaIndiansTest(pima.gbm, n.trees = best.iter, prob = TRUE)
```

## Package: `xgboost`
```{r}
set.seed(101)
pima.xgb <- xgboost::xgboost(
  data = data.matrix(subset(pima, select = -diabetes)),
  label = unclass(pima$diabetes) - 1, objective = "binary:logistic",
  nrounds = 100, max_depth = 3, eta = 0.1, gamma = 0, colsample_bytree = 0.8, 
  min_child_weight = 1, subsample = 0.7, verbose = 0
)
pimaIndiansTest(pima.xgb)
pimaIndiansTest(pima.xgb, prob = TRUE)
```

# Neural networks

## Package: `nnet`
```{r}
set.seed(101)
pima.nnet <- nnet::nnet(diabetes ~ ., data = na.omit(pima), size = 10, 
                        decay = 0.1, maxit = 500, trace = FALSE)
pimaIndiansTest(pima.nnet)
pimaIndiansTest(pima.nnet, prob = TRUE)
```

# Support vector machines

## Package: `e1071`

```{r}
pima.svm <- e1071::svm(diabetes ~ ., data = na.omit(pima), type = "C-classification", 
                       probability = TRUE)
pimaIndiansTest(pima.svm)
pimaIndiansTest(pima.svm, prob = TRUE)
```

## Package: `kernlab`

```{r}
pima.ksvm <- kernlab::ksvm(diabetes ~ ., data = na.omit(pima), type = "C-svc", 
                           prob.model = TRUE)
pimaIndiansTest(pima.ksvm)
pimaIndiansTest(pima.ksvm, prob = TRUE)
```

# Multinomial models

## Package: `nnet`

```{r}
set.seed(101)
pima.multinom <- nnet::multinom(diabetes ~ . ^ 2, data = na.omit(pima), 
                                trace = FALSE)
pimaIndiansTest(pima.multinom)
pimaIndiansTest(pima.multinom, prob = TRUE)
```

### 2018-02-22 R/getParPredCls.R

@@ -105,7 +105,7 @@ getParClsLogit.gbm <- function(object, newdata, which.class, ...) {
   # It seems that when the response has more than two levels, predict.gbm
   # returns an array. When the response is binary, a vector with predictions
   # for the positive class is returned returned.
-  if (ncol(pr) == 1) {
+  if (NCOL(pr) == 1) {
     mean(multiClassLogit(cbind(pr, 1 - pr), which.class = which.class),
          na.rm = TRUE)
   } else {
@@ -122,7 +122,7 @@ getParClsProb.gbm <- function(object, newdata, which.class, ...) {
   # It seems that when the response has more than two levels, predict.gbm
   # returns an array. When the response is binary, a vector with predictions
   # for the positive class is returned returned.
-  if (ncol(pr) == 1) {
+  if (NCOL(pr) == 1) {
     mean(cbind(pr, 1 - pr)[, which.class], na.rm = TRUE)
   } else {
     mean(pr[, which.class, 1], na.rm = TRUE)

### 2018-02-22 tools/pdp-logo-img.R

@@ -1,6 +1,9 @@
 # Load required packages
+library(ggimage)
 library(ggplot2)
+library(grid)
 library(pdp)
+library(png)
 library(randomForest)
 
 # Fit a random forest to the Boston housing data
@@ -9,16 +12,16 @@ set.seed(101)  # for reproducibility
 boston.rf <- randomForest(cmedv ~ ., data = boston)
 
 # Partial dependence of cmedv on lstat and rm
-pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = FALSE,
+pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE,
               progress = "text")
 
 # Create image for logo (i.e., remove axis labels, etc.)
-pdp <- autoplot(pd, contour = TRUE, contour.color = "black") +
-  annotate("text", label = "pdp",
-           x = mean(range(boston$lstat)),
-           y = mean(range(boston$rm)),
-           size = 15,
-           color = "white") +
+p <- autoplot(pd, contour = TRUE, contour.color = "black") +
+  # annotate("text", label = "pdp",
+  #          x = mean(range(boston$lstat)),
+  #          y = mean(range(boston$rm)),
+  #          size = 15,
+  #          color = "white") +
   theme(axis.line = element_blank(),
         axis.text.x = element_blank(),
         axis.text.y = element_blank(),
@@ -31,16 +34,50 @@ pdp <- autoplot(pd, contour = TRUE, contour.color = "black") +
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         plot.background = element_blank())
-pdp  # text will look smaller
+print(p)  # text will look smaller
 
-# Save logo image
-ggsave("/home/w108bmg/Desktop/Dropbox/devel/pdp/tools/pdp-logo-img.png",
-       plot = pdp,
-       device = NULL,
-       path = NULL,
-       scale = 1,
-       width = 1.7321,
-       height = 2,
-       units = "in",
-       dpi = 300,
-       limitsize = TRUE)
+# # Save logo image
+# ggsave("/home/w108bmg/Desktop/Dropbox/devel/pdp/tools/pdp-logo-img.png",
+#        plot = pdp,
+#        device = NULL,
+#        path = NULL,
+#        scale = 1,
[…]

### 2018-05-17 README.Rmd
update logo; fix CRAN badge

---
output:
  md_document:
    variant: markdown_github
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/README-"
)
```

# pdp <img src="tools/pdp-logo.png" align="right" width="120" height="139" />

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/pdp)](https://cran.r-project.org/package=pdp)
[![Build Status](https://travis-ci.org/bgreenwell/pdp.svg?branch=master)](https://travis-ci.org/bgreenwell/pdp)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/bgreenwell/pdp?branch=master&svg=true)](https://ci.appveyor.com/project/bgreenwell/pdp)
[![Coverage Status](https://img.shields.io/codecov/c/github/bgreenwell/pdp.svg)](https://codecov.io/github/bgreenwell/pdp?branch=master)
[![Downloads](http://cranlogs.r-pkg.org/badges/pdp)](http://cranlogs.r-pkg.org/badges/pdp)
[![Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/pdp)](http://cranlogs.r-pkg.org/badges/grand-total/pdp)

## Overview

Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be more easily understood. These plots are especially useful in explaining the output from black box models. The `pdp` package offers a general framework for constructing partial dependence plots for various types of fitted models in R.

A detailed introduction to `pdp` has been published in The R Journal: "pdp: An R Package for Constructing Partial Dependence Plots", https://journal.r-project.org/archive/2017/RJ-2017-016/index.html. You can track development at https://github.com/bgreenwell/pdp. To report bugs or issues, contact the main author directly or submit them to https://github.com/bgreenwell/pdp/issues.

As of right now, `pdp` exports four functions:

* `partial` - compute partial dependence functions (i.e., objects of class `"partial"`) from various fitted model objects;
* `plotPartial"` - plot partial dependence functions (i.e., objects of class `"partial"`) using `lattice` graphics;
* `autoplot` - plot partial dependence functions (i.e., objects of class `"partial"`) using `ggplot2` graphics;
* `topPredictors` extract most "important" predictors from various types of fitted models.

## Installation

The `pdp` package is [currently listed on CRAN](https://CRAN.R-project.org/package=pdp) and can easily be installed:
```r
  # Install from CRAN (recommended)
  install.packages("pdp")

  # Alternatively, install the development version from GitHub
  devtools::install_github("bgreenwell/pdp")
```

## Random forest example

As a first example, we'll fit a random forest to the famous Boston housing data included with the package (see `?boston` for details). In fact the original motivation for this package was to be able to compute two-predictor partial dependence plots from random forest models in R. 
```{r example-rf, fig.width=8, fig.height=4, message=FALSE}
# Fit a random forest to the Boston housing data
library(randomForest)  # install.packages("randomForest")
data (boston)  # load the boston housing data
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston)

# Partial dependence of cmedv on lstat and rm
library(pdp)
pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE)
head(pd)  # print first 6 rows

# Lattice version
p1 <- plotPartial(pd, main = "lattice version")

# ggplot2 version
library(ggplot2)
p2 <- autoplot(pd, contour = TRUE, main = "ggplot2 version", 
               legend.title = "Partial\ndependence")

# Show both plots in one figure
grid.arrange(p1, p2, ncol = 2)
```

Next, we'll fit a classification model to the Pima Indians Diabetes data.

## Support vector machine (SVM) example

As a second example, we'll fit an SVM to the Pima Indians diabetes data included with the package (see `?pima` for details). Note that for some fitted model objects (e.g., `"ksvm"` objects) it is necessary to supply the original training data via the `train` argument in the call to `partial`.

```{r example-svm, fig.width=8, fig.height=4, message=FALSE}
# Fit an SVM to the Pima Indians diabetes data
library(kernlab)  # install.packages("kernlab")
data (pima)  # load the Pima Indians diabetes data
pima.svm <- ksvm(diabetes ~ ., data = pima, type = "C-svc", kernel = "rbfdot",
                 C = 0.5, prob.model = TRUE)

# Partial dependence of diabetes test result on glucose (default is logit scale)
pd.glucose <- partial(pima.svm, pred.var = "glucose", train = pima)

# Partial dependence of diabetes test result on glucose (probability scale)
pd.glucose.prob <- partial(pima.svm, pred.var = "glucose", prob = TRUE, 
                           train = pima)

# Show both plots in one figure
grid.arrange(autoplot(pd.glucose, main = "Logit scale"), 
             autoplot(pd.glucose.prob, main = "Probability scale"), 
             ncol = 2)
```

### 2018-05-17 tools/pdp-logo-img.R

@@ -11,73 +11,71 @@ data (boston)  # load the boston housing data
 set.seed(101)  # for reproducibility
 boston.rf <- randomForest(cmedv ~ ., data = boston)
 
+# Function to rescale vector to be between a and b
+rescale <- function(x, a, b) {
+  ((x - min(x)) / (max(x) - min(x))) * (b - a) + a
+}
+
 # Partial dependence of cmedv on lstat and rm
-pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE,
-              progress = "text")
+pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = FALSE,
+              progress = "text", grid.resolution = 100)
 
-# Create image for logo (i.e., remove axis labels, etc.)
-p <- autoplot(pd, contour = TRUE, contour.color = "black") +
-  # annotate("text", label = "pdp",
-  #          x = mean(range(boston$lstat)),
-  #          y = mean(range(boston$rm)),
-  #          size = 15,
-  #          color = "white") +
-  theme(axis.line = element_blank(),
-        axis.text.x = element_blank(),
-        axis.text.y = element_blank(),
-        axis.ticks = element_blank(),
-        axis.title.x = element_blank(),
-        axis.title.y = element_blank(),
-        legend.position = "none",
-        panel.background = element_blank(),
-        panel.border = element_blank(),
-        panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(),
-        plot.background = element_blank())
-print(p)  # text will look smaller
-
-# # Save logo image
-# ggsave("/home/w108bmg/Desktop/Dropbox/devel/pdp/tools/pdp-logo-img.png",
-#        plot = pdp,
-#        device = NULL,
-#        path = NULL,
-#        scale = 1,
-#        width = 1.7321,
-#        height = 2,
-#        units = "in",
-#        dpi = 300,
-#        limitsize = TRUE)
-#
-# # Load vip image
-# img <- rasterGrob(readPNG("tools/pdp-logo-img.png"),
-#                   interpolate = TRUE, width = 1)
-
-# Hexagon data
+# Boundaries of the hexagon
 hex <- data.frame(x = 1.35 * 1 * c(-sqrt(3) / 2, 0, rep(sqrt(3) / 2, 2), 0,
[…]

### 2018-05-23 tools/pdp-logo-img.R
fix typo

@@ -1,5 +1,5 @@
 # Load required packages
-library(ggimage)
+library(earth)
 library(ggplot2)
 library(grid)
 library(pdp)
@@ -10,6 +10,11 @@ library(randomForest)
 data (boston)  # load the boston housing data
 set.seed(101)  # for reproducibility
 boston.rf <- randomForest(cmedv ~ ., data = boston)
+boston.earth <- earth(cmedv ~ ., data = boston, degree = 2,
+                      pmethod = "exhaustive", nfold = 5, ncross = 5)
+
+# Print MARS model coefficients
+coef(boston.earth)
 
 # Function to rescale vector to be between a and b
 rescale <- function(x, a, b) {
@@ -63,7 +68,7 @@ make_pdp_sticker <- function(option) {
 
 # Plot range of different logos
 logos <- lapply(LETTERS[1L:5L], make_pdp_sticker)
-png("tools/pdp-logol.png", width = 900, height = 500, bg = "transparent",
+png("tools/pdp-logos.png", width = 900, height = 500, bg = "transparent",
     type = "cairo-png")
 grid.arrange(grobs = logos, ncol = 3)
 dev.off()

### 2018-05-28 R/autoplot.R
lots pf changes

@@ -43,14 +43,14 @@
 #' on to the modelling function defined by \code{smooth.method}.
 #'
 #' @param contour Logical indicating whether or not to add contour lines to the
-#' level plot. Only used when \code{levelplot = TRUE}. Default is \code{FALSE}.
+#' level plot.
 #'
 #' @param contour.color Character string specifying the color to use for the
 #' contour lines when \code{contour = TRUE}. Default is \code{"white"}.
 #'
-#' @param palette If a string, will use that named palette. If a number, will
-#' index into the list of palettes of appropriate type. Default is
-#' \code{"Spectral"}.
+#' @param palette Character string indicating the colormap option to use. Five
+#' options are available: "viridis" (the default), "magma", "inferno", "plasma",
+#' and "cividis".
 #'
 #' @param train Data frame containing the original training data. Only required
 #' if \code{rug = TRUE} or \code{chull = TRUE}.
@@ -112,15 +112,15 @@
 #'   ncol = 2
 #' )
 #' }
-autoplot.partial <- function(object, center = FALSE, plot.pdp = TRUE,
-                             pdp.color = "red", pdp.size = 1, pdp.linetype = 1,
-                             rug = FALSE, smooth = FALSE,
-                             smooth.method = "auto", smooth.formula = y ~ x,
-                             smooth.span = 0.75, smooth.method.args = list(),
-                             contour = FALSE, contour.color = "white",
-                             palette = "Spectral", train = NULL,
-                             xlab = NULL, ylab = NULL, main = NULL,
-                             legend.title = NULL, ...) {
+autoplot.partial <- function(
+  object, center = FALSE, plot.pdp = TRUE, pdp.color = "red", pdp.size = 1,
+  pdp.linetype = 1, rug = FALSE, smooth = FALSE, smooth.method = "auto",
+  smooth.formula = y ~ x, smooth.span = 0.75, smooth.method.args = list(),
+  contour = FALSE, contour.color = "white",
+  palette = c("viridis", "magma", "inferno", "plasma", "cividis"),
+  train = NULL, xlab = NULL, ylab = NULL, main = NULL, legend.title = "yhat",
+  ...
+) {
 
   # Determine if object contains an ID column (i.e., multiple curves)
   multi <- "yhat.id" %in% names(object)
@@ -147,6 +147,7 @@ autoplot.partial <- function(object, center = FALSE, plot.pdp = TRUE,
[…]

### 2018-05-28 R/partial.R

@@ -85,6 +85,13 @@
 #' function directly (\code{TRUE}). Default is \code{FALSE}. See
 #' \code{\link{plotPartial}} for plotting details.
 #'
+#' @param plot.engine Character string specifying which plotting engine to use
+#' whenever \code{plot = TRUE}. Options include \code{"lattice"} (defasult) or
+#' \code{"ggplot2"}.
+#'
+#' @param alpha Numeric values between 0 and 1 indicating the alpha-transparency
+#' to use for ICE anc-ICE curves. Default is 0.5.
+#'
 #' @param smooth Logical indicating whether or not to overlay a LOESS smooth.
 #' Default is \code{FALSE}.
 #'
@@ -93,6 +100,21 @@
 #' predictor distributions. This helps reduce the risk of interpreting the
 #' partial dependence plot outside the region of the data (i.e., extrapolating).
 #' Only used when \code{plot = TRUE}. Default is \code{FALSE}.
+#'
+#' @param contour Logical indicating whether or not to add contour lines to the
+#' level plot. Only used when \code{levelplot = TRUE}. Default is \code{FALSE}.
+#'
+#' @param contour.color Character string specifying the color to use for the
+#' contour lines when \code{contour = TRUE}. Default is \code{"white"}.
+#'
+#' @param palette Character string indicating the colormap option to use. Five
+#' options are available: "viridis" (the default), "magma", "inferno", "plasma",
+#' and "cividis".
+#'
+#' @param alpha Numeric value in \code{[0, 1]} specifying the opacity alpha (
+#' most useful when plotting ICE/c-ICE curves). Default is 1 (i.e., no
+#' transparency). In fact, this option only affects ICE/c-ICE curves and level
+#' plots.
 #'
 #' @param chull Logical indicating wether or not to restrict the values of the
 #' first two variables in \code{pred.var} to lie within the convex hull of their
@@ -244,18 +266,20 @@ partial <- function(object, ...) {
 
 
 #' @rdname partial
+#'
 #' @export
-partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
-                            grid.resolution = NULL, ice = FALSE, center = FALSE,
-                            quantiles = FALSE, probs = 1:9/10,
-                            trim.outliers = FALSE,
-                            type = c("auto", "regression", "classification"),
-                            inv.link = NULL,
[…]

### 2018-05-28 R/plotPartial.R

@@ -39,6 +39,17 @@
 #' @param contour Logical indicating whether or not to add contour lines to the
 #' level plot. Only used when \code{levelplot = TRUE}. Default is \code{FALSE}.
 #'
+#' @param contour.color Character string specifying the color to use for the
+#' contour lines when \code{contour = TRUE}. Default is \code{"white"}.
+#'
+#' @param palette Character string indicating the colormap option to use. Five
+#' options are available: "viridis" (the default), "magma", "inferno", "plasma",
+#' and "cividis".
+#'
+#' @param alpha Numeric value in \code{[0, 1]} specifying the opacity alpha (
+#' most useful when plotting ICE/c-ICE curves). Default is 1 (i.e., no
+#' transparency).
+#'
 #' @param number Integer specifying the number of conditional intervals to use
 #' for the continuous panel variables. See \code{\link[graphics]{co.intervals}}
 #' and \code{\link[lattice]{equal.count}} for further details.
@@ -106,34 +117,37 @@ plotPartial <- function(x, ...) {
 
 #' @rdname plotPartial
 #' @export
-plotPartial.ice <- function(x, center = FALSE, plot.pdp = TRUE,
-                            pdp.col = "red2", pdp.lwd = 2, pdp.lty = 1,
-                            rug = FALSE, train = NULL, ...) {
+plotPartial.ice <- function(
+  x, center = FALSE, plot.pdp = TRUE, pdp.col = "red2", pdp.lwd = 2,
+  pdp.lty = 1, rug = FALSE, train = NULL, alpha = 1, ...
+) {
   plotIceCurves(x, center = center, plot.pdp = plot.pdp, pdp.col = pdp.col,
                 pdp.lwd = pdp.lwd, pdp.lty = pdp.lty, rug = rug, train = train,
-                ...)
+                alpha = alpha, ...)
 }
 
 
 #' @rdname plotPartial
 #' @export
-plotPartial.cice <- function(x, plot.pdp = TRUE, pdp.col = "red2", pdp.lwd = 2,
-                             pdp.lty = 1, rug = FALSE, train = NULL, ...) {
+plotPartial.cice <- function(
+  x, plot.pdp = TRUE, pdp.col = "red2", pdp.lwd = 2, pdp.lty = 1, rug = FALSE,
+  train = NULL, alpha = 1, ...
+) {
[…]

### 2018-05-28 README.Rmd
update badge size
fix typo
update graphics

@@
 knitr::opts_chunk$set(
+  cache = TRUE,
   collapse = TRUE,
   comment = "#>",
+  fig.align = "center",
   fig.path = "tools/README-"
@@
 
-# pdp <img src="tools/pdp-logo.png" align="right" width="120" height="139" />
+# pdp <img src="tools/pdp-logo.png" align="right" width="130" height="150" />
 
@@
 
+You can also plot the output directly from `partial()` (although, this is typically not recommended; see the [above-mentioned paper](https://journal.r-project.org/archive/2017/RJ-2017-016/index.html) for details):
+```{r example-rf-2, fig.width=7, fig.height=5, message=FALSE}
+partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE, plot = TRUE,
+        plot.engine = "ggplot2", palette = "magma", alpha = 0.9,
+        contour = TRUE, contour.color = "white")
+```
+
 Next, we'll fit a classification model to the Pima Indians Diabetes data.
@@
 As a second example, we'll fit an SVM to the Pima Indians diabetes data included with the package (see `?pima` for details).
-Note that for some fitted model objects (e.g., `"ksvm"` objects) it is necessary to supply the original training data via the `train` argument in the call to `partial`.
+Note that for some fitted model objects (e.g., `"ksvm"` objects) it is necessary to supply the original training data via the `train` argument in the call to `partial()`.
 
-```{r example-svm, fig.width=8, fig.height=4, message=FALSE}
+```{r example-svm, fig.width=8, fig.height=4, message=TRUE}
 # Fit an SVM to the Pima Indians diabetes data

### 2018-08-10 README.Rmd
code cleanup

@@
   cache = TRUE,
+  cache.path = "tools/README-cache/",
   collapse = TRUE,
@@
   fig.align = "center",
-  fig.path = "tools/README-"
+  fig.path = "man/figures/README-"
 )
@@
 ```
+
+## Individual conditional expectation (ICE) plots
+
+PDPs can be misleading in the presence of substantial interactions [(Goldstein et al., 2015)](https://www.tandfonline.com/doi/full/10.1080/10618600.2014.907095).
+To overcome this issue Goldstein, Kapelner, Bleich, and Pitkin developed the concept of individual conditional expectation (ICE) plots—available in the [`ICEbox`](https://cran.r-project.org/package=ICEbox) package.
+ICE plots display the estimated relationship between the response and a predictor of interest for each observation.
+Consequently, the PDP for a predictor of interest can be obtained by averaging the corresponding ICE curves across all observations.
+Starting with `pdp` version 0.6.0, ICE and centered ICE (c-ICE) curves can be easily obtained using the `ice` and `center` options; c-ICE plots help visualize heterogeneity in the modeled relationship between observations.
+Derivative ICE (d-ICE) plots (**coming soon**) help to explore interaction effects.
+
+```{r example-svm-2, fig.width=8, fig.height=4}
+# Load required packages
+library(ggplot2)  # for additional themes
+
+# ICE and c-ICE curves of diabetes test result on glucose (default logit scale)
+ice.glucose <- partial(
+  pima.svm, pred.var = "glucose", train = pima, ice = TRUE
+)
+cice.glucose <- partial(
+  pima.svm, pred.var = "glucose", train = pima, ice = TRUE, center = TRUE
+)
+
+# Show both plots in one figure
+grid.arrange(ncol = 2,
+  autoplot(ice.glucose, alpha = 0.1, main = "ICE curves") + theme_light(), 
+  autoplot(cice.glucose, alpha = 0.1, main = "c-ICE curves") + theme_light()
+)
+```
+
+```{r example-rf-3, fig.width=7, fig.height=5}
+partial(
+  boston.rf, pred.var = "age", ice = TRUE, center = TRUE, plot = TRUE,
+  plot.engine = "ggplot2", alpha = 0.1
+)
+```
+
+## References
+
+Alex Goldstein, Adam Kapelner, Justin Bleich & Emil Pitkin (2014) Peeking Inside the Black Box: Visualizing Statistical Learning With Plots of Individual Conditional Expectation, Journal of Computational and Graphical Statistics, 24:1, 44-65, DOI: 10.1080/10618600.2014.907095

### 2018-08-10 R/autoplot.R

@@ -69,7 +69,7 @@
 #'
 #' @return A \code{"ggplot"} object.
 #'
-#' @importFrom ggplot2 aes_string autoplot facet_wrap geom_contour geom_line
+#' @importFrom ggplot2 aes autoplot facet_wrap geom_contour geom_line
 #'
 #' @importFrom ggplot2 geom_point geom_rug geom_smooth geom_tile ggplot ggtitle
 #'
@@ -113,12 +113,26 @@
 #' )
 #' }
 autoplot.partial <- function(
-  object, center = FALSE, plot.pdp = TRUE, pdp.color = "red", pdp.size = 1,
-  pdp.linetype = 1, rug = FALSE, smooth = FALSE, smooth.method = "auto",
-  smooth.formula = y ~ x, smooth.span = 0.75, smooth.method.args = list(),
-  contour = FALSE, contour.color = "white",
+  object,
+  center = FALSE,
+  plot.pdp = TRUE,
+  pdp.color = "red",
+  pdp.size = 1,
+  pdp.linetype = 1,
+  rug = FALSE,
+  smooth = FALSE,
+  smooth.method = "auto",
+  smooth.formula = y ~ x,
+  smooth.span = 0.75,
+  smooth.method.args = list(),
+  contour = FALSE,
+  contour.color = "white",
   palette = c("viridis", "magma", "inferno", "plasma", "cividis"),
-  train = NULL, xlab = NULL, ylab = NULL, main = NULL, legend.title = "yhat",
+  train = NULL,
+  xlab = NULL,
+  ylab = NULL,
+  main = NULL,
+  legend.title = "yhat",
   ...
 ) {
 
@@ -133,66 +147,119 @@ autoplot.partial <- function(
   }
 
   # Generate plot
-  if (multi) {
-    ggPlotIceCurves(object, center = center, plot.pdp = plot.pdp,
-                    pdp.color = pdp.color, pdp.size = pdp.size,
-                    pdp.linetype = 1, rug = rug, train = train, xlab = xlab,
-                    ylab = ylab, main = main, ...)
-  } else if (nx == 1L) {  # single predictor
-    ggPlotOnePredictorPDP(object, rug = rug, smooth = smooth,
-                          smooth.method = smooth.method,
-                          smooth.formula = smooth.formula,
-                          smooth.span = smooth.span,
-                          smooth.method.args = smooth.method.args,
-                          train = train, xlab = xlab, ylab = ylab, main = main,
-                          ...)
[…]

### 2018-08-10 R/getIcePredCls.R

@@ -13,7 +13,7 @@ getIceClsProb <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getIceClsLogit.default <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
-  multiClassLogit(pr, which.class = which.class)
+  multiclass_logit(pr, which.class = which.class)
 }
 
 
@@ -27,7 +27,7 @@ getIceClsProb.default <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getIceClsLogit.BinaryTree <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
-  multiClassLogit(do.call(rbind, pr), which.class = which.class)
+  multiclass_logit(do.call(rbind, pr), which.class = which.class)
 }
 
 
@@ -41,7 +41,7 @@ getIceClsProb.BinaryTree <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getIceClsLogit.bagging <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, ...)$prob
-  multiClassLogit(pr, which.class = which.class)
+  multiclass_logit(pr, which.class = which.class)
 }
 
 
@@ -55,7 +55,7 @@ getIceClsProb.bagging <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getIceClsLogit.boosting <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, ...)$prob
-  multiClassLogit(pr, which.class = which.class)
+  multiclass_logit(pr, which.class = which.class)
 }
 
 
@@ -69,7 +69,7 @@ getIceClsProb.boosting <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getIceClsLogit.earth <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, type = "response", ...)
-  multiClassLogit(cbind(pr, 1 - pr), which.class = which.class)
+  multiclass_logit(cbind(pr, 1 - pr), which.class = which.class)
 }
 
 
@@ -83,7 +83,7 @@ getIceClsProb.earth <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getIceClsLogit.fda <- function(object, newdata, which.class, ...) {
[…]

### 2018-08-10 R/getParDepGBM.R

@@ -53,7 +53,7 @@ getParDepGBM <- function(object, pred.var, pred.grid, which.class, prob, ...) {
     if (prob) {  # use class probabilities
       pd.df$yhat <- y[, which.class]
     } else {  # use centered logit
-      pd.df$yhat <- multiClassLogit(y, which.class = which.class)
+      pd.df$yhat <- multiclass_logit(y, which.class = which.class)
     }
   } else if (object$distribution$name %in% c("bernoulli", "pairwise")) {
     pr <- stats::plogis(y)

### 2018-08-10 R/getParPredCls.R

@@ -13,7 +13,7 @@ getParClsProb <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getParClsLogit.default <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
-  mean(multiClassLogit(pr, which.class = which.class), na.rm = TRUE)
+  mean(multiclass_logit(pr, which.class = which.class), na.rm = TRUE)
 }
 
 
@@ -27,7 +27,7 @@ getParClsProb.default <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getParClsLogit.BinaryTree <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
-  mean(multiClassLogit(do.call(rbind, pr), which.class = which.class),
+  mean(multiclass_logit(do.call(rbind, pr), which.class = which.class),
        na.rm = TRUE)
 }
 
@@ -42,7 +42,7 @@ getParClsProb.BinaryTree <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getParClsLogit.bagging <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, ...)$prob
-  mean(multiClassLogit(pr, which.class = which.class), na.action = TRUE)
+  mean(multiclass_logit(pr, which.class = which.class), na.action = TRUE)
 }
 
 
@@ -56,7 +56,7 @@ getParClsProb.bagging <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getParClsLogit.boosting <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, ...)$prob
-  mean(multiClassLogit(pr, which.class = which.class), na.rm = TRUE)
+  mean(multiclass_logit(pr, which.class = which.class), na.rm = TRUE)
 }
 
 
@@ -70,7 +70,7 @@ getParClsProb.boosting <- function(object, newdata, which.class, ...) {
 #' @keywords internal
 getParClsLogit.earth <- function(object, newdata, which.class, ...) {
   pr <- stats::predict(object, newdata = newdata, type = "response", ...)
-  mean(multiClassLogit(cbind(pr, 1 - pr), which.class = which.class),
+  mean(multiclass_logit(cbind(pr, 1 - pr), which.class = which.class),
[…]

### 2018-08-10 R/partial.R

@@ -89,9 +89,6 @@
 #' whenever \code{plot = TRUE}. Options include \code{"lattice"} (defasult) or
 #' \code{"ggplot2"}.
 #'
-#' @param alpha Numeric values between 0 and 1 indicating the alpha-transparency
-#' to use for ICE anc-ICE curves. Default is 0.5.
-#'
 #' @param smooth Logical indicating whether or not to overlay a LOESS smooth.
 #' Default is \code{FALSE}.
 #'
@@ -100,6 +97,9 @@
 #' predictor distributions. This helps reduce the risk of interpreting the
 #' partial dependence plot outside the region of the data (i.e., extrapolating).
 #' Only used when \code{plot = TRUE}. Default is \code{FALSE}.
+#'
+#' @param levelplot Logical indicating whether or not to use a false color level
+#' plot (\code{TRUE}) or a 3-D surface (\code{FALSE}). Default is \code{TRUE}.
 #'
 #' @param contour Logical indicating whether or not to add contour lines to the
 #' level plot. Only used when \code{levelplot = TRUE}. Default is \code{FALSE}.
@@ -297,7 +297,7 @@ partial.default <- function(
 
   # Try to extract training data (hard problem) if not provided
   if (missing(train)) {
-    train <- getTrainingData(object)
+    train <- get_training_data(object)
   }
 
   # Convert the training data to a matrix for XGBoost models
@@ -330,9 +330,10 @@ partial.default <- function(
 
   # Generate grid of predictor values
   pred.grid <- if (missing(pred.grid)) {
-    predGrid(train = train, pred.var = pred.var,
-             grid.resolution = grid.resolution, quantiles = quantiles,
-             probs = probs, trim.outliers = trim.outliers)
+    pred_grid(
+      train = train, pred.var = pred.var, grid.resolution = grid.resolution,
+      quantiles = quantiles, probs = probs, trim.outliers = trim.outliers
+    )
   } else {
     if (!is.data.frame(pred.grid)) {
       stop("`pred.grid` shoud be a data frame.")
@@ -347,14 +348,14 @@ partial.default <- function(
           warning(paste("Options `quantiles` and `trim.outliers`",
                         "ignored whenever `pred.grid` is specified."))
         }
-        orderGrid(pred.grid)
+        order_grid(pred.grid)
       }
     }
   }
 
   # Make sure each column has the correct class, levels, etc.
   if (inherits(train, "data.frame") && check.class) {
-    pred.grid <- copyClasses(pred.grid, train)
[…]

### 2018-08-10 R/plotPartial.R

@@ -3,7 +3,7 @@
 #' Plots partial dependence functions (i.e., marginal effects) using
 #' \code{\link[lattice]{lattice}} graphics.
 #'
-#' @param x An object that inherits from the \code{"partial"} class.
+#' @param object An object that inherits from the \code{"partial"} class.
 #'
 #' @param center Logical indicating whether or not to produce centered ICE
 #' curves (c-ICE curves). Only useful when \code{object} represents a set of ICE
@@ -110,53 +110,95 @@
 #' p2 <- plotPartial(age.ice, center = TRUE, alpha = 0.5)
 #' grid.arrange(p1, p2, ncol = 2)
 #' }
-plotPartial <- function(x, ...) {
+plotPartial <- function(object, ...) {
   UseMethod("plotPartial")
 }
 
 
 #' @rdname plotPartial
+#'
 #' @export
 plotPartial.ice <- function(
-  x, center = FALSE, plot.pdp = TRUE, pdp.col = "red2", pdp.lwd = 2,
-  pdp.lty = 1, rug = FALSE, train = NULL, alpha = 1, ...
+  object,
+  center = FALSE,
+  plot.pdp = TRUE,
+  pdp.col = "red2",
+  pdp.lwd = 2,
+  pdp.lty = 1,
+  rug = FALSE,
+  train = NULL,
+  alpha = 1,
+  ...
 ) {
-  plotIceCurves(x, center = center, plot.pdp = plot.pdp, pdp.col = pdp.col,
-                pdp.lwd = pdp.lwd, pdp.lty = pdp.lty, rug = rug, train = train,
-                alpha = alpha, ...)
+
+  # Call workhorse function
+  plot_ice_curves(
+    object = object, center = center, plot.pdp = plot.pdp, pdp.col = pdp.col,
+    pdp.lwd = pdp.lwd, pdp.lty = pdp.lty, rug = rug, train = train,
+    alpha = alpha, ...
+  )
+
 }
 
 
 #' @rdname plotPartial
+#'
 #' @export
 plotPartial.cice <- function(
-  x, plot.pdp = TRUE, pdp.col = "red2", pdp.lwd = 2, pdp.lty = 1, rug = FALSE,
-  train = NULL, alpha = 1, ...
+  object,
+  plot.pdp = TRUE,
+  pdp.col = "red2",
+  pdp.lwd = 2,
+  pdp.lty = 1,
+  rug = FALSE,
+  train = NULL,
[…]

### 2018-08-10 R/super_type.R

@@ -1,18 +1,18 @@
 #' @keywords internal
-superType <- function(object) {
-  UseMethod("superType")
+super_type <- function(object) {
+  UseMethod("super_type")
 }
 
 
 #' @keywords internal
-superType.default <- function(object) {
+super_type.default <- function(object) {
   warning('`type` could not be determined; assuming `type = "regression"`')
   "regression"
 }
 
 
 #' @keywords internal
-superType.BinaryTree <- function(object) {
+super_type.BinaryTree <- function(object) {
   if (object@responses@is_nominal) {
     "classification"
   } else if (object@responses@is_ordinal || object@responses@is_censored) {
@@ -24,24 +24,24 @@ superType.BinaryTree <- function(object) {
 
 
 #' @keywords internal
-superType.bagging <- function(object) {
+super_type.bagging <- function(object) {
   "classification"
 }
 
 
 #' @keywords internal
-superType.boosting <- function(object) {
+super_type.boosting <- function(object) {
   "classification"
 }
 
 
 #' @keywords internal
-superType.C5.0 <- function(object) {
+super_type.C5.0 <- function(object) {
   "classification"
 }
 
 
-superType.cforest <- function(object) {
+super_type.cforest <- function(object) {
   if (attr(object$terms, "response") == 1 &&
       attr(object$terms, "dataClasses")[1L] == "numeric") {
     "regression"
@@ -55,19 +55,19 @@ superType.cforest <- function(object) {
 
 
 #' @keywords internal
-superType.classbagg <- function(object) {
+super_type.classbagg <- function(object) {
   "classification"
 }
 
 
 #' @keywords internal
-superType.cubist <- function(object) {
+super_type.cubist <- function(object) {
   "regression"
 }
 
 
 #' @keywords internal
-superType.earth <- function(object) {
+super_type.earth <- function(object) {
   if (!is.null(object$glm.list) &&
       object$glm.list[[1L]]$family$family == "binomial") {
     "classification"
@@ -82,13 +82,13 @@ superType.earth <- function(object) {
 
 
 #' @keywords internal
-superType.fda<- function(object) {
+super_type.fda<- function(object) {
   "classification"
 }
 
 
 #' @keywords internal
-superType.gbm <- function(object) {
+super_type.gbm <- function(object) {
   if (object$distribution %in%
       c("gaussian", "laplace", "tdist", "gamma", "poisson", "tweedie")) {
     "regression"
[…]

### 2018-08-10 R/utils.R

@@ -24,7 +24,7 @@ NULL
 NULL
 
 
-#' Retrieve the last Trellis Object
+#' Retrieve the last trellis object
 #'
 #' See \code{\link[lattice]{trellis.last.object}} for more details.
 #'
@@ -32,67 +32,68 @@ NULL
 #' @rdname trellis.last.object
 #' @keywords internal
 #' @export
-#' @importFrom lattice  trellis.last.object
+#' @importFrom lattice trellis.last.object
 #' @usage trellis.last.object(..., prefix)
 NULL
 
 
 #' @keywords internal
-averageIceCurves <- function(object) {
-  UseMethod("averageIceCurves")
+average_ice_curves <- function(object) {
+  UseMethod("average_ice_curves")
 }
 
 
 #' @keywords internal
-averageIceCurves.ice <- function(object) {
-  yhat <- tapply(object[["yhat"]], INDEX = as.factor(object[[1L]]),
-                 FUN = mean, simplify = FALSE)
-  res <- data.frame("x" = object[seq_len(length(yhat)), 1L, drop = TRUE],
-                    "yhat" = unlist(yhat))
+average_ice_curves.ice <- function(object) {
+  yhat <- tapply(
+    object[["yhat"]], INDEX = as.factor(object[[1L]]), FUN = mean,
+    simplify = FALSE
+  )
+  res <- data.frame(
+    "x" = object[seq_len(length(yhat)), 1L, drop = TRUE],
+    "yhat" = unlist(yhat)
+  )
   names(res)[1L] <- names(object)[1L]
   res
 }
 
 
 #' @keywords internal
-averageIceCurves.cice <- function(object) {
-  averageIceCurves.ice(object)
+average_ice_curves.cice <- function(object) {
+  average_ice_curves.ice(object)
 }
 
 
 #' @keywords internal
-averageIceCurves.partial <- function(object) {
-  averageIceCurves.ice(object)
+average_ice_curves.partial <- function(object) {
+  average_ice_curves.ice(object)
 }
 
 
 #' @keywords internal
-centerIceCurves <- function(object) {
-  UseMethod("centerIceCurves")
+center_ice_curves <- function(object) {
+  UseMethod("center_ice_curves")
 }
 
 
 #' @keywords internal
-centerIceCurves.ice <- function(object) {
-  yhat <- tapply(object[["yhat"]], INDEX = as.factor(object[["yhat.id"]]),
[…]

### 2018-08-10 R/autoplot.R
minor tweaks to pass R CMD checks

@@ -263,16 +263,17 @@ ggplot_ice_curves <- function(
 
   # Should the curves be centered to start at yhat = 0?
   if (center) {
-    object <- centerIceCurves(object)
+    object <- center_ice_curves(object)
   }
 
   # Use the first column to determine which type of plot to construct
   if (is.factor(object[[1L]])) {
 
     # Draw scatterplots
-    p <- ggplot(object, aes(x = object[[1L]], y = yhat, group = 1)) +
-      geom_line(aes(group = yhat.id), ...) +
-      geom_point(aes(group = yhat.id), alpha = 1)
+    p <- ggplot(object,
+                aes(x = object[[1L]], y = object[["yhat"]], group = 1)) +
+      geom_line(aes(group = object[["yhat.id"]]), ...) +
+      geom_point(aes(group = object[["yhat.id"]]), alpha = 1)
 
     # Should the PDP be displayed too?
     if (plot.pdp) {
@@ -285,8 +286,8 @@ ggplot_ice_curves <- function(
   } else {
 
     # Draw lineplots
-    p <- ggplot(object, aes(x = object[[1L]], y = yhat)) +
-      geom_line(aes(group = yhat.id), ...)
+    p <- ggplot(object, aes(x = object[[1L]], y = object[["yhat"]])) +
+      geom_line(aes(group = object[["yhat.id"]]), ...)
 
     # Should the PDP be displayed too?
     if (plot.pdp) {
@@ -353,13 +354,13 @@ ggplot_one_predictor_pdp <- function(
   if (is.factor(object[[1L]])) {
 
     # Draw a scatterplot
-    p <- ggplot(object, aes(x = object[[1L]], y = yhat)) +
+    p <- ggplot(object, aes(x = object[[1L]], y = object[["yhat"]])) +
       geom_point(...)
 
   } else {
 
     # Draw a lineplot
-    p <- ggplot(object, aes(x = object[[1L]], y = yhat)) +
+    p <- ggplot(object, aes(x = object[[1L]], y = object[["yhat"]])) +
       geom_line(...)
 
     # Add rug plot to x-axis
@@ -431,14 +432,14 @@ ggplot_two_predictor_pdp <- function(
   if (is.factor(object[[1L]]) && is.factor(object[[2L]])) {
 
     # Draw a faceted scatterplot
-    p <- ggplot(object, aes(x = object[[1L]], y = yhat)) +
[…]

### 2018-08-10 R/plotPartial.R

@@ -266,7 +266,7 @@ plot_ice_curves <- function(
 
   # Determine if ICE curves should be centered
   if (center) {
-    object <- centerIceCurves(object)  # converts ICE curves to c-ICE curves
+    object <- center_ice_curves(object)  # converts ICE curves to c-ICE curves
   }
 
   # Determine plot type

### 2018-08-10 R/super_type.R

@@ -285,7 +285,7 @@ super_type.train <- function(object) {
 #' @keywords internal
 super_type.xgb.Booster <- function(object) {
   if (object$params$objective %in%
-      c("reg:linear", "count:poisson", "reg:gamma")) {
+      c("reg:linear", "reg:logistic", "count:poisson", "reg:gamma")) {
     "regression"
   } else if (object$params$objective %in%
              c("binary:logistic", "multi:softprob")) {
@@ -293,7 +293,7 @@ super_type.xgb.Booster <- function(object) {
     # not return the predicted probabilities (e.g., "binary:logitraw").
     "classification"
   } else if (object$params$objective %in%
-             c("reg:logistic", "binary:logitraw", "multi:softmax")) {
+             c("binary:logitraw", "multi:softmax")) {
     stop(paste("For classification, switch to an objective function",
                "that returns the predicted probabilities."))
   } else {

### 2018-08-14 README.Rmd
start vignettes

--- 2018-05-17 README.Rmd
@@
 knitr::opts_chunk$set(
+  cache = TRUE,
+  cache.path = "tools/README-cache/",
   collapse = TRUE,
   comment = "#>",
-  fig.path = "tools/README-"
+  fig.align = "center",
+  fig.path = "man/figures/README-"
 )
@@
 
-Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality).
-Unfortunately, it can be challenging to understand the results of such models and explain them to management.
-Partial dependence plots offer a simple solution.
-Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be more easily understood.
-These plots are especially useful in explaining the output from black box models.
-The `pdp` package offers a general framework for constructing partial dependence plots for various types of fitted models in R.
+[pdp](https://cran.r-project.org/package=pdp) is an R package for constructing _**p**artial **d**ependence **p**lots_ (PDPs) and _**i**ndividual **c**onditional **e**xpectation_ (ICE) curves.
+PDPs and ICE curves are part of a larger framework referred to as *interpretable machine learning* (IML), which also includes (but not limited to) _**v**ariable **i**mportance **p**lots_ (VIPs).
+While VIPs (available in the R package [vip](https://koalaverse.github.io/vip/index.html)) help visualize feature impact (either locally or globally), PDPs and ICE curves help visualize feature effects.
+An in-progress, but comprehensive, overview of IML can be found at the following URL: https://github.com/christophM/interpretable-ml-book.
 
-A detailed introduction to `pdp` has been published in The R Journal: "pdp: An R Package for Constructing Partial Dependence Plots", https://journal.r-project.org/archive/2017/RJ-2017-016/index.html.
+A detailed introduction to [pdp](https://cran.r-project.org/package=pdp) has been published in The R Journal: "pdp: An R Package for Constructing Partial Dependence Plots", https://journal.r-project.org/archive/2017/RJ-2017-016/index.html.
 You can track development at https://github.com/bgreenwell/pdp.
@@
 
-* `partial` - compute partial dependence functions (i.e., objects of class `"partial"`) from various fitted model objects;
-* `plotPartial"` - plot partial dependence functions (i.e., objects of class `"partial"`) using `lattice` graphics;
-* `autoplot` - plot partial dependence functions (i.e., objects of class `"partial"`) using `ggplot2` graphics;
-* `topPredictors` extract most "important" predictors from various types of fitted models.
+  * `partial()` - compute partial dependence functions and individual conditional expectations (i.e., objects of class `"partial"` and `"ice"`, respectively) from various fitted model objects;
+
+  * `plotPartial()"` - construct `lattice`-based PDPs and ICE curves;
+
+  * `autoplot()` - construct `ggplot2`-based PDPs and ICE curves;
+
+  * `topPredictors()` extract most "important" predictors from various types of fitted models. (Will soon be replaced by functionality from [vip](https://koalaverse.github.io/vip/index.html).)
 
@@
 
-The `pdp` package is [currently listed on CRAN](https://CRAN.R-project.org/package=pdp) and can easily be installed:
-```r
-  # Install from CRAN (recommended)
+```{r, eval=FALSE}
+# The easiest way to get pdp is to install it from CRAN:
 install.packages("pdp")
 
-  # Alternatively, install the development version from GitHub
+# Alternatively, you can install the development version from GitHub:
+if (!requireNamespace("devtools")) {
+  install.packages("devtools")
+}
 devtools::install_github("bgreenwell/pdp")
 ```
-
-## Random forest example
-
-As a first example, we'll fit a random forest to the famous Boston housing data included with the package (see `?boston` for details).
-In fact the original motivation for this package was to be able to compute two-predictor partial dependence plots from random forest models in R.
-```{r example-rf, fig.width=8, fig.height=4, message=FALSE}
-# Fit a random forest to the Boston housing data
-library(randomForest)  # install.packages("randomForest")
-data (boston)  # load the boston housing data
-set.seed(101)  # for reproducibility
-boston.rf <- randomForest(cmedv ~ ., data = boston)
-
-# Partial dependence of cmedv on lstat and rm
-library(pdp)
-pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE)
-head(pd)  # print first 6 rows
-
-# Lattice version
-p1 <- plotPartial(pd, main = "lattice version")
-
-# ggplot2 version
-library(ggplot2)
-p2 <- autoplot(pd, contour = TRUE, main = "ggplot2 version", 
-               legend.title = "Partial\ndependence")
-
-# Show both plots in one figure
-grid.arrange(p1, p2, ncol = 2)
-```
-
-Next, we'll fit a classification model to the Pima Indians Diabetes data.
-
-## Support vector machine (SVM) example
-
-As a second example, we'll fit an SVM to the Pima Indians diabetes data included with the package (see `?pima` for details).
-Note that for some fitted model objects (e.g., `"ksvm"` objects) it is necessary to supply the original training data via the `train` argument in the call to `partial`.
-
-```{r example-svm, fig.width=8, fig.height=4, message=FALSE}
-# Fit an SVM to the Pima Indians diabetes data
-library(kernlab)  # install.packages("kernlab")
-data (pima)  # load the Pima Indians diabetes data
-pima.svm <- ksvm(diabetes ~ ., data = pima, type = "C-svc", kernel = "rbfdot",
-                 C = 0.5, prob.model = TRUE)
-
-# Partial dependence of diabetes test result on glucose (default is logit scale)
-pd.glucose <- partial(pima.svm, pred.var = "glucose", train = pima)
-
-# Partial dependence of diabetes test result on glucose (probability scale)
-pd.glucose.prob <- partial(pima.svm, pred.var = "glucose", prob = TRUE, 
-                           train = pima)
-
-# Show both plots in one figure
-grid.arrange(autoplot(pd.glucose, main = "Logit scale"), 
-             autoplot(pd.glucose.prob, main = "Probability scale"), 
-             ncol = 2)
-```

### 2018-08-15 R/plotPartial.R
bug fix; more vignette material

@@ -42,6 +42,11 @@
 #' @param contour.color Character string specifying the color to use for the
 #' contour lines when \code{contour = TRUE}. Default is \code{"white"}.
 #'
+#' @param col.regions Color vector to be used for trivariate displays. If
+#' \code{levelplot} is \code{TRUE}, defaults to the wonderful Matplotlib
+#' 'viridis' color map provided by the \code{viridis} package. See
+#' \code{\link[viridis]{viridis}} for details.
+#'
 #' @param palette Character string indicating the colormap option to use. Five
 #' options are available: "viridis" (the default), "magma", "inferno", "plasma",
 #' and "cividis".
@@ -60,11 +65,6 @@
 #'
 #' @param train Data frame containing the original training data. Only required
 #' if \code{rug = TRUE} or \code{chull = TRUE}.
-#'
-#' @param col.regions Color vector to be used for trivariate displays. If
-#' \code{levelplot} is \code{TRUE}, defaults to the wonderful Matplotlib
-#' 'viridis' color map provided by the \code{viridis} package. See
-#' \code{\link[viridis]{viridis}} for details.
 #'
 #' @param ... Additional optional arguments to be passed onto \code{dotplot},
 #' \code{levelplot}, \code{xyplot}, or \code{wireframe}.
@@ -241,8 +241,8 @@ plotPartial.partial <- function(
     plot_three_predictor_pdp(  # three predictors (paneled)
       object = object, nx = nx, smooth = smooth, levelplot = levelplot,
       rug = rug, chull = chull, train = train, contour = contour,
-      contour.color = contour.color, palette = palette, alpha = alpha,
-      number = number, overlap = overlap, ...
+      contour.color = contour.color, col.regions = col.regions,
+      palette = palette, alpha = alpha, number = number, overlap = overlap, ...
     )
 
   }
@@ -477,6 +477,7 @@ plot_three_predictor_pdp <- function(
   train,
   contour,
   contour.color,
+  col.regions,
   palette,
   alpha,
   number,

### 2018-08-21 R/autoplot.R
add first draft of website and vignettes; minor bug fixes

@@ -306,7 +306,7 @@ ggplot_ice_curves <- function(
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))
-        p <- p + geom_rug(data = x.rug, aes(x = x.rug[1L]),csides = "b",
+        p <- p + geom_rug(data = x.rug, aes(x = x.rug[[1L]]),csides = "b",
                           inherit.aes = FALSE)
       }
     }
@@ -372,7 +372,7 @@ ggplot_one_predictor_pdp <- function(
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))
-        p <- p + geom_rug(data = x.rug, aes(x = x.rug[1L]), sides = "b",
+        p <- p + geom_rug(data = x.rug, aes(x = x.rug[[1L]]), sides = "b",
                           inherit.aes = FALSE)
       }
     }
@@ -452,7 +452,7 @@ ggplot_two_predictor_pdp <- function(
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))
-        p <- p + geom_rug(data = x.rug, aes(x = x.rug[1L]), sides = "b",
+        p <- p + geom_rug(data = x.rug, aes(x = x.rug[[1L]]), sides = "b",
                           inherit.aes = FALSE)
       }
     }
@@ -481,7 +481,7 @@ ggplot_two_predictor_pdp <- function(
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))
-        p <- p + geom_rug(data = x.rug, aes(x = x.rug[1L]), sides = "b",
+        p <- p + geom_rug(data = x.rug, aes(x = x.rug[[1L]]), sides = "b",
                           inherit.aes = FALSE)
       }
     }

### 2018-08-21 R/partial.R

@@ -482,7 +482,7 @@ partial.default <- function(
   # Plot partial dependence function (if requested)
   if (plot) {  # return a graph (i.e., a "trellis" or "ggplot" object)
     plot.engine <- match.arg(plot.engine)
-    res <- if (ice) {
+    res <- if (inherits(pd.df, what = "ice")) {
       if (plot.engine == "ggplot2") {
         autoplot(
           object = pd.df, plot.pdp = TRUE, rug = rug, train = train,

### 2018-08-24 README.Rmd
move logos
lighten logo gradient

@@
 
-# pdp <img src="tools/pdp-logo.png" align="right" width="130" height="150" />
+# pdp <img src="man/figures/pdp-logo.png" align="right" width="130" height="150" />
 
@@
 [![Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/pdp)](http://cranlogs.r-pkg.org/badges/grand-total/pdp)
+[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
 

### 2018-08-24 R/getParDepH2O.R

@@ -1,0 +1,25 @@
+#' @keywords internal
+getParDepH2O <- function(
+  object, pred.var, pred.grid, pred.fun, train, progress, parallel, paropts,
+  ...
+) {
+  plyr::adply(
+    pred.grid, .margins = 1, .progress = progress, .parallel = parallel,
+    .paropts = paropts, .id = NULL,
+    .fun = function(x) {
+      temp <- train
+      temp[, pred.var] <- x[1L, 1L]
+      out <- mean(predict(object, newdata = temp))
+      print(out)
+      if (length(out) == 1) {
+        stats::setNames(out, "yhat")
+      } else {
+        if (is.null(names(out))) {
+          stats::setNames(out, paste0("yhat.", 1L:length(out)))
+        } else {
+          stats::setNames(out, paste0("yhat.", names(out)))
+        }
+      }
+    }
+  )
+}

### 2018-08-24 tools/pdp-logo-img.R

@@ -1,86 +1,0 @@
-# Load required packages
-library(earth)
-library(ggplot2)
-library(grid)
-library(pdp)
-library(png)
-library(randomForest)
-
-# Fit a random forest to the Boston housing data
-data (boston)  # load the boston housing data
-set.seed(101)  # for reproducibility
-boston.rf <- randomForest(cmedv ~ ., data = boston)
-boston.earth <- earth(cmedv ~ ., data = boston, degree = 2,
-                      pmethod = "exhaustive", nfold = 5, ncross = 5)
-
-# Print MARS model coefficients
-coef(boston.earth)
-
-# Function to rescale vector to be between a and b
-rescale <- function(x, a, b) {
-  ((x - min(x)) / (max(x) - min(x))) * (b - a) + a
-}
-
-# Partial dependence of cmedv on lstat and rm
-pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = FALSE,
-              progress = "text", grid.resolution = 100)
-
-# Boundaries of the hexagon
-hex <- data.frame(x = 1.35 * 1 * c(-sqrt(3) / 2, 0, rep(sqrt(3) / 2, 2), 0,
-                                   rep(-sqrt(3) / 2, 2)),
-                  y = 1.35 * 1 * c(0.5, 1, 0.5, -0.5, -1, -0.5, 0.5))
-
-# Restrict PDP to the boundaries of the hexagon
-pd_hex <- pd
-pd_hex$lstat <- rescale(pd_hex$lstat, a = min(hex$x), b = max(hex$x))
-pd_hex$rm <- rescale(pd_hex$rm, a = min(hex$y), b = max(hex$y))
-pd_hex <- pd_hex[mgcv::in.out(as.matrix(hex), as.matrix(pd_hex[, 1L:2L])), ]
-
-# Hexagon logo
-make_pdp_sticker <- function(option) {
-  ggplot(pd_hex, aes(lstat, rm)) +
-    geom_polygon(data = hex, aes(x, y), color = "black", fill = grey(0.25),
-                 size = 3) +
-    geom_tile(data = pd_hex, aes(x = lstat, y = rm, z = yhat, fill = yhat)) +
-    # scale_fill_distiller(name = "yhat", palette = "Spectral") +
-    viridis::scale_fill_viridis(option = option) +
-    geom_polygon(data = hex, aes(x, y), color = "black", fill = "transparent",
-                 size = 3) +
-    geom_contour(aes(z = yhat), color = "black") +
[…]

### 2018-08-24 R/autoplot.R
fix spelling errors

@@ -40,7 +40,7 @@
 #' smoother lines. Default is \code{0.75}.
 #'
 #' @param smooth.method.args List containing additional arguments to be passed
-#' on to the modelling function defined by \code{smooth.method}.
+#' on to the modeling function defined by \code{smooth.method}.
 #'
 #' @param contour Logical indicating whether or not to add contour lines to the
 #' level plot.
@@ -55,14 +55,14 @@
 #' @param train Data frame containing the original training data. Only required
 #' if \code{rug = TRUE} or \code{chull = TRUE}.
 #'
-#' @param xlab Charater string specifying the text for the x-axis label.
+#' @param xlab Character string specifying the text for the x-axis label.
 #'
-#' @param ylab Charater string specifying the text for the y-axis label.
+#' @param ylab Character string specifying the text for the y-axis label.
 #'
 #' @param main Character string specifying the text for the main title of the
 #' plot.
 #'
-#' @param legend.title Charater string specifying the text for the legend title.
+#' @param legend.title Character string specifying the text for the legend title.
 #' Default is \code{"yhat"}.
 #'
 #' @param ... Additional optional arguments to be passed onto \code{geom_line}.

### 2018-08-24 R/partial.R

@@ -56,7 +56,7 @@
 #'
 #' @param inv.link Function specifying the transformation to be applied to the
 #' predictions before the partial dependence function is computed
-#' (experimental). Default is \code{NULL} (i.e., no transofrmation). This option
+#' (experimental). Default is \code{NULL} (i.e., no transformation). This option
 #' is intended to be used for models that allow for non-Gaussian response
 #' variables (e.g., counts). For these models, predictions are not typically
 #' returned on the original response scale by default. For example, Poisson GBMs
@@ -86,7 +86,7 @@
 #' \code{\link{plotPartial}} for plotting details.
 #'
 #' @param plot.engine Character string specifying which plotting engine to use
-#' whenever \code{plot = TRUE}. Options include \code{"lattice"} (defasult) or
+#' whenever \code{plot = TRUE}. Options include \code{"lattice"} (default) or
 #' \code{"ggplot2"}.
 #'
 #' @param smooth Logical indicating whether or not to overlay a LOESS smooth.
@@ -116,7 +116,7 @@
 #' transparency). In fact, this option only affects ICE/c-ICE curves and level
 #' plots.
 #'
-#' @param chull Logical indicating wether or not to restrict the values of the
+#' @param chull Logical indicating whether or not to restrict the values of the
 #' first two variables in \code{pred.var} to lie within the convex hull of their
 #' training values; this affects \code{pred.grid}. This helps reduce the risk of
 #' interpreting the partial dependence plot outside the region of the data

### 2018-08-24 R/plotPartial.R

@@ -29,7 +29,7 @@
 #' @param rug Logical indicating whether or not to include rug marks on the
 #' predictor axes. Default is \code{FALSE}.
 #'
-#' @param chull Logical indicating wether or not to restrict the first two
+#' @param chull Logical indicating whether or not to restrict the first two
 #' variables in \code{pred.var} to lie within the convex hull of their training
 #' values; this affects \code{pred.grid}. Default is \code{FALSE}.
 #'

### 2018-08-25 R/getParDepH2O.R
ignore vignettes

@@ -1,25 +1,0 @@
-#' @keywords internal
-getParDepH2O <- function(
-  object, pred.var, pred.grid, pred.fun, train, progress, parallel, paropts,
-  ...
-) {
-  plyr::adply(
-    pred.grid, .margins = 1, .progress = progress, .parallel = parallel,
-    .paropts = paropts, .id = NULL,
-    .fun = function(x) {
-      temp <- train
-      temp[, pred.var] <- x[1L, 1L]
-      out <- mean(predict(object, newdata = temp))
-      print(out)
-      if (length(out) == 1) {
-        stats::setNames(out, "yhat")
-      } else {
-        if (is.null(names(out))) {
-          stats::setNames(out, paste0("yhat.", 1L:length(out)))
-        } else {
-          stats::setNames(out, paste0("yhat.", names(out)))
-        }
-      }
-    }
-  )
-}

### 2018-08-27 slowtests/slowtests-boston.Rmd
update tests

---
title: "Tests: Bouston Housing Data"
author: "Brandon M. Greenwell"
output: html_document
---

```{r setup}
# Load required packages
library(ggplot2)
library(grid)
library(pdp)

# Test function
boston_test <- function(object, ...) {
  trn <- subset(boston, select = -cmedv)
  pd1 <- partial(object, pred.var = "lstat", train = trn, ...)
  pd2 <- partial(object, pred.var = c("lstat", "rm"), chull = TRUE,
                 train = trn, ...)
  pd3 <- partial(object, pred.var = "rm", ice = TRUE, train = trn, ...)
  pd4 <- partial(object, pred.var = "rm", ice = TRUE, center = TRUE,
                 train = trn, ...)
  grid.arrange(
    autoplot(pd1, rug = TRUE, train = trn, main = "PDP for lstat"),
    autoplot(pd2, main = "PDP for lstat and rm"),
    autoplot(pd3, main = "ICE curves for rm", alpha = 0.1),
    autoplot(pd4, main = "c-ICE curves for rm", alpha = 0.1),
    ncol = 2, top = textGrob(deparse(substitute(object)),
                             gp = gpar(fontsize = 20, font = 3))
  )
}
```

# Cubist

## Package: `Cubist`

```{r Cubist_cubist}
boston_cubist <- Cubist::cubist(x = subset(boston, select = -cmedv),
                                y = boston$cmedv, committees = 100)
boston_test(boston_cubist)
```

# Decision trees

## Package: `rpart`

```{r rpart_rpart}
boston_rpart <- rpart::rpart(cmedv ~ ., data = boston)
boston_test(boston_rpart)
```

## Package: `party`

```{r party_ctree}
boston_ctree <- party::ctree(cmedv ~ ., data = boston)
boston_test(boston_ctree)
```

## Package: `partykit`

```{r partykit_ctree}
boston_ctree2 <- partykit::ctree(cmedv ~ ., data = boston)
boston_test(boston_ctree2)
```

# Bagging

## Package: `ipred`

```{r ipred_bagging}
set.seed(101)
boston_ipred <- ipred::bagging(cmedv ~ ., data = boston, nbagg = 500)
boston_test(boston_ipred)
```

# Random forests

## Package: `randomForest`

```{r randomForest_randomForest}
set.seed(101)
boston_rf <- randomForest::randomForest(cmedv ~ ., data = boston)
boston_test(boston_rf)
```

## Package: `party`

```{r party_cforest}
set.seed(101)
boston_crf <- party::cforest(cmedv ~ ., data = boston)
boston_test(boston_crf, quantiles = TRUE)
```

## Package: `ranger`

```{r ranger_ranger}
set.seed(101)
boston_ranger <- ranger::ranger(cmedv ~ ., data = boston)
boston_test(boston_ranger)
```

# Boosting

## Package: `gbm`

```{r gbm_gbm}
set.seed(101)
boston_gbm <- gbm::gbm(cmedv ~ ., data = boston, distribution = "gaussian",
                       n.trees = 5000, interaction.depth = 3, shrinkage = 0.001,
                       cv.folds = 5)
best.iter <- gbm::gbm.perf(boston_gbm, method = "OOB", plot.it = FALSE)
boston_test(boston_gbm, n.trees = best.iter)
```

## Package: `xgboost`

```{r xgboost_xgboost}
set.seed(101)
boston_xgb <- xgboost::xgboost(
  data = data.matrix(subset(boston, select = -cmedv)), label = boston$cmedv,
  objective = "reg:linear", nrounds = 5000, max_depth = 3, eta = 0.001,
  colsample_bytree = 0.8, subsample = 0.5,
  save_period = NULL, verbose = 0
)
boston_test(boston_xgb)
```

# Neural networks

## Package: `nnet`

```{r nnet_nnet}
set.seed(101)
boston_nnet <- nnet::nnet(cmedv ~ ., data = boston, size = 6, decay = 0.1,
                          linout = TRUE, trace = FALSE)
boston_test(boston_nnet)
```

# Support vector machines

## Package: `e1071`

```{r e1071_svm}
boston_svm <- e1071::svm(cmedv ~ ., data = boston, type = "eps-regression")
boston_test(boston_svm)
```

## Package: `kernlab`

```{r kernlab_ksvm}
boston_ksvm <- kernlab::ksvm(cmedv ~ ., data = boston, type= "eps-svr")
boston_test(boston_ksvm)
```

# Linear/generalized linear models

## Package: `stats`

```{r stats_lm}
boston_lm <- lm(cmedv ~ . ^ 2, data = boston)
boston_test(boston_lm)
```

```{r stats_glm}
boston_glm <- glm(cmedv ~ . ^ 2, data = boston)
boston_test(boston_glm)
```

# Multivariate adaptive regression splines

## Package: `earth`

```{r earth_earth}
boston_earth <- earth::earth(cmedv ~ ., degree = 3, data = boston)
boston_test(boston_earth)
```

## Package: `mda`

```{r mda_mars}
boston_mars <- mda::mars(x = subset(boston, select = - cmedv), 
                         y = boston$cmedv, degree = 3)
boston_test(boston_mars)
```

### 2018-08-27 slowtests/slowtests-iris.Rmd
update tests

---
title: "Tests: Edgar Anderson's Iris Data"
author: "Brandon M. Greenwell"
output: html_document
---

```{r setup}
# Load required packages
library(ggplot2)
library(grid)
library(pdp)

# Test function
iris_test <- function(object, ...) {
  trn <- subset(iris, select = -Species)
  pd1 <- partial(object, pred.var = "Petal.Length", which.class = 1, 
                 prob = TRUE, train = trn, ...)
  pd2 <- partial(object, pred.var = "Petal.Length", which.class = 2, 
                 prob = TRUE, train = trn, ...)
  pd3 <- partial(object, pred.var = "Petal.Length", which.class = 3, 
                 prob = TRUE, train = trn, ...)
  pd <- rbind(cbind(pd1, "Class" = "1"), cbind(pd2, "Class" = "2"),
              cbind(pd3, "Class" = "3"))
  ggplot(pd, aes(x = Petal.Length, y = yhat, color = Class)) +
    geom_line() +
    xlab("Petal length") +
    ylab("Probability") +
    ggtitle(deparse(substitute(object))) +
    theme_light()
}
```

# Discriminant analysis

## Package: `MASS`

```{r mass_lda}
iris_lda <- MASS::lda(Species ~ ., data = iris)
iris_test(iris_lda)
```

```{r mass_qda}
iris_qda <- MASS::qda(Species ~ ., data = iris)
iris_test(iris_qda)
```

## Package: `mda`

```{r mda_fda}
iris_fda <- mda::fda(Species ~ ., data = iris)
iris_test(iris_fda)
```

```{r mda_mda}
iris_fda <- mda::mda(Species ~ ., data = iris)
iris_test(iris_fda)
```

# Decision trees

## Package: `rpart`

```{r rpart_rpart}
iris_rpart <- rpart::rpart(Species ~ ., data = iris)
iris_test(iris_rpart)
```

## Package: `C50`

```{r C50_C50}
set.seed(101)  # for reproducibility
iris_C5.0 <- C50::C5.0(Species ~ ., data = iris, trials = 100)
iris_test(iris_C5.0)
```

## Package: `party`

```{r party_ctree}
iris_ctree <- party::ctree(Species ~ ., data = iris)
iris_test(iris_ctree)
```

## Package: `partykit`

```{r partykit_ctree}
iris_ctree2 <- partykit::ctree(Species ~ ., data = iris)
iris_test(iris_ctree2)
```

# Bagging

## Package: `adabag`

```{r adabag_bagging}
set.seed(101)  # for reproducibility
iris_bagging <- adabag::bagging(Species ~ ., data = iris)
iris_test(iris_bagging)
```

## Package: `ipred`

```{r ipred_bagging}
set.seed(101)  # for reproducibility
iris_ipred <- ipred::bagging(Species ~ ., data = iris, nbagg = 500)
iris_test(iris_ipred)
```

# Random forests

## Package: `randomForest`

```{r randomForest_randomForest}
set.seed(101)  # for reproducibility
iris_rf <- randomForest::randomForest(Species ~ ., data = iris)
iris_test(iris_rf)
```

## Package: `party`

```{r party_cforest}
set.seed(101)  # for reproducibility
iris_crf <- party::cforest(Species ~ ., data = iris)
iris_test(iris_crf, quantiles = TRUE)
```

## Package: `partykit`

```{r partykit_cforest}
set.seed(101)  # for reproducibility
iris_crf2 <- partykit::cforest(Species ~ ., data = iris)
iris_test(iris_crf2, quantiles = TRUE)
```

## Package: `ranger`

```{r ranger_ranger}
set.seed(101)  # for reproducibility
iris_ranger <- ranger::ranger(Species ~ ., data = iris, probability = TRUE)
iris_test(iris_ranger)
```

# Boosting

## Package: `adabag`

```{r adabag_boosting}
set.seed(101)  # for reproducibility
iris_boosting <- adabag::boosting(Species ~ ., data = iris)
iris_test(iris_boosting)
```

## Package: `gbm`

```{r gbm_gbm}
set.seed(101)  # for reproducibility
iris_gbm <- gbm::gbm(Species ~ ., 
                     data = iris,
                     distribution = "multinomial",
                     n.trees = 5000,
                     interaction.depth = 3,
                     shrinkage = 0.001,
                     cv.folds = 5,
                     verbose = FALSE)
best.iter <- gbm::gbm.perf(iris_gbm, method = "cv", plot.it = FALSE)
iris_test(iris_gbm, n.trees = best.iter)
```

## Package: `xgboost`

```{r xgboost_xgboost}
set.seed(101)  # for reproducibility
iris_xgb <- xgboost::xgboost(
  data = data.matrix(subset(iris, select = -Species)), 
  label = unclass(iris$Species) - 1, num_class = 3, 
  objective = "multi:softprob", nrounds = 100, max_depth = 3, eta = 0.1, 
  gamma = 0, colsample_bytree = 0.8, min_child_weight = 1, subsample = 0.7,
  verbose = 0
)
iris_test(iris_xgb)
```

# Neural networks

## Package: `nnet`

```{r nnet_nnet}
set.seed(101)  # for reproducibility
iris_nnet <- nnet::nnet(Species ~ ., data = iris, size = 10, decay = 0.1, 
                        maxit = 500, trace = FALSE)
iris_test(iris_nnet)
```

# Support vector machines

## Package: `e1071`

```{r e1071_svm}
iris_svm <- e1071::svm(Species ~ ., data = iris, type = "C-classification", 
                       probability = TRUE)
iris_test(iris_svm)
```

## Package: `kernlab`

```{r kernlab_ksvm}
iris_ksvm <- kernlab::ksvm(Species ~ ., data = iris, type = "C-svc", 
                           prob.model = TRUE)
iris_test(iris_ksvm)
```

# Linear/generalized linear models

## Package: `nnet`

```{r nnet_multinom}
set.seed(101)  # for reproducibility
iris_multinom <- nnet::multinom(Species ~ ., data = iris, trace = FALSE)
iris_test(iris_multinom)
```

### 2018-08-27 slowtests/slowtests-pima.Rmd
update tests

@@
 ---
-title: "Pima  Tests"
+title: "Tests: Pima Indians Diabetes Data"
 author: "Brandon M. Greenwell"
-date: "July 15, 2017"
 output: html_document
@@
 
-```{r setup, include=FALSE}
+```{r setup}
 # Load required packages
@@
 
-# Set knitr chunk options
-knitr::opts_chunk$set(echo = TRUE)
-
-# Function to run tests for a fitted model object using the Pima Indians 
-# diabetes data
-pimaIndiansTest <- function(object, ...) {
-
-  # Training data
+# Test function
+pima_test <- function(object, ...) {
   pima <- na.omit(pima)
   trn <- subset(pima, select = -diabetes)
-
-  # Partial dependence plot for lstat
   pd1 <- partial(object, pred.var = "glucose", train = trn, ...)
-
-  # Partial dependence plot for lstat and rm
   pd2 <- partial(object, pred.var = c("glucose", "age"), chull = TRUE,
                  train = trn, ...)
-
-  # Individual conditional expectation curves for age
-  pd3 <- partial(object, pred.var = "glucose", ice = TRUE, train = trn,
-                 ...)
-
-  # Centered individual conditional expectation curves for lstat
+  pd3 <- partial(object, pred.var = "glucose", ice = TRUE, train = trn, ...)
   pd4 <- partial(object, pred.var = "glucose", ice = TRUE, center = TRUE,
                  train = trn, ...)
-
-  # Display plots
   grid.arrange(
@@
     autoplot(pd3, rug = TRUE, train = pima,
-             main = "ICE curves for glucose", alpha = 0.25),
+             main = "ICE curves for glucose", alpha = 0.1),
     autoplot(pd4, rug = TRUE, train = pima,
-             main = "c-ICE curves for glucose", alpha = 0.25),
+             main = "c-ICE curves for glucose", alpha = 0.1),
     ncol = 2,
@@
   )
-
 }
@@
 
-```{r}
-pima.lda <- MASS::lda(diabetes ~ . ^ 2, data = na.omit(pima))
-pimaIndiansTest(pima.lda)
-pimaIndiansTest(pima.lda, prob = TRUE)
+```{r mass_lda}
+pima_lda <- MASS::lda(diabetes ~ . ^ 2, data = na.omit(pima))
+pima_test(pima_lda)
+pima_test(pima_lda, prob = TRUE)
 ```
 
-```{r}
-pima.qda <- MASS::qda(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.qda)
-pimaIndiansTest(pima.qda, prob = TRUE)
+```{r mass_qda}
+pima_qda <- MASS::qda(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_qda)
+pima_test(pima_qda, prob = TRUE)
 ```
@@
 
-```{r}
-pima.mda <- mda::mda(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.mda)
-pimaIndiansTest(pima.mda, prob = TRUE)
+```{r mda_fda}
+pima_fda <- mda::fda(diabetes ~ ., data = na.omit(pima), method = mda::mars,
+                     degree = 2)
+pima_test(pima_fda)
+pima_test(pima_fda, prob = TRUE)
 ```
 
-```{r}
-pima.fda <- mda::fda(diabetes ~ ., data = na.omit(pima), method = mda::mars,
-                     degree = 2)
-pimaIndiansTest(pima.fda)
-pimaIndiansTest(pima.fda, prob = TRUE)
+```{r mda_mda}
+pima_mda <- mda::mda(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_mda)
+pima_test(pima_mda, prob = TRUE)
 ```
@@
 
-```{r}
-pima.rpart <- rpart::rpart(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.rpart)
-pimaIndiansTest(pima.rpart, prob = TRUE)
+```{r rpart_rpart}
+pima_rpart <- rpart::rpart(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_rpart)
+pima_test(pima_rpart, prob = TRUE)
 ```
@@
 ## Package: `C50`
-```{r}
-set.seed(101)
-pima.C5.0 <- C50::C5.0(diabetes ~ ., data = na.omit(pima), trials = 100)
-pimaIndiansTest(pima.C5.0)
-pimaIndiansTest(pima.C5.0, prob = TRUE)
+
+```{r C50_C50}
+set.seed(101)  # for reproducibility
+pima_C5.0 <- C50::C5.0(diabetes ~ ., data = na.omit(pima), trials = 100)
+pima_test(pima_C5.0)
+pima_test(pima_C5.0, prob = TRUE)
 ```
@@
 
-```{r}
-pima.ctree <- party::ctree(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.ctree)
-pimaIndiansTest(pima.ctree, prob = TRUE)
+```{r party_ctree}
+pima_ctree <- party::ctree(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_ctree)
+pima_test(pima_ctree, prob = TRUE)
 ```
@@
 
-```{r}
-pima.ctree2 <- partykit::ctree(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.ctree2)
-pimaIndiansTest(pima.ctree2, prob = TRUE)
+```{r partykit_ctree}
+pima_ctree2 <- partykit::ctree(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_ctree2)
+pima_test(pima_ctree2, prob = TRUE)
 ```
@@
 
-```{r}
-set.seed(101)
-pima.bagging <- adabag::bagging(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.bagging, quantiles = TRUE)
-pimaIndiansTest(pima.bagging, quantiles = TRUE, prob = TRUE)
+```{r adabag_bagging}
+set.seed(101)  # for reproducibility
+pima_bagging <- adabag::bagging(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_bagging, quantiles = TRUE)
+pima_test(pima_bagging, quantiles = TRUE, prob = TRUE)
 ```
@@
 
-```{r}
-set.seed(101)
-pima.ipred <- ipred::bagging(diabetes ~ ., data = na.omit(pima), nbagg = 500)
-pimaIndiansTest(pima.ipred, quantiles = TRUE)
-pimaIndiansTest(pima.ipred, quantiles = TRUE, prob = TRUE)
+```{r ipred_bagging}
+set.seed(101)  # for reproducibility
+pima_ipred <- ipred::bagging(diabetes ~ ., data = na.omit(pima), nbagg = 500)
+pima_test(pima_ipred, quantiles = TRUE)
+pima_test(pima_ipred, quantiles = TRUE, prob = TRUE)
 ```
@@
 
-```{r}
-set.seed(101)
-pima.rf <- randomForest::randomForest(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.rf)
-pimaIndiansTest(pima.rf, prob = TRUE)
+```{r randomForest_randomForest}
+set.seed(101)  # for reproducibility
+pima_rf <- randomForest::randomForest(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_rf)
+pima_test(pima_rf, prob = TRUE)
 ```
@@
 
-```{r}
-set.seed(101)
-pima.crf <- party::cforest(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.crf, quantiles = TRUE)
-pimaIndiansTest(pima.crf, quantiles = TRUE, prob = TRUE)
+```{r party_cforest}
+set.seed(101)  # for reproducibility
+pima_crf <- party::cforest(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_crf, quantiles = TRUE)
+pima_test(pima_crf, quantiles = TRUE, prob = TRUE)
 ```
@@
 ## Package: `partykit`
-```{r}
-set.seed(101)
-pima.crf2 <- partykit::cforest(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.crf2, quantiles = TRUE)
-pimaIndiansTest(pima.crf2, quantiles = TRUE, prob = TRUE)
+
+```{r partykit_cforest}
+set.seed(101)  # for reproducibility
+pima_crf2 <- partykit::cforest(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_crf2, quantiles = TRUE)
+pima_test(pima_crf2, quantiles = TRUE, prob = TRUE)
 ```
@@
 
-```{r}
-set.seed(101)
-pima.ranger <- ranger::ranger(diabetes ~ ., data = na.omit(pima), 
+```{r ranger_ranger}
+set.seed(101)  # for reproducibility
+pima_ranger <- ranger::ranger(diabetes ~ ., data = na.omit(pima), 
                               probability = TRUE)
-pimaIndiansTest(pima.ranger)
-pimaIndiansTest(pima.ranger, prob = TRUE)
+pima_test(pima_ranger)
+pima_test(pima_ranger, prob = TRUE)
 ```
@@
 
-```{r}
-set.seed(101)
-pima.boosting <- adabag::boosting(diabetes ~ ., data = na.omit(pima))
-pimaIndiansTest(pima.boosting, quantiles = TRUE)
-pimaIndiansTest(pima.boosting, quantiles = TRUE, prob = TRUE)
+```{r adabag_boosting}
+set.seed(101)  # for reproducibility
+pima_boosting <- adabag::boosting(diabetes ~ ., data = na.omit(pima))
+pima_test(pima_boosting, quantiles = TRUE)
+pima_test(pima_boosting, quantiles = TRUE, prob = TRUE)
 ```
@@
 
-```{r}
-set.seed(101)
-pima.gbm <- gbm::gbm(ifelse(diabetes == "neg", 1, 0) ~ ., 
+```{r gbm_gbm}
+set.seed(101)  # for reproducibility
+pima_gbm <- gbm::gbm(ifelse(diabetes == "neg", 1, 0) ~ ., 
                      data = na.omit(pima),
@@
                      verbose = FALSE)
-best.iter <- gbm::gbm.perf(pima.gbm, method = "OOB", plot.it = FALSE)
-pimaIndiansTest(pima.gbm, n.trees = best.iter)
-pimaIndiansTest(pima.gbm, n.trees = best.iter, prob = TRUE)
+best.iter <- gbm::gbm.perf(pima_gbm, method = "OOB", plot.it = FALSE)
+pima_test(pima_gbm, n.trees = best.iter)
+pima_test(pima_gbm, n.trees = best.iter, prob = TRUE)
 ```
@@
 ## Package: `xgboost`
-```{r}
-set.seed(101)
-pima.xgb <- xgboost::xgboost(
+
+```{r xgboost_xgboost}
+set.seed(101)  # for reproducibility
+pima_xgb <- xgboost::xgboost(
   data = data.matrix(subset(pima, select = -diabetes)),
@@
 )
-pimaIndiansTest(pima.xgb)
-pimaIndiansTest(pima.xgb, prob = TRUE)
+pima_test(pima_xgb, which.class = 2)
+pima_test(pima_xgb, prob = TRUE, which.class = 2)
 ```
@@
 ## Package: `nnet`
-```{r}
-set.seed(101)
-pima.nnet <- nnet::nnet(diabetes ~ ., data = na.omit(pima), size = 10, 
+
+```{r nnet_nnet}
+set.seed(101)  # for reproducibility
+pima_nnet <- nnet::nnet(diabetes ~ ., data = na.omit(pima), size = 10, 
                         decay = 0.1, maxit = 500, trace = FALSE)
-pimaIndiansTest(pima.nnet)
-pimaIndiansTest(pima.nnet, prob = TRUE)
+pima_test(pima_nnet)
+pima_test(pima_nnet, prob = TRUE)
 ```
@@
 
-```{r}
-pima.svm <- e1071::svm(diabetes ~ ., data = na.omit(pima), type = "C-classification", 
+```{r e1071_svm}
+pima_svm <- e1071::svm(diabetes ~ ., data = na.omit(pima), type = "C-classification", 
                        probability = TRUE)
-pimaIndiansTest(pima.svm)
-pimaIndiansTest(pima.svm, prob = TRUE)
+pima_test(pima_svm)
+pima_test(pima_svm, prob = TRUE)
 ```
@@
 
-```{r}
-pima.ksvm <- kernlab::ksvm(diabetes ~ ., data = na.omit(pima), type = "C-svc", 
+```{r kernlab_svm}
+pima_ksvm <- kernlab::ksvm(diabetes ~ ., data = na.omit(pima), type = "C-svc", 
                            prob.model = TRUE)
-pimaIndiansTest(pima.ksvm)
-pimaIndiansTest(pima.ksvm, prob = TRUE)
+pima_test(pima_ksvm)
+pima_test(pima_ksvm, prob = TRUE)
 ```
@@
 
-```{r}
-set.seed(101)
-pima.multinom <- nnet::multinom(diabetes ~ . ^ 2, data = na.omit(pima), 
+```{r nnet_multinom}
+set.seed(101)  # for reproducibility
+pima_multinom <- nnet::multinom(diabetes ~ . ^ 2, data = na.omit(pima), 
                                 trace = FALSE)
-pimaIndiansTest(pima.multinom)
-pimaIndiansTest(pima.multinom, prob = TRUE)
+pima_test(pima_multinom)
+pima_test(pima_multinom, prob = TRUE)
 ```

### 2018-08-27 R/plotPartial.R

@@ -457,7 +457,7 @@ plot_two_predictor_pdp <- function(
 
         # Draw a wireframe plot
         wireframe(
-          form, data = object, col.regions = col.regions, drape = TRUE, ...
+          form, data = object, col.regions = col.regions, ...
         )
 
       }
@@ -565,7 +565,7 @@ plot_three_predictor_pdp <- function(
 
       # Draw a wireframe plot
       wireframe(
-        form, data = object, col.regions = col.regions, drape = TRUE, ...
+        form, data = object, col.regions = col.regions, ...
       )
 
     }

### 2018-08-27 R/super_type.R

@@ -42,12 +42,10 @@ super_type.C5.0 <- function(object) {
 
 
 super_type.cforest <- function(object) {
-  if (attr(object$terms, "response") == 1 &&
-      attr(object$terms, "dataClasses")[1L] == "numeric") {
-    "regression"
-  } else if (attr(object$terms, "response") == 1 &&
-             attr(object$terms, "dataClasses")[1L] == "factor") {
+  if (is.factor(object$fitted[["(response)"]])) {
     "classification"
+  } else if (is.numeric(object$fitted[["(response)"]])) {
+    "regression"
   } else {
     "other"
   }
@@ -127,7 +125,7 @@ super_type.ksvm <- function(object) {
 
 
 #' @keywords internal
-super_type.lda<- function(object) {
+super_type.lda <- function(object) {
   "classification"
 }
 
@@ -178,12 +176,10 @@ super_type.nnet <- function(object) {
 
 
 super_type.party <- function(object) {
-  if (attr(object$terms, "response") == 1 &&
-      attr(object$terms, "dataClasses")[1L] == "numeric") {
-    "regression"
-  } else if (attr(object$terms, "response") == 1 &&
-             attr(object$terms, "dataClasses")[1L] == "factor") {
+  if (is.factor(object$fitted[["(response)"]])) {
     "classification"
+  } else if (is.numeric(object$fitted[["(response)"]])) {
+    "regression"
   } else {
     "other"
   }

### 2018-08-27 slowtests/slowtests-autoplot.R

@@ -38,24 +38,24 @@ boston.rf %>%
 # ICE curves (numeric)
 boston.rf %>%
   partial(pred.var = "lstat", ice = TRUE) %>%
-  autoplot(alpha = 0.3, main = "ICE curves")
+  autoplot(alpha = 0.1, main = "ICE curves")
 
 # c-ICE curves (numeric)
 boston.rf %>%
   partial(pred.var = "lstat", ice = TRUE) %>%
-  autoplot(center = TRUE, alpha = 0.3, main = "c-ICE curves")
+  autoplot(center = TRUE, alpha = 0.1, main = "c-ICE curves")
 
 # c-ICE curves (numeric)
 boston.rf %>%
   partial(pred.var = "lstat", ice = TRUE, center = TRUE) %>%
-  autoplot(alpha = 0.3, main = "c-ICE curves")
+  autoplot(alpha = 0.1, main = "c-ICE curves")
 
 # ICE curves (factor)
 boston.rf %>%
   partial(pred.var = "chas", ice = TRUE) %>%
-  autoplot(alpha = 0.3, main = "ICE curves")
+  autoplot(alpha = 0.1, main = "ICE curves")
 
 # c-ICE curves (factor)
 boston.rf %>%
   partial(pred.var = "chas", ice = TRUE, center = TRUE) %>%
-  autoplot(alpha = 0.3, main = "ICE curves")
+  autoplot(alpha = 0.1, main = "ICE curves")

### 2018-08-27 slowtests/slowtests-calhousing.R

@@ -5,36 +5,22 @@
 # Description: Reproducing the California housing example from Section 14.1 of
 # Chapter 10 in
 #
-#   Hastie, Trevor, Robert Tibshirani, and J. H. Friedman. The elements of 
-#   statistical learning : data mining, inference, and prediction. New York: 
+#   Hastie, Trevor, Robert Tibshirani, and J. H. Friedman. The elements of
+#   statistical learning : data mining, inference, and prediction. New York:
 #   Springer, 2009. Print.
 #
-# WARNING: This is simply a test file. These models are not trained to be 
+# WARNING: This is simply a test file. These models are not trained to be
 # "optimal" in any sense.
 #
 #-------------------------------------------------------------------------------
-
-
-################################################################################
-# Setup
-################################################################################
 
 # Load required packages
 library(gbm)
 library(pdp)
 
-# Download function from gist
-url <- "b1330460eec5acf1c81fae71902e331c"
-devtools::source_gist(url, filename = "fetchCaliforniaHousingData.R")
-
 # Fetch California housing data
-cal <- fetchCaliforniaHousingData()
+cal <- read.csv("slowtests/california_housing.csv")
 head(cal)
-
-
-################################################################################
-# Fit models
-################################################################################
 
 # Fit a GBM with the same parameters as in Hastie et al. (2009, pp 371-375)
 set.seed(102)
@@ -48,13 +34,8 @@ cal.gbm <- gbm(AvgValue ~ ., data = cal,
                verbose = TRUE)
 best.iter <- gbm.perf(cal.gbm, method = "test")
 
-
-################################################################################
-# Construct and display partial dependence plots
-################################################################################
-
 # There seems to be a fair amount of outliers in the datset so we have two
-# options: 
+# options:
 #
 #   (1) use the quantiles options with specific probabilities;
 #   (2) use the trim.outliers option.
@@ -92,8 +73,10 @@ grid.arrange(
 grid.arrange(
   plotPartial(pd.HouseAge.AveOccup, levelplot = FALSE, zlab = "",
               scales = list(arrows = FALSE),
+              col.regions = viridis::magma(100, begin = 0.4),
+              # palette = "magma",
               drape = TRUE, colorkey = TRUE,
-              screen = list(z = 120, x = -60)),
+              screen = list(z = 140, x = -60)),
   plotPartial(pd.HouseAge.AveOccup),
   ncol = 2
 )

### 2018-08-27 slowtests/slowtests-mtcars.R

@@ -46,6 +46,7 @@ p4 <- bst %>%
 # Compare plots
 grid.arrange(p1, p2, p3, p4, ncol = 2)
 
+# Fit a gbm model
 set.seed(101)
 mtcars.gbm <- gbm(carb ~ ., data = mtcars, distribution = "poisson",
                   n.trees = 2000, interaction.depth = 3, shrinkage = 0.01,
@@ -54,17 +55,21 @@ best.iter <- gbm.perf(mtcars.gbm, method = "cv")
 summary(mtcars.gbm, n.trees = best.iter, las = 1)
 
 # Log scale
-plot(mtcars.gbm, i.var = "mpg")
-partial(mtcars.gbm, pred.var = "mpg", n.trees = best.iter, plot = TRUE)
+plot(mtcars.gbm, i.var = "mpg", main = "Log scale")
+partial(mtcars.gbm, pred.var = "mpg", n.trees = best.iter, plot = TRUE,
+        plot.engine = "ggplot2") +
+  ggtitle("Log scale")
 
 # Response scale
 par(mfrow = c(1, 2))
-plot(mtcars.gbm, i.var = "mpg", type = "link")
-plot(mtcars.gbm, i.var = "mpg", type = "response")
+plot(mtcars.gbm, i.var = "mpg", type = "link", main = "link")
+plot(mtcars.gbm, i.var = "mpg", type = "response", main = "response")
 grid.arrange(
   partial(mtcars.gbm, pred.var = "mpg", recursive = FALSE,
-          n.trees = best.iter, plot = TRUE),
+          n.trees = best.iter, plot = TRUE, plot.engine = "ggplot2") +
+    ggtitle("Link scale"),
   partial(mtcars.gbm, pred.var = "mpg", recursive = FALSE, inv.link = exp,
-          n.trees = best.iter, plot = TRUE),
+          n.trees = best.iter, plot = TRUE, plot.engine = "ggplot2") +
+    ggtitle("Response scale"),
   ncol = 2
 )

### 2018-08-27 slowtests/slowtests-plotPartial.R

@@ -37,24 +37,24 @@ boston.rf %>%
 # ICE curves (numeric)
 boston.rf %>%
   partial(pred.var = "lstat", ice = TRUE) %>%
-  plotPartial(alpha = 0.3, main = "ICE curves")
+  plotPartial(alpha = 0.1, main = "ICE curves")
 
 # c-ICE curves (numeric)
 boston.rf %>%
   partial(pred.var = "lstat", ice = TRUE) %>%
-  plotPartial(center = TRUE, alpha = 0.3, main = "c-ICE curves")
+  plotPartial(center = TRUE, alpha = 0.1, main = "c-ICE curves")
 
 # c-ICE curves (numeric)
 boston.rf %>%
   partial(pred.var = "lstat", ice = TRUE, center = TRUE) %>%
-  plotPartial(alpha = 0.3, main = "c-ICE curves")
+  plotPartial(alpha = 0.1, main = "c-ICE curves")
 
 # ICE curves (factor)
 boston.rf %>%
   partial(pred.var = "chas", ice = TRUE) %>%
-  plotPartial(alpha = 0.3, main = "ICE curves")
+  plotPartial(alpha = 0.1, main = "ICE curves")
 
 # c-ICE curves (factor)
 boston.rf %>%
   partial(pred.var = "chas", ice = TRUE, center = TRUE) %>%
-  plotPartial(alpha = 0.3, main = "ICE curves")
+  plotPartial(alpha = 0.1, main = "ICE curves")

### 2018-08-27 slowtests/slowtests-xgboost.R

@@ -6,18 +6,13 @@
 # formats:
 #
 #   (1) "matrix" - base R;
-#   (2) "dgCMatrix" (sparce matrix format from Matrix pkg);
-#   (3) "xgb.DMatrix" (XGBoost matrix format).
+#   (2) "dgCMatrix" - sparse matrix format from package Matrix;
+#   (3) "xgb.DMatrix" - XGBoost matrix format.
 #
 # WARNING: This is simply a test file. These models are not trained to be
 # "optimal" in any sense.
 #
 #-------------------------------------------------------------------------------
-
-
-################################################################################
-# Setup
-################################################################################
 
 # Load required packages
 library(ggplot2)
@@ -26,12 +21,7 @@ library(pdp)
 library(xgboost)
 
 # Load the data
-data(pima)
-
-
-################################################################################
-# Fit models
-################################################################################
+data(pima)  # xgboost can handle missing values, so no need for na.omit()
 
 # Set up training data
 X <- subset(pima, select = -diabetes)
@@ -61,11 +51,6 @@ bst.dgCMatrix <- xgboost(data = X.dgCMatrix, label = y, params = plist,
 set.seed(101)
 bst.xgb.DMatrix <- xgboost(data = xgb.DMatrix(data.matrix(X), label = y),
                            params = plist, nrounds = 100, save_period = NULL)
-
-
-################################################################################
-# Construct and display partial dependence plots
-################################################################################
 
 # Function to construct a PDP for glucose on the probability scale
 parDepPlot <- function(object, train, ...) {

### 2018-08-27 R/autoplot.R
cleanup

@@ -113,27 +113,12 @@
 #' )
 #' }
 autoplot.partial <- function(
-  object,
-  center = FALSE,
-  plot.pdp = TRUE,
-  pdp.color = "red",
-  pdp.size = 1,
-  pdp.linetype = 1,
-  rug = FALSE,
-  smooth = FALSE,
-  smooth.method = "auto",
-  smooth.formula = y ~ x,
-  smooth.span = 0.75,
-  smooth.method.args = list(),
-  contour = FALSE,
-  contour.color = "white",
-  palette = c("viridis", "magma", "inferno", "plasma", "cividis"),
-  train = NULL,
-  xlab = NULL,
-  ylab = NULL,
-  main = NULL,
-  legend.title = "yhat",
-  ...
+  object, center = FALSE, plot.pdp = TRUE, pdp.color = "red", pdp.size = 1,
+  pdp.linetype = 1, rug = FALSE, smooth = FALSE, smooth.method = "auto",
+  smooth.formula = y ~ x, smooth.span = 0.75, smooth.method.args = list(),
+  contour = FALSE, contour.color = "white",
+  palette = c("viridis", "magma", "inferno", "plasma", "cividis"), train = NULL,
+  xlab = NULL, ylab = NULL, main = NULL, legend.title = "yhat", ...
 ) {
 
   # Determine if object contains an ID column (i.e., multiple curves)
@@ -194,18 +179,9 @@ autoplot.partial <- function(
 #'
 #' @export
 autoplot.ice <- function(
-  object,
-  center = FALSE,
-  plot.pdp = TRUE,
-  pdp.color = "red",
-  pdp.size = 1,
-  pdp.linetype = 1,
-  rug = FALSE,
-  train = NULL,
-  xlab = NULL,
-  ylab = NULL,
-  main = NULL,
-  ...
+  object, center = FALSE, plot.pdp = TRUE, pdp.color = "red", pdp.size = 1,
+  pdp.linetype = 1, rug = FALSE, train = NULL, xlab = NULL, ylab = NULL,
+  main = NULL, ...
 ) {
 
   # Call workhorse function
@@ -222,17 +198,8 @@ autoplot.ice <- function(
 #'
 #' @export
 autoplot.cice <- function(
-  object,
-  plot.pdp = TRUE,
-  pdp.color = "red",
-  pdp.size = 1,
-  pdp.linetype = 1,
-  rug = FALSE,
-  train = NULL,
-  xlab = NULL,
-  ylab = NULL,
-  main = NULL,
-  ...
[…]

### 2018-08-27 R/plotPartial.R

@@ -119,16 +119,8 @@ plotPartial <- function(object, ...) {
 #'
 #' @export
 plotPartial.ice <- function(
-  object,
-  center = FALSE,
-  plot.pdp = TRUE,
-  pdp.col = "red2",
-  pdp.lwd = 2,
-  pdp.lty = 1,
-  rug = FALSE,
-  train = NULL,
-  alpha = 1,
-  ...
+  object, center = FALSE, plot.pdp = TRUE, pdp.col = "red2", pdp.lwd = 2,
+  pdp.lty = 1, rug = FALSE, train = NULL, alpha = 1, ...
 ) {
 
   # Call workhorse function
@@ -145,15 +137,8 @@ plotPartial.ice <- function(
 #'
 #' @export
 plotPartial.cice <- function(
-  object,
-  plot.pdp = TRUE,
-  pdp.col = "red2",
-  pdp.lwd = 2,
-  pdp.lty = 1,
-  rug = FALSE,
-  train = NULL,
-  alpha = 1,
-  ...
+  object, plot.pdp = TRUE, pdp.col = "red2", pdp.lwd = 2, pdp.lty = 1,
+  rug = FALSE, train = NULL, alpha = 1, ...
 ) {
 
   # Call workhorse function
@@ -170,25 +155,11 @@ plotPartial.cice <- function(
 #'
 #' @export
 plotPartial.partial <- function(
-  object,
-  center = FALSE,
-  plot.pdp = TRUE,
-  pdp.col = "red2",
-  pdp.lwd = 2,
-  pdp.lty = 1,
-  smooth = FALSE,
-  rug = FALSE,
-  chull = FALSE,
-  levelplot = TRUE,
-  contour = FALSE,
-  contour.color = "white",
-  col.regions = NULL,
+  object, center = FALSE, plot.pdp = TRUE, pdp.col = "red2", pdp.lwd = 2,
+  pdp.lty = 1, smooth = FALSE, rug = FALSE, chull = FALSE, levelplot = TRUE,
+  contour = FALSE, contour.color = "white", col.regions = NULL,
   palette = c("viridis", "magma", "inferno", "plasma", "cividis"),
-  alpha = 1,
-  number = 4,
-  overlap = 0.1,
-  train = NULL,
-  ...
+  alpha = 1, number = 4, overlap = 0.1, train = NULL, ...
 ) {
 
   # Determine if object contains multiple curves
[…]

### 2019-10-27 R/exemplar.R
major changes; see TODO.md and NEWS.md

@@ -1,0 +1,60 @@
+#' Exemplar observation
+#'
+#' Construct an exemplar record from a data frame. For now, all numeric columns
+#' (including \code{"\link{Date}"} objects) are replaced with their
+#' corresponding median value and non-numeric columns are replaced with their
+#' most frequent value.
+#'
+#' @param object A data frame.
+#'
+#' @return A data frame with the same number of columns as \code{object} and a
+#' single row.
+#'
+#' @rdname exemplar
+#'
+#' @export
+#'
+#' @examples
+#' set.seed(1554)  # for reproducibility
+#' train <- data.frame(
+#'   x = rnorm(100),
+#'   y = sample(letters[1L:3L], size = 100, replace = TRUE,
+#'              prob = c(0.1, 0.1, 0.8))
+#' )
+#' exemplar(train)
+exemplar <- function(object) {
+  UseMethod("exemplar")
+}
+
+
+#' @rdname exemplar
+#'
+#' @export
+exemplar.data.frame <- function(object) {
+  res <- as.data.frame(lapply(object, FUN = function(x) {
+    if (inherits(x, what = c("numeric", "integer", "Date"))) {
+      stats::median(x, na.rm = TRUE)
+    } else {
+      names(which.max(table(x, useNA = "no")))
+    }
+  }))
+  copy_classes(res, y = object)  # make sure column types are the same
+}
+
+
+#' @rdname exemplar
+#'
+#' @export
+exemplar.matrix <- function(object) {
+  # FIXME: For now, just use the median rounded up to the nearest integer
+  res <- ceiling(apply(object, MARGIN = 2, FUN = stats::median))
+  data.matrix(t(res))
+}
+
+#' @rdname exemplar
+#'
+#' @export
+exemplar.dgCMatrix <- function(object) {
+  res <- exemplar.matrix(data.matrix(object))
+  methods::as(data.matrix(res), Class = "dgCMatrix")
+}

### 2019-10-27 R/getParPredCls.R

@@ -269,10 +269,10 @@ getParClsProb.RandomForest <- function(object, newdata, which.class, ...) {
 
 #' @keywords internal
 getParClsLogit.ranger <- function(object, newdata, which.class, ...) {
-  if (object$treetype != "Probability estimation") {
-    stop(paste("Cannot obtain predicted probabilities from",
-               deparse(substitute(object))))
-  }
+  # if (object$treetype != "Probability estimation") {
+  #   stop(paste("Cannot obtain predicted probabilities from",
+  #              deparse(substitute(object))))
+  # }
   pr <- stats::predict(object, data = newdata, ...)$predictions
   mean(multiclass_logit(pr, which.class = which.class), na.rm = TRUE)
 }
@@ -280,10 +280,10 @@ getParClsLogit.ranger <- function(object, newdata, which.class, ...) {
 
 #' @keywords internal
 getParClsProb.ranger <- function(object, newdata, which.class, ...) {
-  if (object$treetype != "Probability estimation") {
-    stop(paste("Cannot obtain predicted probabilities from",
-               deparse(substitute(object))))
-  }
+  # if (object$treetype != "Probability estimation") {
+  #   stop(paste("Cannot obtain predicted probabilities from",
+  #              deparse(substitute(object))))
+  # }
   pr <- stats::predict(object, data = newdata, ...)$predictions
   mean(pr[, which.class], na.rm = TRUE)
 }
@@ -291,10 +291,10 @@ getParClsProb.ranger <- function(object, newdata, which.class, ...) {
 
 #' @keywords internal
 getParClsLogit.svm <- function(object, newdata, which.class, ...) {
-  if (is.null(object$call$probability)) {
-    stop(paste("Cannot obtain predicted probabilities from",
-               deparse(substitute(object))))
-  }
+  # if (is.null(object$call$probability)) {
+  #   stop(paste("Cannot obtain predicted probabilities from",
+  #              deparse(substitute(object))))
+  # }
   pr <- attr(stats::predict(object, newdata = newdata, probability = TRUE, ...),
              which = "probabilities")
   mean(multiclass_logit(pr, which.class = which.class), na.rm = TRUE)
@@ -303,10 +303,10 @@ getParClsLogit.svm <- function(object, newdata, which.class, ...) {
 
 #' @keywords internal
 getParClsProb.svm <- function(object, newdata, which.class, ...) {
-  if (is.null(object$call$probability)) {
[…]

### 2019-10-27 R/get_task.R

@@ -1,18 +1,18 @@
 #' @keywords internal
-super_type <- function(object) {
-  UseMethod("super_type")
+get_task <- function(object) {
+  UseMethod("get_task")
 }
 
 
 #' @keywords internal
-super_type.default <- function(object) {
+get_task.default <- function(object) {
   warning('`type` could not be determined; assuming `type = "regression"`')
   "regression"
 }
 
 
 #' @keywords internal
-super_type.BinaryTree <- function(object) {
+get_task.BinaryTree <- function(object) {
   if (object@responses@is_nominal) {
     "classification"
   } else if (object@responses@is_ordinal || object@responses@is_censored) {
@@ -24,24 +24,24 @@ super_type.BinaryTree <- function(object) {
 
 
 #' @keywords internal
-super_type.bagging <- function(object) {
+get_task.bagging <- function(object) {
   "classification"
 }
 
 
 #' @keywords internal
-super_type.boosting <- function(object) {
+get_task.boosting <- function(object) {
   "classification"
 }
 
 
 #' @keywords internal
-super_type.C5.0 <- function(object) {
+get_task.C5.0 <- function(object) {
   "classification"
 }
 
 
-super_type.cforest <- function(object) {
+get_task.cforest <- function(object) {
   if (is.factor(object$fitted[["(response)"]])) {
     "classification"
   } else if (is.numeric(object$fitted[["(response)"]])) {
@@ -53,19 +53,19 @@ super_type.cforest <- function(object) {
 
 
 #' @keywords internal
-super_type.classbagg <- function(object) {
+get_task.classbagg <- function(object) {
   "classification"
 }
 
 
 #' @keywords internal
-super_type.cubist <- function(object) {
+get_task.cubist <- function(object) {
   "regression"
 }
 
 
 #' @keywords internal
-super_type.earth <- function(object) {
+get_task.earth <- function(object) {
   if (!is.null(object$glm.list) &&
       object$glm.list[[1L]]$family$family == "binomial") {
     "classification"
@@ -80,13 +80,13 @@ super_type.earth <- function(object) {
 
 
 #' @keywords internal
-super_type.fda<- function(object) {
+get_task.fda<- function(object) {
   "classification"
 }
 
 
 #' @keywords internal
-super_type.gbm <- function(object) {
+get_task.gbm <- function(object) {
   if (object$distribution %in%
[…]

### 2019-10-27 R/get_training_data.R

@@ -1,8 +1,20 @@
 # Error message to display when training data cannot be extracted form object
-mssg <- paste0(
+msg <- paste0(
   "The training data could not be extracted from object. Please supply the ",
-  "raw training data using the `train` argument in the call to `partial`."
+  "raw training data using the `train` argument in the call to `partial()`."
 )
+
+
+#' #' @keywords internal
+#' where <- function(name, env = parent.frame()) {
+#'   if (identical(env, emptyenv())) {
+#'     stop("Can't find ", name, ".", call. = FALSE)
+#'   } else if (exists(name, envir = env, inherits = FALSE)) {
+#'     env
+#'   } else {
+#'     where(name, parent.env(env))
+#'   }
+#' }
 
 
 #' @keywords internal
@@ -12,59 +24,154 @@ get_training_data <- function(object) {
 
 
 #' @keywords internal
-get_training_data.default <- function(object) {
+get_training_data.default <- function(object, env = parent.frame(),
+                                      arg = "data") {
+
+  # FIXME: This function does not work on objects fit using the :: operator.
+  # For example,
+  #
+  # > fit <- randomForest::randomForest(cmedv ~ ., data = boston)
+  # > pdp:::get_training_data.default(fit)
+  #
+  #  Error in eval(mcall[[1L]], envir = env) : object 'randomForest' not found
+  #
+  #   3. eval(mcall[[1L]], envir = env)
+  #   2. eval(mcall[[1L]], envir = env) at get_training_data.R#35
+  #   1. pdp:::get_training_data.default(fit)
+
+  # Throw error message for S4 objects (for now)
   if (isS4(object)) {
-    stop(mssg)
+    stop(msg, call. = FALSE)
+  }
+
+  # Grab the call
+  mcall <- tryCatch(stats::getCall(object), error = function(e) {
+    stop(msg, call. = FALSE)
+  })
+
+  # If data component of the call is NULL, then try to make sure each
+  # component is named before proceeding (taken from Advanced R, 2nd ed.)
+  if (is.null(mcall[[arg]])) {
+    f <- tryCatch(eval(mcall[[1L]], envir = env), error = function(e) {
+      stop(msg, call. = FALSE)
+    })
+    if (!is.primitive(f)) {
+      mcall <- match.call(f, call = mcall)
+    }
+  }
+
+  # Grab the data component (if it exists)
+  n <- 1
+  while(length(env) != 0) {
[…]

### 2019-10-27 R/partial.R

@@ -1,3 +1,5 @@
+# FIXME: partial() returns character columns for factors.
+
 #' Partial Dependence Functions
 #'
 #' Compute partial dependence functions (i.e., marginal effects) for various
@@ -31,6 +33,18 @@
 #' @param center Logical indicating whether or not to produce centered ICE
 #' curves (c-ICE curves). Only used when \code{ice = TRUE}. Default is
 #' \code{FALSE}. See Goldstein et al. (2014) for details.
+#'
+#' @param approx Logical indicating whether or not to compute a faster, but
+#' approximate, marginal effect plot (similar in spirit to
+#' \code{\link[plotmo]{plotmo}}). If \code{TRUE}, then `partial()` will compute
+#' predictions across the predictors specified in \code{pred.var} while holding
+#' the other predictors constant (a "poor man's partial dependence" function as
+#' Stephen Milborrow, the author of \code{\link[plotmo]{plotmo}}, puts it).
+#' Default is \code{FALSE}. Note this works with \code{ice = TRUE} as well.
+#' WARNING: This option is currently experimental. Use at your own risk. It is
+#' possible (and arguably safer) to do this manually by passing a specific
+#' "exemplar" observation to the train argument and specifying `pred.grid`
+#' manually.
 #'
 #' @param quantiles Logical indicating whether or not to use the sample
 #' quantiles of the continuous predictors listed in \code{pred.var}. If
@@ -270,7 +284,7 @@ partial <- function(object, ...) {
 #' @export
 partial.default <- function(
   object, pred.var, pred.grid, pred.fun = NULL, grid.resolution = NULL,
-  ice = FALSE, center = FALSE, quantiles = FALSE, probs = 1:9/10,
+  ice = FALSE, center = FALSE, approx = FALSE, quantiles = FALSE, probs = 1:9/10,
   trim.outliers = FALSE, type = c("auto", "regression", "classification"),
   inv.link = NULL, which.class = 1L, prob = FALSE, recursive = TRUE,
   plot = FALSE, plot.engine = c("lattice", "ggplot2"),
@@ -375,13 +389,18 @@ partial.default <- function(
   # Determine the type of supervised learning used
   type <- match.arg(type)
   if (type == "auto" && is.null(pred.fun)) {
-    type <- super_type(object)  # determine if regression or classification
+    type <- get_task(object)  # determine if regression or classification
   }
 
[…]

### 2019-10-27 R/pred_grid.R

@@ -22,7 +22,8 @@ pred_grid <- function(
   pred.val <- lapply(pred.var, function(x) {
     if (is.factor(train[, x, drop = TRUE])) {
       levels(train[, x, drop = TRUE])
-    } else if (x %in% cats) {
+    } else if (inherits(train[, x, drop = TRUE], what = "character") ||
+               (x %in% cats)) {
       sort(unique(train[, x, drop = TRUE]))  # martices cannot contain factors
     } else {
       if (!is.null(grid.resolution) && quantiles) {

### 2019-10-27 R/utils.R

@@ -115,11 +115,11 @@ copy_classes <- function(x, y) {
       # Convert to factor or ordered class
       if (is.factor(y[[name]])) {
         if (is.ordered(y[[name]])) {
-          x[[name]] <- as.ordered(x[[name]])
+          x[[name]] <- ordered(x[[name]], levels = levels(y[[name]]))
         } else {
-          x[[name]] <- as.factor(x[[name]])
+          x[[name]] <- factor(x[[name]], levels = levels(y[[name]]))
         }
-        levels(x[[name]]) <- levels(y[[name]])
+        # levels(x[[name]]) <- levels(y[[name]])
         # if (!all(levels(y[[name]]) %in% x[[name]])) {
         #   stop("Factors levels ", paste0("{", paste(
         #     levels(y[[name]])[!(levels(y[[name]]) %in% x[[name]])],
@@ -160,12 +160,13 @@ multiclass_logit <- function(x, which.class = 1L) {
     rowMeans(log(ifelse(x > 0, x, eps)))
 }
 
+
 #' @keywords internal
 train_chull <- function(pred.var, pred.grid, train) {
   if (length(pred.var) >= 2 && is.numeric(pred.grid[, 1L]) &&
       is.numeric(pred.grid[, 2L])) {  # if the first two columns are numeric
     if (is.data.frame(train)) {
-      train <- data.matrix(train)  # mgcv::in.out requires a matrix
+      train <- data.matrix(train)  # `mgcv::in.out()` requires a matrix
     }
     X <- stats::na.omit(train[, pred.var[1L:2L]])
     Y <- stats::na.omit(data.matrix(pred.grid[, 1L:2L]))

### 2019-10-27 inst/tinytest/test_exemplar.R

@@ -1,0 +1,19 @@
+# Generate some sample data
+df1 <- data.frame(
+  x = sample(1L:3L, size = 100, replace = TRUE, prob = c(0.1, 0.1, 0.8)),
+  y = sample(letters[1L:3L], size = 100, replace = TRUE,
+             prob = c(0.1, 0.1, 0.8)),
+  z = sample(c(Sys.Date(), Sys.Date() + 1, Sys.Date() + 2), size = 100,
+             replace = TRUE, prob = c(0.1, 0.1, 0.8))
+)
+
+# Data frame that pdp::exemplar() should return
+df2 <- data.frame(
+  x = 3L,
+  y = factor("c", levels = letters[1L:3L]),
+  z = Sys.Date() + 2
+)
+
+# Expectation(s)
+df3 <- exemplar(df1)
+expect_equal(exemplar(df1), target = df2)

### 2019-10-27 inst/tinytest/test_partial.R

@@ -1,0 +1,178 @@
+if (require(gbm, quietly = TRUE) && require(ggplot2, quietly = TRUE)) {
+
+  # Load the Ames housing data
+  ames <- AmesHousing::make_ames()
+
+  # Fit a gbm model
+  set.seed(1918)  # for reproducibility
+  fit <- gbm(
+    Sale_Price ~ .,
+    data = ames,
+    distribution = "gaussian",
+    n.trees = 1500,
+    interaction.depth = 6,
+    shrinkage = 0.1,
+    bag.fraction = 1,
+    cv.folds = 5
+  )
+  best_iter <- gbm.perf(fit, method = "cv")
+
+  # Partial dependence w/wo setting approx = TRUE
+  pd1 <- partial(fit, pred.var = "Gr_Liv_Area", n.trees = best_iter,
+                 recursive = FALSE, progress = "progress")
+  pd2 <- partial(fit, pred.var = "Gr_Liv_Area", approx = TRUE,
+                 recursive = FALSE, progress = "progress",
+                 n.trees = best_iter)
+
+  # Basic expectation(s)
+  expect_true(inherits(pd1, what = "partial"))
+  expect_true(inherits(pd2, what = "partial"))
+  expect_identical(dim(pd1), target = c(51L, 2L))
+  expect_identical(dim(pd2), target = c(51L, 2L))
+
+  # Display plots side by side
+  grid.arrange(
+    plotPartial(pd1),
+    plotPartial(pd2),
+    nrow = 1
+  )
+
+  # Users should be able to supply the predictor name(s) or column position(s)
+  # in train
+  expect_identical(
+    current = partial(fit, pred.var = "Gr_Liv_Area", n.trees = best_iter),
+    target = partial(fit, pred.var = which(names(ames) == "Gr_Liv_Area"),
+                     n.trees = best_iter)
+  )
+
+  # An error should be thrown whenever pred.var is not found in train
+  expect_error(partial(fit, pred.var = "ABC", n.trees = best_iter))
+  expect_error(partial(fit, pred.var = 100, n.trees = best_iter))
+
+  # Users can supply their own predictor grid
+  grid1 <- data.frame("Gr_Liv_Area" = range(ames$Gr_Liv_Area))
+  grid2 <- data.frame("x" = range(ames$Gr_Liv_Area))
+  pd_grid1 <- partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1,
[…]

### 2019-10-27 inst/tinytest/test_pkg_e1071.R

@@ -1,0 +1,34 @@
+# Load Friedman benchmark data
+friedman1 <- readRDS("friedman.rds")$friedman1
+friedman2 <- readRDS("friedman.rds")$friedman2
+X <- friedman1[, paste0("x.", 1L:10L)]
+
+# Tests for package e1071
+if (require(e1071, quietly = TRUE)) {
+
+  # Fit model(s)
+  fit1 <- svm(y ~ ., data = friedman1)
+  fit2 <- svm(x = X, y = friedman1$y)
+  fit3 <- svm(y ~ ., data = friedman2, probability = TRUE)
+  fit4 <- svm(y ~ ., data = friedman2)
+
+  # Compute partial dependence for x.3
+  pd1 <- partial(fit1, pred.var = "x.3")
+  pd2 <- partial(fit2, pred.var = "x.3")
+  pd3 <- partial(fit3, pred.var = "x.3")
+  pd4 <- partial(fit3, pred.var = "x.3", prob = TRUE)
+  ice1 <- partial(fit1, pred.var = "x.3", ice = TRUE)
+  ice2 <- partial(fit3, pred.var = "x.3", ice = TRUE)
+  ice3 <- partial(fit3, pred.var = "x.3", ice = TRUE, prob = TRUE)
+
+  # Expectations: partial()
+  expect_true(inherits(pd1, what = "partial"))
+  expect_true(inherits(pd2, what = "partial"))
+  expect_true(inherits(pd3, what = "partial"))
+  expect_true(inherits(pd4, what = "partial"))
+  expect_true(inherits(ice1, what = "ice"))
+  expect_true(inherits(ice2, what = "ice"))
+  expect_true(inherits(ice3, what = "ice"))
+  expect_error(partial(fit4, pred.var = "x.3"))
+
+}

### 2019-10-27 inst/tinytest/test_pkg_gbm.R

@@ -1,0 +1,84 @@
+# Load required packages
+library(ggplot2)
+
+# Load Friedman benchmark data
+friedman1 <- readRDS("friedman.rds")$friedman1
+friedman2 <- readRDS("friedman.rds")$friedman2
+friedman3 <- readRDS("friedman.rds")$friedman3
+friedman3$x.5 <- cut(friedman3$x.5, breaks = 10)
+
+# Tests for package gbm
+if (require(gbm, quietly = TRUE)) {
+
+  # Fit model(s)
+  fit1 <- gbm(
+    formula = y ~ .,
+    data = friedman1,
+    shrinkage = 0.01,
+    n.trees = 500,
+    interaction.depth = 2
+  )
+  fit2 <- gbm(
+    formula = ifelse(y == "class1", 1, 0) ~ .,
+    data = friedman2,
+    distribution = "bernoulli",
+    shrinkage = 0.01,
+    n.trees = 500,
+    interaction.depth = 2
+  )
+  fit3 <- gbm(
+    formula = y ~ .,
+    data = friedman3,
+    distribution = "multinomial",
+    shrinkage = 0.01,
+    n.trees = 500,
+    interaction.depth = 2
+  )
+  iter1 <- gbm.perf(fit1, plot.it = FALSE, method = "OOB")
+  iter2 <- gbm.perf(fit2, plot.it = FALSE, method = "OOB")
+  iter3 <- gbm.perf(fit3, plot.it = FALSE, method = "OOB")
+
+  # Compute partial dependence, ICE, etc. for x.3
+  pd1 <- partial(fit1, pred.var = "x.3", n.trees = iter1)
+  pd1.brute <- partial(fit1, pred.var = "x.3", n.trees = iter1,
+                       recursive = FALSE)
+  pd2 <- partial(fit2, pred.var = "x.3", n.trees = iter2)
+  pd2.prob <- partial(fit2, pred.var = "x.3", prob = TRUE, n.trees = iter1)
+  ice1 <- partial(fit1, pred.var = "x.3", n.trees = iter1, ice = TRUE)
+  ice1.brute <- partial(fit1, pred.var = "x.3", n.trees = iter1,
+                        recursive = FALSE, ice = TRUE)
+  ice2 <- partial(fit2, pred.var = "x.3", n.trees = iter2, ice = TRUE)
+  ice2.prob <- partial(fit2, pred.var = "x.3", prob = TRUE, n.trees = iter1,
+                       ice = TRUE)
[…]

### 2019-10-27 inst/tinytest/test_pkg_party.R

@@ -1,0 +1,58 @@
+# Load Friedman benchmark data
+friedman1 <- readRDS("friedman.rds")$friedman1  # regression
+friedman2 <- readRDS("friedman.rds")$friedman2  # classification (binary)
+
+
+# Tests for package party (S4 methods)
+if (require(party, quietly = TRUE)) {
+
+  # ctree()
+
+  # Fit model(s)
+  fit1 <- ctree(y ~ ., data = friedman1)
+  fit2 <- ctree(y ~ ., data = friedman2)
+  fit3 <- cforest(y ~ ., data = friedman1)
+  fit4 <- cforest(y ~ ., data = friedman2)
+
+  # Expectations: get_training_data()
+  expect_equal(
+    current = pdp:::get_training_data.BinaryTree(fit1),
+    target = friedman1[, paste0("x.", 1L:10L)],
+    check.attributes = FALSE
+  )
+  expect_equal(
+    current = pdp:::get_training_data.BinaryTree(fit2),
+    target = friedman2[, paste0("x.", 1L:10L)],
+    check.attributes = FALSE
+  )
+  expect_equal(
+    current = pdp:::get_training_data.RandomForest(fit3),
+    target = friedman1[, paste0("x.", 1L:10L)],
+    check.attributes = FALSE
+  )
+  expect_equal(
+    current = pdp:::get_training_data.RandomForest(fit4),
+    target = friedman2[, paste0("x.", 1L:10L)],
+    check.attributes = FALSE
+  )
+  expect_error(pdp:::get_training_data.default(fit1))
+
+  # Expectations: get_task()
+  expect_identical(
+    current = pdp:::get_task.BinaryTree(fit1),
+    target = "regression"
+  )
+  expect_identical(
+    current = pdp:::get_task.BinaryTree(fit2),
+    target = "classification"
+  )
+  expect_identical(
+    current = pdp:::get_task.RandomForest(fit3),
+    target = "regression"
+  )
+  expect_identical(
+    current = pdp:::get_task.RandomForest(fit4),
+    target = "classification"
+  )
+
+}

### 2019-10-27 inst/tinytest/test_pkg_ranger.R

@@ -1,0 +1,33 @@
+# Load Friedman benchmark data
+friedman1 <- readRDS("friedman.rds")$friedman1
+friedman2 <- readRDS("friedman.rds")$friedman2
+
+# Tests for package ranger (using ::)
+if (require(ranger, quietly = TRUE)) {
+
+  # Fit model(s)
+  fit1 <- ranger(y ~ ., friedman1)  # NOTE: Data arg not named!
+  fit2 <- ranger(y ~ ., data = friedman1)
+  fit3 <- ranger(y ~ ., data = friedman2, probability = TRUE)
+  fit4 <- ranger(y ~ ., data = friedman2, probability = FALSE)
+
+  # Compute partial dependence for x.3
+  pd1 <- partial(fit1, pred.var = "x.3")
+  pd2 <- partial(fit2, pred.var = "x.3")
+  pd3 <- partial(fit3, pred.var = "x.3")
+  pd4 <- partial(fit3, pred.var = "x.3", prob = TRUE)
+  ice1 <- partial(fit2, pred.var = "x.3", ice = TRUE)
+  ice2 <- partial(fit3, pred.var = "x.3", ice = TRUE)
+  ice3 <- partial(fit3, pred.var = "x.3", ice = TRUE, prob = TRUE)
+
+  # Expectations: partial()
+  expect_true(inherits(pd1, what = "partial"))
+  expect_true(inherits(pd2, what = "partial"))
+  expect_true(inherits(pd3, what = "partial"))
+  expect_true(inherits(pd4, what = "partial"))
+  expect_true(inherits(ice1, what = "ice"))
+  expect_true(inherits(ice2, what = "ice"))
+  expect_true(inherits(ice3, what = "ice"))
+  expect_error(partial(fit4, pred.var = "x.3"))
+
+}

### 2019-10-27 inst/tinytest/test_progress.R

@@ -1,0 +1,37 @@
+# Load required packages
+library(Matrix)
+
+# Load Friedman benchmark data
+friedman1 <- readRDS("friedman.rds")$friedman1
+X1 <- data.matrix(friedman1[, paste0("x.", 1L:10L)])
+
+# Tests progress bar with XGBoost models
+if (require(xgboost, quietly = TRUE)) {
+
+  # Regression model -----------------------------------------------------------
+
+  # Fit regression model
+  fit1 <- xgboost(
+    data = X1,
+    label = friedman1$y,
+    params = list(
+      max_depth = 2,
+      eta = 0.1,
+      objective = "reg:squarederror",  # formerly "reg:linear"
+      eval_metric = "rmse"
+    ),
+    nrounds = 827,
+    verbose = 0,
+    save_period = NULL
+  )
+
+  # Compute partial dependence with progress bar
+  grid_size <- 1000L
+  pd <- partial(fit1, pred.var = "x.3", train = X1, progress = "progress",
+                grid.resolution = grid_size)
+
+  # Expectations
+  expect_true(nrow(pd) == grid_size)
+  expect_true(inherits(pd, what = "partial"))
+
+}

### 2019-10-27 inst/tinytest/test_topPredictors.R

@@ -1,0 +1,21 @@
+# Load Friedman benchmark data
+friedman1 <- readRDS("friedman.rds")$friedman1
+
+# Tests for topPredictors() using caret and randomForest
+if (require(caret, quietly = TRUE) && require(randomForest, quietly = TRUE)) {
+
+  # Regression model -----------------------------------------------------------
+
+  # Fit a random forest
+  fit <- train(y ~ ., data = friedman1, method = "rf", importance = TRUE,
+               trControl = trainControl(method = "none"))
+
+  # Extract "top" predictors
+  vi1 <- topPredictors(fit, n = 100L)
+  vi2 <- topPredictors(fit$finalModel, n = 100L)
+
+  # Expectations
+  expect_identical(vi1, vi2)
+  expect_true(length(vi1) == ncol(friedman1) - 1)
+
+}

### 2019-10-27 slowtests/slowtests-calhousing.R

@@ -34,7 +34,7 @@ cal.gbm <- gbm(AvgValue ~ ., data = cal,
                verbose = TRUE)
 best.iter <- gbm.perf(cal.gbm, method = "test")
 
-# There seems to be a fair amount of outliers in the datset so we have two
+# There seems to be a fair amount of outliers in this data set so we have two
 # options:
 #
 #   (1) use the quantiles options with specific probabilities;
@@ -55,6 +55,7 @@ pd.AveRooms <- partial(cal.gbm, pred.var = "AveRooms", quantiles = TRUE,
 # Partial dependence of AvgValue on AveOccup and HouseAge (together)
 pd.HouseAge.AveOccup <- partial(
   cal.gbm, pred.var = c("HouseAge", "AveOccup"), quantiles = TRUE,
+  # chull = TRUE,
   probs = 5:95/100, n.trees = best.iter
 )
 

### 2019-10-27 tests/testthat/test-partial.R

@@ -1,144 +1,0 @@
-context("Partial dependence functions")
-
-# Data frames used in the following tests
-pima2 <- na.omit(pima)
-set.seed(101)
-df.reg <- boston[sample(nrow(boston), size = 50, replace = FALSE), ]
-df.class <- pima2[sample(nrow(pima2), size = 50, replace = FALSE), ]
-
-# Switch to generic response names
-df.reg$y <- df.reg$cmedv
-df.class$y <- df.class$diabetes
-df.reg$cmedv <- NULL
-df.class$diabetes <- NULL
-
-# Boston housing
-boston2 <- boston
-boston2$chas <- ifelse(boston2$chas == "0", yes = "a", no = "b")
-
-# Fit regression (lm) and classification (glm) models
-df.reg.lm <- lm(y ~ ., data = df.reg)
-df.class.glm <- glm(y ~ ., data = df.class, family = binomial)
-
-test_that("partial works correctly for regression", {
-
-  # Specifying resolution of automatic grid
-  pd1 <- partial(df.reg.lm, pred.var = "rm", grid.resolution = 5,
-                 train = df.reg)
-
-  # Specifying a grid manually
-  pd2 <- partial(df.reg.lm, pred.var = "rm", pred.grid = data.frame(rm = 1),
-                 check.class = FALSE, train = df.reg)
-
-  # Return a trellis object
-  pd3 <- partial(df.reg.lm, pred.var = "rm", grid.resolution = 5, plot = TRUE,
-                 train = df.reg)
-
-  # Using specific quantiles
-  pd4 <- partial(df.reg.lm, pred.var = "rm", quantiles = TRUE, probs = 1:9/10,
-                 train = df.reg)
-
-  # Expectations
-  expect_is(pd1, "data.frame")
-  expect_is(pd2, "data.frame")
-  expect_is(pd3, "trellis")
-  expect_is(pd4, "data.frame")
-  expect_identical(pd4$rm, as.numeric(quantile(df.reg$rm, probs = 1:9/10)))
-
-  # Trellis object should still contain partial dependence data
-  expect_is(attr(pd3, "partial.data"), "data.frame")
-  expect_identical(pd1, attr(pd3, "partial.data"))
-
-  # Warnings
-  expect_warning(partial(df.reg.lm, pred.var = "rm",
-                         pred.grid = df.reg[,"rm", drop = FALSE],
-                         trim.outliers = TRUE, check.class = FALSE,
-                         train = df.reg))
[…]

### 2019-10-27 inst/tinytest/test_partial.R
fix CRAN issue w/ implicit parallelization in GBM tests

@@ -13,7 +13,8 @@ if (require(gbm, quietly = TRUE) && require(ggplot2, quietly = TRUE)) {
     interaction.depth = 6,
     shrinkage = 0.1,
     bag.fraction = 1,
-    cv.folds = 5
+    cv.folds = 5,  # implicitly uses more cores than R CMD Check will allow
+    n.cores = 2  # set to CRAN max
   )
   best_iter <- gbm.perf(fit, method = "cv")
 

### 2019-10-27 README.Rmd
ignore vignettes
update install instructions

@@
 [![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/bgreenwell/pdp?branch=master&svg=true)](https://ci.appveyor.com/project/bgreenwell/pdp)
-[![Coverage Status](https://img.shields.io/codecov/c/github/bgreenwell/pdp.svg)](https://codecov.io/github/bgreenwell/pdp?branch=master)
+[![Codecov test coverage](https://codecov.io/gh/bgreenwell/pdp/branch/master/graph/badge.svg)](https://codecov.io/gh/bgreenwell/pdp?branch=master)
 [![Downloads](http://cranlogs.r-pkg.org/badges/pdp)](http://cranlogs.r-pkg.org/badges/pdp)
@@
 To report bugs or issues, contact the main author directly or submit them to https://github.com/bgreenwell/pdp/issues.
+For additional documentation and examples, visit the [package website](https://bgreenwell.github.io/pdp/index.html).
 
@@
 # Alternatively, you can install the development version from GitHub:
-if (!requireNamespace("devtools")) {
-  install.packages("devtools")
-}
-devtools::install_github("bgreenwell/pdp")
+# install.packages("remotes")
+remotes::install_github("bgreenwell/pdp")
 ```

### 2019-10-28 README.Rmd
bug fixes

@@
 
-As of right now, `pdp` exports four functions:
+As of right now, `pdp` exports the following functions:
 
@@
 
-  * `topPredictors()` extract most "important" predictors from various types of fitted models. (Will soon be replaced by functionality from [vip](https://koalaverse.github.io/vip/index.html).)
+  * ~~`topPredictors()` extract most "important" predictors from various types of fitted models.~~ see [vip](https://koalaverse.github.io/vip/index.html) instead for a more robust and flexible replacement;
+
+  * `exemplar()` - construct an exemplar record from a data frame (**experimental** feature that may be useful for constructing fast, approximate feature effect plots.)
 
@@
 # Alternatively, you can install the development version from GitHub:
-# install.packages("remotes")
+if (!("remotes" %in% installed.packages()[, "Package"])) {
+  install.packages("remotes")
+}
 remotes::install_github("bgreenwell/pdp")

### 2019-10-28 R/autoplot.R

@@ -263,7 +263,7 @@ ggplot_ice_curves <- function(
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))
-        p <- p + geom_rug(data = x.rug, aes(x = x.rug[[1L]]),csides = "b",
+        p <- p + geom_rug(data = x.rug, aes(x = x.rug[[1L]]), sides = "b",
                           inherit.aes = FALSE)
       }
     }

### 2019-10-28 R/getParDepGBM.R

@@ -18,14 +18,39 @@ getParDepGBM <- function(object, pred.var, pred.grid, which.class, prob, ...) {
     object$num.classes <- 1
   }
 
+  ##############################################################################
+
+  # FIXME: What's the best way to do this?
+
   # Convert categorical variables to integer (i.e., 0, 1, 2, ..., K)
   for (i in seq_len(length(pred.grid))) {
-    if (!is.numeric(pred.grid[[i]])) {
-      factor.vals <- as.character(pred.grid[[i]])  # save original factor values
-      pred.grid[[i]] <- as.numeric(pred.grid[[i]]) - 1  # convert to 0, 1, ..., K
-      attr(pred.grid[[i]], which = "factor.vals") <- factor.vals  # store as attribute
+
+    # For `"gbm"` objects, possibilities are "numeric", "ordered", or "factor".
+    # But ordered factors actually inherit from class `"factor"`, so only need
+    # to check for that here.
+    if (inherits(pred.grid, "factor")) {
+
+      # Save original factor values (could possibly be "ordered")
+      levs <- levels(pred.grid[[i]])
+      vals <- as.character(pred.grid[[i]])
+      type <- class(pred.grid[[i]])
+
+      # Convert from categorical to integer (i.e., 0, 1, ..., K). For example,
+      # c("low", "hot", "med"), w/ low < med < hot, should be converted to
+      # c(0, 2, 1).
+      pred.grid[[i]] <- as.numeric(pred.grid[[i]]) - 1
+
+      # Store original categorical values, class information, etc.
+      attr(pred.grid[[i]], which = "cat") <- TRUE  # categorical indicator
+      attr(pred.grid[[i]], which = "levs") <- levs  # factor levels
+      attr(pred.grid[[i]], which = "vals") <- vals  # factor values
+      attr(pred.grid[[i]], which = "original_class") <- type  # factor type
+
     }
+
   }
+
+  ##############################################################################
 
   # Partial dependence values
   y <- .Call("PartialGBM",
@@ -69,10 +94,24 @@ getParDepGBM <- function(object, pred.var, pred.grid, which.class, prob, ...) {
     pd.df$yhat <- y
   }
 
+  ##############################################################################
+
+  # FIXME: What's the best way to do this?
+
   # Transform categorical variables back to factors
   for (i in seq_len(length(pred.var))) {
[…]

### 2019-10-28 R/partial.R

@@ -399,7 +399,7 @@ partial.default <- function(
   }
 
   # Compute "poor man's partial dependence"
-  if (isTRUE(approx)) {
+  if (isTRUE(approx)) {       # FIXME: What about when `rug/chull = TRUE`
     train <- exemplar(train)  # FIXME: Better handling for matrix-like objects
   }
 

### 2019-10-28 R/plotPartial.R

@@ -257,7 +257,7 @@ plot_ice_curves <- function(
           stop("The training data must be supplied for rug display.")
         } else {
           panel.rug(
-            stats::quantile(train[, names(object)[1L]], probs = 0:10/10,
+            stats::quantile(train[[names(object)[1L]]], probs = 0:10/10,
                             na.rm = TRUE)
           )
         }
@@ -286,7 +286,7 @@ plot_one_predictor_pdp <- function(object, smooth, rug, train = NULL, ...) {
             stop("The training data must be supplied for rug display.")
           } else {
             panel.rug(
-              stats::quantile(train[, names(object)[1L]],
+              stats::quantile(train[[names(object)[1L]]],
                               probs = 0:10/10, na.rm = TRUE)
             )
           }
@@ -343,7 +343,7 @@ plot_two_predictor_pdp <- function(
             stop("The training data must be supplied for rug display.")
            } else {
              panel.rug(
-               stats::quantile(train[, names(object)[1L]], probs = 0:10/10,
+               stats::quantile(train[[names(object)[1L]]], probs = 0:10/10,
                                na.rm = TRUE)
              )
            }
@@ -383,10 +383,10 @@ plot_two_predictor_pdp <- function(
             if (rug) {
               panel.rug(
                 x = stats::quantile(
-                  train[, names(object)[1L]], probs = 0:10/10, na.rm = TRUE
+                  train[[names(object)[1L]]], probs = 0:10/10, na.rm = TRUE
                 ),
                 y = stats::quantile(
-                  train[, names(object)[2L]], probs = 0:10/10, na.rm = TRUE
+                  train[[names(object)[2L]]], probs = 0:10/10, na.rm = TRUE
                 ),
                 col = "black"
               )
@@ -471,7 +471,7 @@ plot_three_predictor_pdp <- function(
             stop("The training data must be supplied for rug display.")
           } else {
             panel.rug(
-              stats::quantile(train[, names(object)[1L]], probs = 0:10/10,
+              stats::quantile(train[[names(object)[1L]]], probs = 0:10/10,
                               na.rm = TRUE)
             )
           }

### 2019-10-28 inst/tinytest/test_partial.R

@@ -1,7 +1,7 @@
 if (require(gbm, quietly = TRUE) && require(ggplot2, quietly = TRUE)) {
 
   # Load the Ames housing data
-  ames <- AmesHousing::make_ames()
+  ames <- as.data.frame(AmesHousing::make_ames())
 
   # Fit a gbm model
   set.seed(1918)  # for reproducibility
@@ -37,6 +37,16 @@ if (require(gbm, quietly = TRUE) && require(ggplot2, quietly = TRUE)) {
     plotPartial(pd2),
     nrow = 1
   )
+
+  # Users can call plotPartial() to construct lattice-based PDPs
+  pdp1_lattice <- plotPartial(pd1, smooth = TRUE, rug = TRUE, train = ames)
+
+  # Users can call autoplot() to construct ggplot2-based PDPs
+  pdp1_ggplot2 <- autoplot(pd1, smooth = TRUE, rug = TRUE, train = ames)
+
+  # Expectation(s)
+  expect_true(inherits(pdp1_lattice, what = "trellis"))
+  expect_true(inherits(pdp1_ggplot2, what = "ggplot"))
 
   # Users should be able to supply the predictor name(s) or column position(s)
   # in train
@@ -127,6 +137,19 @@ if (require(gbm, quietly = TRUE) && require(ggplot2, quietly = TRUE)) {
     partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1, pred.fun = pfun2)
   )
 
+  # ICE curves constructed manually
+  pdp_pfun3_ggplot2 <- autoplot(pd_pfun3, center = TRUE, rug = TRUE,
+                                train = ames, alpha = 0.1)
+  pdp_pfun3_lattice <- plotPartial(pd_pfun3, center = TRUE, rug = TRUE,
+                                   train = ames, alpha = 0.1)
+
+  # Expectation(s)
+  expect_true(inherits(pdp_pfun3_ggplot2, what = "ggplot"))
+  expect_true(inherits(pdp_pfun3_lattice, what = "trellis"))
+
+  # Display plots side by side
+  grid.arrange(pdp_pfun3_ggplot2, pdp_pfun3_lattice, nrow = 1)
+
   # c-ICE curves
   ice <- partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1, ice = TRUE,
                  center = TRUE, n.trees = best_iter)
[…]

### 2019-10-28 inst/tinytest/test_pkg_party.R

@@ -1,58 +1,81 @@
-# Load Friedman benchmark data
-friedman1 <- readRDS("friedman.rds")$friedman1  # regression
-friedman2 <- readRDS("friedman.rds")$friedman2  # classification (binary)
-
-
 # Tests for package party (S4 methods)
 if (require(party, quietly = TRUE)) {
 
-  # ctree()
+  # Load Friedman benchmark data
+  friedman1 <- readRDS("friedman.rds")$friedman1  # regression
+  friedman2 <- readRDS("friedman.rds")$friedman2  # classification (binary)
+
+  # Conditional inference tree; party::ctree() ---------------------------------
 
   # Fit model(s)
   fit1 <- ctree(y ~ ., data = friedman1)
   fit2 <- ctree(y ~ ., data = friedman2)
+
+  # Partial dependence for x.4
+  pd1 <- partial(fit1, pred.var = "x.4")
+  pd2 <- partial(fit2, pred.var = "x.4")  # default (class-centered) logit scale
+  pd2_prob <- partial(fit2, pred.var = "x.4", prob = TRUE)  # probability scale
+
+  # ICE curves for x.4
+  ice1 <- partial(fit1, pred.var = "x.4", ice = TRUE, center = TRUE)
+  ice2 <- partial(fit2, pred.var = "x.4",
+                  ice = TRUE, center = TRUE)  # default (class-centered) logit scale
+  ice2_prob <- partial(fit2, pred.var = "x.4", prob = TRUE,
+                       ice = TRUE, center = TRUE)  # probability scale
+
+  # Expectation(s)
+  expect_true(inherits(pd1, what = "partial"))
+  expect_true(inherits(pd2, what = "partial"))
+  expect_true(inherits(pd2_prob, what = "partial"))
+  expect_true(inherits(ice1, what = "cice"))
+  expect_true(inherits(ice2, what = "cice"))
+  expect_true(inherits(ice2_prob, what = "cice"))
+
+  # Display plots in a grid
+  grid.arrange(
+    plotPartial(pd1),
+    plotPartial(pd2),
+    plotPartial(pd2_prob),
+    plotPartial(ice1),
+    plotPartial(ice2),
+    plotPartial(ice2_prob),
+    nrow = 2
+  )
+
+  # Conditional inference forest; party::cforest() -----------------------------
+
   fit3 <- cforest(y ~ ., data = friedman1)
   fit4 <- cforest(y ~ ., data = friedman2)
 
[…]

### 2020-02-23 rjournal/reviews/response.Rmd
refactor code, etc.

---
title: "Response to Comments"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
  word_document: default
---

Thank you for your time and thoughtful comments. I believe that they have helped to greatly improved, not only the manuscript, but the package it describes as well. Please find below a copy of all the comments along with my responses.

# Reviewer 1

### Comment 1.1
**Bug**: A graph is produced as expected if the code below is cut-and-pasted at the command line .  However, if the code is put into a file (say `foo.R`) and `source("foo.R")` is executed at the command line, no graph is produced.  (But note that if `plot=TRUE` is added to the call to `partial()`, then a plot is produced as expected.)
```{r, eval=FALSE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
library(pdp)
pd <- partial(ozone.rf, pred.var="Temp")
plotPartial(pd)
```

### Response 1.1
This is simply a side effect of using `lattice` to produce the plots. If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))` or `print(partial(fit, pred.var = "x", plot = TRUE))`). This has been clarified in the package documentation and the manuscript (see **Note:** at the bottom of page 3). The fact that this worked when calling `partial` with `plot = TRUE` was a bug (which is now fixed).

### Comment 1.2
I couldn't get `partial` to work with `gbm` (version 2.1.1) models (although I didn't try very hard).  Perhaps this is because the `gbm` maintainer changed the `gbm` interface in a non-backcompatible way. The following (where `gbm.mod` is a model built using the gbm package)
```{r, eval=FALSE}
partial(gbm.mod, pred.var="age")
```
gives
```{r, eval=FALSE}
Error in paste("Using", n.trees, "trees...\n") :
argument "n.trees" is missing, with no default
```
I also tried
```{r, eval=FALSE}
partial(gbm.mod, pred.var="age", n.trees=100)
```
but that gives
```{r, eval=FALSE}
Error in .fun(piece, ...) : unused argument (n.trees = 100)
```

### Response 1.2
This has been fixed in `pdp` (version 0.3.0). The user may still have to supply `n.trees` in the call to `partial` (depending on which version of `gbm` they are using). Also, as suggested in another comment, a `/slowtests` directory is underway and will help prevent future issues like this. Thanks for cathing this.

### Comment 1.3
Standard partial matching isn't implemented (e.g., `partial(..., type="r")`).

### Response 1.3
Partial matching is now implemented (as of `pdp` version 0.3.0), but only for the `type` argument. Thanks for the suggestion.

### Comment 1.4
The following code incorrectly gives an error message
```
"Error in names(pd_df) <- c(pred.var, "y") :
  'names' attribute [2] must be the same length as the vector [1]"
```
It seems to work ok if you change the variable names.
```{r, eval=FALSE}
library(pdp)
data <- data.frame(V1 = 1:10, V2 = 1:10)
mod <- lm(V2 ~ V1, data = data)
partial(mod, pred.var = "V1")  # error
```

### Response 1.4
This has been fixed in `pdp` (version 0.3.0). This was simply a side effect of `partial`'s internal use of `plyr::adply` which uses default variable names (e.g., `V1`, `V2`, etc.). This caused the column `V1` in the output data frame to be overwritten, producing the error. 
```{r, eval=TRUE}
library(pdp)
data <- data.frame(V1 = 1:10, V2 = 1:10)
mod <- lm(V2 ~ V1, data = data)
partial(mod, pred.var = "V1")  # no more error
```

```{r, include=FALSE}
rm(data)
```

### Comment 1.5
I think you should mention in the documentation something about the assumptions you make about accessing the model data. The following (somewhat contrived) example fails with
`"Error in train[[x]] : subscript out of bounds":`
```{r, eval=FALSE}
data(trees)
foo <- function(data) {
  lm(Volume ~ ., data = data)
}
mod <- foo(trees)
library(pdp)
partial(mod, pred.var = "Girth", plot = TRUE)  # fails
```
and the following silently gives misleading results:

```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)  # ok
trees <- trees[1:2, ]
partial(mod, pred.var = "Girth", plot = TRUE)  # wrong xlim
```
I'm not necessarily saying that you need to resolve the above two
issues, but at least mention in the documentation that this kind of
thing can happen, and how to avoid it (basically call partial in the
same environment used to build the model, and don't change any of the
data used to build the model).

### Response 1.5
Agreed! Both the manuscript (see page 3) and the help page for `partial` have been expanded to discuss this. Unfortunately, some functions don't store a copy of the training data. However, I have expanded the code to work harder---especially for `S4` methods---to inform the user to supply the training data when necessary. Re-running your first example should give an informative error message:
```{r, eval=TRUE, error=TRUE}
data(trees)
foo <- function(data)
{
  lm(Volume ~ ., data = data)
}
mod <- foo(trees)
library(pdp)
partial(mod, pred.var = "Girth", plot = TRUE)  # fails with better error mssg
```

### Comment 1.6
The following code gives an obscure error message:
```
"Error in seq.default(from = min(train[[x]], na.rm = TRUE),
   to = max(train[[x]],  :
  'from' cannot be NA, NaN or infinite"
```
Some better handholding for the user may be helpful here.
```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var="nosuchvariable", plot=TRUE) # obscure err msg
```

### Response 1.6
This has been fixed in `pdp` (version 0.3.0) by adding an informative error message listing which variables are the cause of the issue.
```{r, error=TRUE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var=c("Girth", "foo", "bar"), plot=TRUE)
```

### Comment 1.7
On page 2 of the paper it says "The columns of the data frame are labeled in the same order as the features supplied to `pred.var`, and the last column is always labeled `y`". What happens if `y` is one of the predictors?

### Response 1.7
The last column is now labeled `yhat`, which is less likely to be a predictor name. When one of the predictors is labeled `"y"` `randomForest::partialPlot` fails while `gbm::plot.gbm` does not. I think regardless of what the default column name is, there is always the possibility of this problem. This has been filed as an issue on the `pdp` GitHib page: https://github.com/bgreenwell/pdp/issues/22.

### Comment 1.8
Hastie et al. in the book cited in the paper say "Although such a collection [of partial dependence plots] can seldom provide a comprehensive depiction of the approximation, it can often produce helpful clues".  Therefore in in the abstract for the paper, to avoid overselling consider changing "relationship between the outcome and predictors of interest can be easily understood" to "relationship between the outcome and predictors of interest can be more easily understood".

### Response 1.8
Fixed. Thanks for the suggestion.

### Comment 1.9
On page 3 in the PDF of the paper, add a newline to the error message in the paper:
```
Error: The training data could not be extracted from object. Please
supply the raw training data using the ...
```

### Response 1.9
Fixed. Thanks for pointing that out.

### Comment 1.10
Is there any reason the package is called `pdp` but the main function is called `partial`?  Consider giving the function the same name as the package, so e.g., the following works (to help the user get started)
```{r, eval=FALSE}
library(pdp)
?pdp  # fails
```

### Response 1.10
The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letters). So, I decided to change the name of the package to `pdp`; besides, I think `partial` and `plotPartial` work well together as function names. However, I did add a help page describing the package, main functions, etc. So, `?pdp` no longer fails. However, I am not opposed to creating an alias to `partial` called `pdp` in the future.

### Comment 1.11
`partial` should return the same value regardless of whether or not the `plot` argument is used?

### Response 1.11
I dont think so, because I want `partial` to behave like a `lattice` function when using `plot = TRUE`; that is, I want it to return a `"trellis"` object. However, I did add the partial dependence data as an attribute to the returned `"trellis"` object so that it can easily be extracted if necessary (e.g., if they do not like the plot and don't want to re-run partial). For example,
```{r}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
pd.Girth <- partial(mod, pred.var = "Girth")  # data only
pdp.Girth <- partial(mod, pred.var = "Girth", plot = TRUE)  # plot
identical(pd.Girth, attr(pdp.Girth, "partial.data"))  # should be identical
```
Also, the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded. This would be useful if the user did not store the PDP but needs to retrieve the outout data to avoid re-running `partial` (e.g., when computation time is a concern). For example,
```{r, fig.width=4, fig.height=4}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```
This has been noted in the package documentation (see `?partial`). Thanks for the idea!

### Comment 1.12
A call to say `par(mfrow = c(2, 2))` before calling `plotPartial` is ignored by `plotPartial`.  This is a pity because it would allow multiple plots to be put on one page like this:
```{r, eval=FALSE}
par(mfrow = c(1, 2))
partial(ozone.rf, pred.var = "Temp", plot = TRUE)
partial(ozone.rf, pred.var = "Wind", plot = TRUE)
```

### Response 1.12
The command `par(mfrow = c(2, 2))` does not work with `"trellis"` objects, such as those produced by `lattice`---which `partial` relies on for producing its plots. Other methods are available. For convenience, `pdp` now imports the `grid.arrange` function from the `gridxtra` package (this has been discussed in the manuscript and incorporated into the examples where appropriate).
```{r, include=FALSE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data = airquality, na.action = na.omit)
library(pdp)
```
```{r, eval=TRUE, fig.width=8, fig.height=4}
grid.arrange(
  partial(ozone.rf, pred.var = "Temp", plot = TRUE),
  partial(ozone.rf, pred.var = "Wind", plot = TRUE),
  ncol = 2
)
```

However, since `partial` returns a data frame by default, using `par` would work in the following case
```{r, eval=TRUE, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))
plot(partial(ozone.rf, pred.var = "Temp"), type = "l")
plot(partial(ozone.rf, pred.var = "Wind"), type = "l")
```

### Comment 1.13
It would be nice if the functions automatically plotted all the (important) variables in a grid of plots, without forcing the user to explicitly specify them.

### Response 1.13
I agree, this would be a nice feature, but the problems I see are (1) not all models naturally emit a "variable importance" score; the `caret` package has come a long way in this regard, and (2) too many plots will look cluttered. At this point, I'd rather encourage the user to select which variables they would like to see plotted. I am not opposed to including this feature in the future. I added this as a feature request: https://github.com/bgreenwell/pdp/issues/19. Thanks for the suggestion!

### Comment 1.14
It would be nice if the functions automatically got the data from the model used to build the model in more cases without forcing the user to explicitly specify the data.  Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only---one can't access the data stored in `xgb.DMatrix`).

### Response 1.14
`kernlab::ksvm` does not store a copy of the training data either, as far as I can tell. As time goes on, the coverage will certainly improve, which will make it easier on the user. For example, `partial` now works with `caret` which, by default, stores the taining data in a component called `trainingData`. Also, I suspect that the package maintainers for `xgboost` (and similar packages designed for working with "large" data sets) would be hesitent in storing a copy of the training data with the fitted model for memory reasons.

### Comment 1.15
Consider adding a `slowtests` directory (below `pdp/inst/` or elsewhere) that has tests for all the models supported by the package (Table 1 in the paper) to easily check back compatibility when packages change.  (For justification for such tests see for example my comments on `gbm` back compatibility above.)  These tests will probably be quite slow, but that won't be an issue because they won't be invoked by CRAN check; the slow tests would be just for the maintainer of the `pdp` package to run manually.

### Response 1.15
This is a good suggestion and is currently under development: https://github.com/bgreenwell/pdp/blob/master/slowtests/slowtests.R. In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.

# Reviewer 2

### Comment 2.1
The Introduction immediately introduces the PDP. You must first motivate why it is important. In a linear model, they are not important. In a linear model with simple interactions, likewise. They first become important if you have non-parametric machine-learning techniques. Further, they are not the only idea extant when trying to explain a predictor's impact on the response. This paper would benefit if you can give a bit more of a lit review here.

### Response 2.1
I expanded the introduction a bit to try and motivate why PDPs are important (the new material links directly to the Boston housing example used throughout the manuscript). I also included more background on the `plotmo` and `ICEbox` packages (and there corresponding methodologies). `plotmo` very recently introduced a `partdep` option for displaying PDPs. Because of this, more detail was added to differentiate `pdp` and motivate the use of `partial`.

I agree that PDPs are not as useful in "simple" linear models (e.g., few variables, only two-way interactions, etc.), but certainly there will be cases where PDPs will be extremely useful (e.g., many terms, complex interactions, etc.). In fact, linear models are still common in machine learning/computer science/data science (they just typically contain many terms). I also agree that PDPs are not the only, or sometimes best, approach to understanding the impact of a subset of the predictor's on the response. Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few other techniques available in the `car` package.

## Important Clarifications Required:

### Comment 2.2
Top of p2. The `ICEbox` package allows for general construction and visualization of PDP's. However, it is limited as you only get one dimensional representations. The selling point of your package should focus on the flexibility of graphics and the illustration of non-extrapoloation zones.

### Response 2.2
`ICEbox` does not actually construct PDPs, rather individual curves for each observation (which can be difficults to interpret when you have millions of observations). I think flexible graphics and non-extrapoloation zones are only half of the selling point here. I think that being able to run `partial` in parallel is huge, and also that this packge works with many types of machine learning models---not just tree-based models. For example, as described in the manuscript, `pdp` can produce single or multi-predictor PDPs for conditional random forests and XGBoost models (among many others) in parallel. I expanded the last example to describe the interface with `caret`, which I think opens the door for wonderful exploratory data analysis pipe lines (e.g., choose model --> plot importance --> plot relationships, etc.). Also, a new section was added to the manuscript: **User-defined prediction functions**. This section describes the new `pred.fun` argument (as suggested by a later comment) that makes `partial` even more flexible.

### Comment 2.3
All `pdp` figures throughout the paper: Can you make explicit in the caption what code produces the left and right subfigures? I would use the subfigure package in latex and then put a comment in the R code e.g. # Fig 3b.

### Response 2.3
Rather than using subfigures, which I think would look messy with all the code, I made quite an attempt to comment which pieces of code produced which figure. Also, in response to another reviewer's comments, the code which produces figures containing multiple plots is better described and demonstrated. See, for example, the **Multi-predictor PDPs** section.

### Edits:

### Comment 2.4
p1 abstract: strict assumptions imposed by traditional statistical models

### Response 2.4
Added some clarification here; in particular, normality, constant variance, and linearity.

### Comment 2.5
p1 pairs should be using curly braces of angle braces, not square braces as this is easily confused with closed intervals.

### Response 2.5
Fixed. Switched to curly braces.

### Comment 2.6
p2 unless the editor disagrees, there is no reason to add "install.packages("pdp")" in the text of the paper as it is assumed a user knows how to install a package.

### Response 2.6
Fixed. Code chunk was removed.

### Comment 2.7
p3 "Error: The training..." should be wrapped.

### Response 2.7
Fixed. The text is now wrapped appropriately. Thanks for catching this!

### Comment 2.8
Fig 6: the sidebar should be labeled "ozone". Also, would it be terrible to label all axes even though it is redundant?

### Response 2.8
The sidebar (i.e., colorkey) has been labeled, but for this example only. As sad as it sounds, labeling the colorkey is not as trivial as we'd expect for `lattice` plots. The code for doing so has been incorporated in the example in question (see pages 8-9). Also, the text describing the first code chunck that produces a display with a colorkey now contains a footnote pointing to the example where the code to produce a label can be found. This is also motivation for why one of the examples showed how to use `ggplot2` for plotting the results from `partial`---`ggplot2` makes it a trivial matter to label the colorkey. I don't think it would be terrible to label all the axes. However, in my option, labelling all the axes would (1) looks cluttered, and (2) require more unnecessary code that would probably distract the reader. This is the standard axis labelling scheme used in trellis plots from S-plus to R (via `lattice`).

### Comment 2.9
Fig 7: as beautiful as this is, you should label the sidebar as "logit" or "centered logit" instead of "y".

### Response 2.9
Fixed. Color key labeled "Centered logit".

### Suggestions for the software

### Comment 2.10
I understand that you like the feature of having a one-shot function partial to both generate and plot the PDP. I realize that the default is `FALSE`. However, I still think it's a disservice to the user as constructing PDP's is computationally intense and if they get the plot wrong or it's not to their liking, they will have to regenerate the whole thing again. So, it is my opinion that you should take it out and force two steps: generation and illustration.

### Response 2.10
While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them first. However, I agree with your concern which is why `partial` has `plot = FALSE` as the default; at the bottom of the second paragraph of section **Constructing PDPs in R** (and in the package documentation), I have added a note empahsizing the use of `plotPartial`. Furthermore, I have made it possible to retrieve the raw PDP data even when `plot = TRUE`. This is because the `"trellis"` object returned by `partial` / `plotPartial` now contains an additional `partial.data` attribute. This is described in the package documentation. For example,
```{r, eval=TRUE, fig.width=5, fig.height=3}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```

### Comment 2.11
The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use random forests to illustrate classification probabilities.

### Response 2.11
This is an excellent suggestion. First, I have changed the behavior of `...` to refer to additional arguments in `predict` (e.g., the `n.trees` argument in `predict.gbm`.). Second, `partial` gained a new argument, `pred.fun`, where the user can provide an optional prediction function. This makes `partial` far more flexible. For example, you can use this argument to construct PDPs on the probability scale, or to construct ICE curves. This is illusrated in the new section labeled **User-defined prediction functions**

### Comment 2.12
p7 I believe there are packages allowing for parallelization in Windows. Can you switch the example so this works on all platforms?

### Response 2.12
Good suggestion. I tweaked the example to run on both Windows and Unix-like systems. This is also reflected in the manuscript text.

### 2020-02-23 rjournal/reviews/response_reviewer_1.Rmd
refactor code, etc.

--- 2020-02-23 rjournal/reviews/response.Rmd
@@
 ---
-title: "Response to Comments"
+title: "Response to Comments from Reviewer 1"
 output:
@@
 I believe that they have helped to greatly improved, not only the manuscript, but the package it describes as well.
-Please find below a copy of all the comments along with my responses.
-
-# Reviewer 1
+Please find below a copy of your comments along with my responses.
 
@@
 In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.
-
-# Reviewer 2
-
-### Comment 2.1
-The Introduction immediately introduces the PDP.
-You must first motivate why it is important.
-In a linear model, they are not important.
-In a linear model with simple interactions, likewise.
-They first become important if you have non-parametric machine-learning techniques.
-Further, they are not the only idea extant when trying to explain a predictor's impact on the response.
-This paper would benefit if you can give a bit more of a lit review here.
-
-### Response 2.1
-I expanded the introduction a bit to try and motivate why PDPs are important (the new material links directly to the Boston housing example used throughout the manuscript).
-I also included more background on the `plotmo` and `ICEbox` packages (and there corresponding methodologies).
-`plotmo` very recently introduced a `partdep` option for displaying PDPs.
-Because of this, more detail was added to differentiate `pdp` and motivate the use of `partial`.
-
-I agree that PDPs are not as useful in "simple" linear models (e.g., few variables, only two-way interactions, etc.), but certainly there will be cases where PDPs will be extremely useful (e.g., many terms, complex interactions, etc.).
-In fact, linear models are still common in machine learning/computer science/data science (they just typically contain many terms).
-I also agree that PDPs are not the only, or sometimes best, approach to understanding the impact of a subset of the predictor's on the response.
-Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few other techniques available in the `car` package.
-
-## Important Clarifications Required:
-
-### Comment 2.2
-Top of p2.
-The `ICEbox` package allows for general construction and visualization of PDP's.
-However, it is limited as you only get one dimensional representations.
-The selling point of your package should focus on the flexibility of graphics and the illustration of non-extrapoloation zones.
-
-### Response 2.2
-`ICEbox` does not actually construct PDPs, rather individual curves for each observation (which can be difficults to interpret when you have millions of observations).
-I think flexible graphics and non-extrapoloation zones are only half of the selling point here.
-I think that being able to run `partial` in parallel is huge, and also that this packge works with many types of machine learning models---not just tree-based models.
-For example, as described in the manuscript, `pdp` can produce single or multi-predictor PDPs for conditional random forests and XGBoost models (among many others) in parallel.
-I expanded the last example to describe the interface with `caret`, which I think opens the door for wonderful exploratory data analysis pipe lines (e.g., choose model --> plot importance --> plot relationships, etc.).
-Also, a new section was added to the manuscript: **User-defined prediction functions**.
-This section describes the new `pred.fun` argument (as suggested by a later comment) that makes `partial` even more flexible.
-
-### Comment 2.3
-All `pdp` figures throughout the paper: Can you make explicit in the caption what code produces the left and right subfigures?
-I would use the subfigure package in latex and then put a comment in the R code e.g.
-# Fig 3b.
-
-### Response 2.3
-Rather than using subfigures, which I think would look messy with all the code, I made quite an attempt to comment which pieces of code produced which figure.
-Also, in response to another reviewer's comments, the code which produces figures containing multiple plots is better described and demonstrated.
-See, for example, the **Multi-predictor PDPs** section.
-
-### Edits:
-
-### Comment 2.4
-p1 abstract: strict assumptions imposed by traditional statistical models
-
-### Response 2.4
-Added some clarification here; in particular, normality, constant variance, and linearity.
-
-### Comment 2.5
-p1 pairs should be using curly braces of angle braces, not square braces as this is easily confused with closed intervals.
-
-### Response 2.5
-Fixed. Switched to curly braces.
-
-### Comment 2.6
-p2 unless the editor disagrees, there is no reason to add "install.packages("pdp")" in the text of the paper as it is assumed a user knows how to install a package.
-
-### Response 2.6
-Fixed. Code chunk was removed.
-
-### Comment 2.7
-p3 "Error: The training..." should be wrapped.
-
-### Response 2.7
-Fixed. The text is now wrapped appropriately. Thanks for catching this!
-
-### Comment 2.8
-Fig 6: the sidebar should be labeled "ozone". Also, would it be terrible to label all axes even though it is redundant?
-
-### Response 2.8
-The sidebar (i.e., colorkey) has been labeled, but for this example only.
-As sad as it sounds, labeling the colorkey is not as trivial as we'd expect for `lattice` plots.
-The code for doing so has been incorporated in the example in question (see pages 8-9).
-Also, the text describing the first code chunck that produces a display with a colorkey now contains a footnote pointing to the example where the code to produce a label can be found.
-This is also motivation for why one of the examples showed how to use `ggplot2` for plotting the results from `partial`---`ggplot2` makes it a trivial matter to label the colorkey.
-I don't think it would be terrible to label all the axes.
-However, in my option, labelling all the axes would (1) looks cluttered, and (2) require more unnecessary code that would probably distract the reader.
-This is the standard axis labelling scheme used in trellis plots from S-plus to R (via `lattice`).
-
-### Comment 2.9
-Fig 7: as beautiful as this is, you should label the sidebar as "logit" or "centered logit" instead of "y".
-
-### Response 2.9
-Fixed. Color key labeled "Centered logit".
-
-### Suggestions for the software
-
-### Comment 2.10
-I understand that you like the feature of having a one-shot function partial to both generate and plot the PDP.
-I realize that the default is `FALSE`.
-However, I still think it's a disservice to the user as constructing PDP's is computationally intense and if they get the plot wrong or it's not to their liking, they will have to regenerate the whole thing again.
-So, it is my opinion that you should take it out and force two steps: generation and illustration.
-
-### Response 2.10
-While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them first.
-However, I agree with your concern which is why `partial` has `plot = FALSE` as the default; at the bottom of the second paragraph of section **Constructing PDPs in R** (and in the package documentation), I have added a note empahsizing the use of `plotPartial`.
-Furthermore, I have made it possible to retrieve the raw PDP data even when `plot = TRUE`.
-This is because the `"trellis"` object returned by `partial` / `plotPartial` now contains an additional `partial.data` attribute.
-This is described in the package documentation.
-For example,
-```{r, eval=TRUE, fig.width=5, fig.height=3}
-library(pdp)
-data(trees)
-mod <- lm(Volume ~ ., data = trees)
-partial(mod, pred.var = "Girth", plot = TRUE)
-pdp.Girth <- lattice::trellis.last.object()
-pd.Girth <- attr(pdp.Girth, "partial.data")
-head(pd.Girth)
-```
-
-### Comment 2.11
-The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use random forests to illustrate classification probabilities.
-
-### Response 2.11
-This is an excellent suggestion.
-First, I have changed the behavior of `...` to refer to additional arguments in `predict` (e.g., the `n.trees` argument in `predict.gbm`.).
-Second, `partial` gained a new argument, `pred.fun`, where the user can provide an optional prediction function.
-This makes `partial` far more flexible.
-For example, you can use this argument to construct PDPs on the probability scale, or to construct ICE curves.
-This is illusrated in the new section labeled **User-defined prediction functions**
-
-### Comment 2.12
-p7 I believe there are packages allowing for parallelization in Windows. Can you switch the example so this works on all platforms?
-
-### Response 2.12
-Good suggestion. I tweaked the example to run on both Windows and Unix-like systems. This is also reflected in the manuscript text.

### 2020-02-23 rjournal/reviews/response_reviewer_2.Rmd
refactor code, etc.

--- 2020-02-23 rjournal/reviews/response.Rmd
@@
 ---
-title: "Response to Comments"
+title: "Response to Comments from Reviewer 2"
 output:
@@
 I believe that they have helped to greatly improved, not only the manuscript, but the package it describes as well.
-Please find below a copy of all the comments along with my responses.
-
-# Reviewer 1
-
-### Comment 1.1
-**Bug**: A graph is produced as expected if the code below is cut-and-pasted at the command line .
-However, if the code is put into a file (say `foo.R`) and `source("foo.R")` is executed at the command line, no graph is produced.
-(But note that if `plot=TRUE` is added to the call to `partial()`, then a plot is produced as expected.)
-```{r, eval=FALSE}
-library(randomForest)
-data(airquality)
-ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
-library(pdp)
-pd <- partial(ozone.rf, pred.var="Temp")
-plotPartial(pd)
-```
-
-### Response 1.1
-This is simply a side effect of using `lattice` to produce the plots.
-If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))` or `print(partial(fit, pred.var = "x", plot = TRUE))`).
-This has been clarified in the package documentation and the manuscript (see **Note:** at the bottom of page 3).
-The fact that this worked when calling `partial` with `plot = TRUE` was a bug (which is now fixed).
-
-### Comment 1.2
-I couldn't get `partial` to work with `gbm` (version 2.1.1) models (although I didn't try very hard).
-Perhaps this is because the `gbm` maintainer changed the `gbm` interface in a non-backcompatible way.
-The following (where `gbm.mod` is a model built using the gbm package)
-```{r, eval=FALSE}
-partial(gbm.mod, pred.var="age")
-```
-gives
-```{r, eval=FALSE}
-Error in paste("Using", n.trees, "trees...\n") :
-argument "n.trees" is missing, with no default
-```
-I also tried
-```{r, eval=FALSE}
-partial(gbm.mod, pred.var="age", n.trees=100)
-```
-but that gives
-```{r, eval=FALSE}
-Error in .fun(piece, ...) : unused argument (n.trees = 100)
-```
-
-### Response 1.2
-This has been fixed in `pdp` (version 0.3.0).
-The user may still have to supply `n.trees` in the call to `partial` (depending on which version of `gbm` they are using).
-Also, as suggested in another comment, a `/slowtests` directory is underway and will help prevent future issues like this.
-Thanks for cathing this.
-
-### Comment 1.3
-Standard partial matching isn't implemented (e.g., `partial(..., type="r")`).
-
-### Response 1.3
-Partial matching is now implemented (as of `pdp` version 0.3.0), but only for the `type` argument. Thanks for the suggestion.
-
-### Comment 1.4
-The following code incorrectly gives an error message
-```
-"Error in names(pd_df) <- c(pred.var, "y") :
-  'names' attribute [2] must be the same length as the vector [1]"
-```
-It seems to work ok if you change the variable names.
-```{r, eval=FALSE}
-library(pdp)
-data <- data.frame(V1 = 1:10, V2 = 1:10)
-mod <- lm(V2 ~ V1, data = data)
-partial(mod, pred.var = "V1")  # error
-```
-
-### Response 1.4
-This has been fixed in `pdp` (version 0.3.0).
-This was simply a side effect of `partial`'s internal use of `plyr::adply` which uses default variable names (e.g., `V1`, `V2`, etc.).
-This caused the column `V1` in the output data frame to be overwritten, producing the error.
-```{r, eval=TRUE}
-library(pdp)
-data <- data.frame(V1 = 1:10, V2 = 1:10)
-mod <- lm(V2 ~ V1, data = data)
-partial(mod, pred.var = "V1")  # no more error
-```
-
-```{r, include=FALSE}
-rm(data)
-```
-
-### Comment 1.5
-I think you should mention in the documentation something about the assumptions you make about accessing the model data. The following (somewhat contrived) example fails with
-`"Error in train[[x]] : subscript out of bounds":`
-```{r, eval=FALSE}
-data(trees)
-foo <- function(data) {
-  lm(Volume ~ ., data = data)
-}
-mod <- foo(trees)
-library(pdp)
-partial(mod, pred.var = "Girth", plot = TRUE)  # fails
-```
-and the following silently gives misleading results:
-
-```{r, eval=FALSE}
-library(pdp)
-data(trees)
-mod <- lm(Volume ~ ., data = trees)
-partial(mod, pred.var = "Girth", plot = TRUE)  # ok
-trees <- trees[1:2, ]
-partial(mod, pred.var = "Girth", plot = TRUE)  # wrong xlim
-```
-I'm not necessarily saying that you need to resolve the above two
-issues, but at least mention in the documentation that this kind of
-thing can happen, and how to avoid it (basically call partial in the
-same environment used to build the model, and don't change any of the
-data used to build the model).
-
-### Response 1.5
-Agreed!
-Both the manuscript (see page 3) and the help page for `partial` have been expanded to discuss this.
-Unfortunately, some functions don't store a copy of the training data.
-However, I have expanded the code to work harder---especially for `S4` methods---to inform the user to supply the training data when necessary.
-Re-running your first example should give an informative error message:
-```{r, eval=TRUE, error=TRUE}
-data(trees)
-foo <- function(data)
-{
-  lm(Volume ~ ., data = data)
-}
-mod <- foo(trees)
-library(pdp)
-partial(mod, pred.var = "Girth", plot = TRUE)  # fails with better error mssg
-```
-
-### Comment 1.6
-The following code gives an obscure error message:
-```
-"Error in seq.default(from = min(train[[x]], na.rm = TRUE),
-   to = max(train[[x]],  :
-  'from' cannot be NA, NaN or infinite"
-```
-Some better handholding for the user may be helpful here.
-```{r, eval=FALSE}
-library(pdp)
-data(trees)
-mod <- lm(Volume~., data=trees)
-partial(mod, pred.var="nosuchvariable", plot=TRUE) # obscure err msg
-```
-
-### Response 1.6
-This has been fixed in `pdp` (version 0.3.0) by adding an informative error message listing which variables are the cause of the issue.
-```{r, error=TRUE}
-library(pdp)
-data(trees)
-mod <- lm(Volume~., data=trees)
-partial(mod, pred.var=c("Girth", "foo", "bar"), plot=TRUE)
-```
-
-### Comment 1.7
-On page 2 of the paper it says "The columns of the data frame are labeled in the same order as the features supplied to `pred.var`, and the last column is always labeled `y`".
-What happens if `y` is one of the predictors?
-
-### Response 1.7
-The last column is now labeled `yhat`, which is less likely to be a predictor name.
-When one of the predictors is labeled `"y"` `randomForest::partialPlot` fails while `gbm::plot.gbm` does not.
-I think regardless of what the default column name is, there is always the possibility of this problem.
-This has been filed as an issue on the `pdp` GitHib page: https://github.com/bgreenwell/pdp/issues/22.
-
-### Comment 1.8
-Hastie et al.
-in the book cited in the paper say "Although such a collection [of partial dependence plots] can seldom provide a comprehensive depiction of the approximation, it can often produce helpful clues".
-Therefore in in the abstract for the paper, to avoid overselling consider changing "relationship between the outcome and predictors of interest can be easily understood" to "relationship between the outcome and predictors of interest can be more easily understood".
-
-### Response 1.8
-Fixed. Thanks for the suggestion.
-
-### Comment 1.9
-On page 3 in the PDF of the paper, add a newline to the error message in the paper:
-```
-Error: The training data could not be extracted from object. Please
-supply the raw training data using the ...
-```
-
-### Response 1.9
-Fixed. Thanks for pointing that out.
-
-### Comment 1.10
-Is there any reason the package is called `pdp` but the main function is called `partial`?
-Consider giving the function the same name as the package, so e.g., the following works (to help the user get started)
-```{r, eval=FALSE}
-library(pdp)
-?pdp  # fails
-```
-
-### Response 1.10
-The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letters).
-So, I decided to change the name of the package to `pdp`; besides, I think `partial` and `plotPartial` work well together as function names.
-However, I did add a help page describing the package, main functions, etc.
-So, `?pdp` no longer fails.
-However, I am not opposed to creating an alias to `partial` called `pdp` in the future.
-
-### Comment 1.11
-`partial` should return the same value regardless of whether or not the `plot` argument is used?
-
-### Response 1.11
-I dont think so, because I want `partial` to behave like a `lattice` function when using `plot = TRUE`; that is, I want it to return a `"trellis"` object.
-However, I did add the partial dependence data as an attribute to the returned `"trellis"` object so that it can easily be extracted if necessary (e.g., if they do not like the plot and don't want to re-run partial).
-For example,
-```{r}
-library(pdp)
-data(trees)
-mod <- lm(Volume ~ ., data = trees)
-pd.Girth <- partial(mod, pred.var = "Girth")  # data only
-pdp.Girth <- partial(mod, pred.var = "Girth", plot = TRUE)  # plot
-identical(pd.Girth, attr(pdp.Girth, "partial.data"))  # should be identical
-```
-Also, the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded.
-This would be useful if the user did not store the PDP but needs to retrieve the outout data to avoid re-running `partial` (e.g., when computation time is a concern).
-For example,
-```{r, fig.width=4, fig.height=4}
-library(pdp)
-data(trees)
-mod <- lm(Volume ~ ., data = trees)
-partial(mod, pred.var = "Girth", plot = TRUE)
-pdp.Girth <- lattice::trellis.last.object()
-pd.Girth <- attr(pdp.Girth, "partial.data")
-head(pd.Girth)
-```
-This has been noted in the package documentation (see `?partial`). Thanks for the idea!
-
-### Comment 1.12
-A call to say `par(mfrow = c(2, 2))` before calling `plotPartial` is ignored by `plotPartial`.  This is a pity because it would allow multiple plots to be put on one page like this:
-```{r, eval=FALSE}
-par(mfrow = c(1, 2))
-partial(ozone.rf, pred.var = "Temp", plot = TRUE)
-partial(ozone.rf, pred.var = "Wind", plot = TRUE)
-```
-
-### Response 1.12
-The command `par(mfrow = c(2, 2))` does not work with `"trellis"` objects, such as those produced by `lattice`---which `partial` relies on for producing its plots.
-Other methods are available.
-For convenience, `pdp` now imports the `grid.arrange` function from the `gridxtra` package (this has been discussed in the manuscript and incorporated into the examples where appropriate).
-```{r, include=FALSE}
-library(randomForest)
-data(airquality)
-ozone.rf <- randomForest(Ozone ~ ., data = airquality, na.action = na.omit)
-library(pdp)
-```
-```{r, eval=TRUE, fig.width=8, fig.height=4}
-grid.arrange(
-  partial(ozone.rf, pred.var = "Temp", plot = TRUE),
-  partial(ozone.rf, pred.var = "Wind", plot = TRUE),
-  ncol = 2
-)
-```
-
-However, since `partial` returns a data frame by default, using `par` would work in the following case
-```{r, eval=TRUE, fig.width=8, fig.height=4}
-par(mfrow = c(1, 2))
-plot(partial(ozone.rf, pred.var = "Temp"), type = "l")
-plot(partial(ozone.rf, pred.var = "Wind"), type = "l")
-```
-
-### Comment 1.13
-It would be nice if the functions automatically plotted all the (important) variables in a grid of plots, without forcing the user to explicitly specify them.
-
-### Response 1.13
-I agree, this would be a nice feature, but the problems I see are (1) not all models naturally emit a "variable importance" score; the `caret` package has come a long way in this regard, and (2) too many plots will look cluttered.
-At this point, I'd rather encourage the user to select which variables they would like to see plotted.
-I am not opposed to including this feature in the future.
-I added this as a feature request: https://github.com/bgreenwell/pdp/issues/19.
-Thanks for the suggestion!
-
-### Comment 1.14
-It would be nice if the functions automatically got the data from the model used to build the model in more cases without forcing the user to explicitly specify the data.
-Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only---one can't access the data stored in `xgb.DMatrix`).
-
-### Response 1.14
-`kernlab::ksvm` does not store a copy of the training data either, as far as I can tell.
-As time goes on, the coverage will certainly improve, which will make it easier on the user.
-For example, `partial` now works with `caret` which, by default, stores the taining data in a component called `trainingData`.
-Also, I suspect that the package maintainers for `xgboost` (and similar packages designed for working with "large" data sets) would be hesitent in storing a copy of the training data with the fitted model for memory reasons.
-
-### Comment 1.15
-Consider adding a `slowtests` directory (below `pdp/inst/` or elsewhere) that has tests for all the models supported by the package (Table 1 in the paper) to easily check back compatibility when packages change.
-(For justification for such tests see for example my comments on `gbm` back compatibility above.)  These tests will probably be quite slow, but that won't be an issue because they won't be invoked by CRAN check; the slow tests would be just for the maintainer of the `pdp` package to run manually.
-
-### Response 1.15
-This is a good suggestion and is currently under development: https://github.com/bgreenwell/pdp/blob/master/slowtests/slowtests.R.
-In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.
-
-# Reviewer 2
+Please find below a copy of your comments along with my responses.
 

### 2020-02-23 rjournal/greenwell.tex
refactor code, etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Front matter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{pdp: An R Package for Constructing Partial Dependence Plots}
\author{by Brandon M. Greenwell}

\maketitle

\abstract{
Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality). Unfortunately, it can be challenging to understand the results of such models and explain them to management. Partial dependence plots offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be more easily understood. These plots are especially useful in explaining the output from black box models. In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}

\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data. One of their goals was to find a housing value equation using data on median home values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see \citet[Table IV]{harrison-1978-hedonic} for a description of each variable. The data violate many classical assumptions like linearity, normality, and constant variance. Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$). Part of the payoff for there time and efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
\begin{equation}
\label{eqn:boston}
\begin{aligned}
\widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
  & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
  & \quad - 0.012 CRIM + 8.03\times10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS \\
  & \quad - 0.0064 NOX^2
\end{aligned}
\end{equation}

Nowadays, many supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy. (We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation since these algorithms typically do not produce simple prediction formulas like Equation~\eqref{eqn:boston}. These models can still provide insight into the data, but it is not in the form of simple equations. For example, quantifying predictor importance has become an essential task in the analysis of "big data", and many supervised learning algorithms, like tree-based methods, can naturally assign variable importance scores to all of the predictors in the training data.

While determining predictor importance is a crucial task in any supervised learning problem, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2001-greedy} for details. PDPs help visualize the relationship between a subset of the features (typically 1-3) and the response while accounting for the average effect of the other predictors in the model. They are particularly effective with black box models like random forests and support vector machines.

Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_p\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$. If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_c = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the "partial dependence" of the response on $\boldsymbol{z}_s$ is defined as
\begin{equation}
\label{eqn:avg_fun}
  f_s\left(\boldsymbol{z}_s\right) = E_{\boldsymbol{z}_c}\left[\widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)\right] = \int \widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)p_{c}\left(\boldsymbol{z}_c\right)d\boldsymbol{z}_c,
\end{equation}
where $p_{c}\left(\boldsymbol{z}_c\right)$ is the marginal probability density of $\boldsymbol{z}_c$: $p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s$.
Equation~\eqref{eqn:avg_fun} can be estimated from a set of training data by
\begin{equation}
\label{eqn:pdf}
\bar{f}_s\left(\boldsymbol{z}_s\right) = \frac{1}{n}\sum_{i = 1}^n\widehat{f}\left(\boldsymbol{z}_s,\boldsymbol{z}_{i, c}\right),
\end{equation}
where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, we average out the effects of all the other predictors in the model.

Constructing a PDP \eqref{eqn:pdf} in practice is rather straightforward. To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:

\begin{algorithm}
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
\caption{A simple algorithm for constructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
\end{algorithm}
Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records. Fortunately, the algorithm can be parallelized quite easily (more on this in Section~\ref{sec:computational}). It can also be easily extended to larger subsets of two or more features as well.

Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm} \citep{gbm-pkg}, among others; these are limited in the sense that they only apply to the models fit using the respective package. For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"} and the \code{plot} function in \pkg{gbm} only applies to \code{"gbm"} objects. While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space. Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., generalized additive models and neural networks). However, to our knowledge, there is no general package for constructing PDPs in R. For example, PDPs for a conditional random forest as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively. The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for constructing PDPs that can be applied to several classes of fitted models.

The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}. According to \citeauthor{plotmo-pkg}, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level). These plots allow for up to two variables at a time. They are also less accurate than PDPs, but are faster to construct. For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs. As of \pkg{plotmo} version 3.3.0, there is now support for constructing PDPs, but it is not the default. The main difference is that \pkg{plotmo}, rather than applying step 1. (a)-(c) in Algorithm~\ref{alg:pdp}, accumulates all the data at once thereby reducing the number of internal calls to \code{predict}. The trade-off is a slight increase in speed at the expense of using more memory. So, why use the \pkg{pdp} package? As will be discussed in the upcoming sections, \pkg{pdp}:
\begin{itemize}
  \item contains only a few functions with relatively few arguments;
  \item does \underline{not} produce a plot by default;
  \item can be used more efficiently with \code{"gbm"} objects (see Section~\ref{sec:computational});
  \item produces graphics based on \CRANpkg{lattice} \citep{lattice-pkg}, which are more flexible than base R graphics;
  \item defaults to using false color level plots for multivariate displays (see Section~\ref{sec:multi});
  \item contains options to mitigate the risks associated with extrapolation (see Section~\ref{sec:computational});
  \item has the option to display progress bars (see Section~\ref{sec:computational});
  \item has the option to construct PDPs in parallel (see Section~\ref{sec:computational});
    \item is extremely flexible in the types of PDPs that can be produced (see Section~\ref{sec:prediction}),
\end{itemize}

PDPs can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}. To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots---available in the \CRANpkg{ICEbox} package. ICE plots display the estimated relationship between the response and a predictor of interest for each observation. Consequently, the PDP for a predictor of interest can be obtained by averaging the corresponding ICE curves across all observations. In Section~\ref{sec:prediction}, it is shown how to obtain ICE curves using the \pkg{pdp} package. It is also possible to display the PDP for a single predictor with \pkg{ICEbox}; see \code{?ICEbox::plot.ice} for an example. \pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor. The ability to construct centered ICE (c-ICE) plots and derivative ICE (d-ICE) plots is also available in \pkg{ICEbox}; c-ICE plots help visualize heterogeneity in the modeled relationship between observations, and d-ICE plots help to explore interaction effects.

Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model. For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots. \dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provide tabular and graphical displays for the terms in parametric models while holding all other predictors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots. However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Constructing PDPs in R
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constructing PDPs in R}
\label{sec:boston}

The \pkg{pdp} package is useful for constructing PDPs for many classes of fitted models in R. PDPs are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest. The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}. The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}. Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}. The two most important functions exported by \pkg{pdp} are:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}

The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the fitted model and predictors are specified using the \code{object} and \code{pred.var} arguments, respectively---these are the only required arguments. If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function. The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is labeled \code{yhat}\footnote{There is one exception to this. When a function supplied via the \code{pred.fun} argument returns multiple predictions, the second to last and last columns will be labeled \code{yhat} and \code{yhat.id}, respectively (see Section~\ref{sec:prediction}).} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$. If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} (with fewer plotting options) and returns the PDP in the form of a \pkg{lattice} plot (i.e., a \code{"trellis"} object). \strong{Note:}  it is recommended to call \code{partial} with \code{plot = FALSE} and store the results; this allows for more flexible plotting, and the user will not have to waste time calling \code{partial} again if the default plot is not sufficient.

The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options. For example, \code{plotPartial} makes it straight forward to add a LOESS smooth, or produce a 3-D surface instead of a false color level plot (the default). Of course, since the default output produced by \code{partial} is still a data frame, the user can easily use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance (see Section~\ref{sec:classification} and Section~\ref{sec:prediction} for examples).

\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}. \pkg{grid} graphics behave a little differently than traditional R graphics, and two points are worth making (see \code{?lattice} for more details):
\begin{enumerate}
  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display. However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line. If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph; hence, the same is true when using \code{partial} with \code{plot = TRUE}.
  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots. Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
\end{enumerate}
A consequence of the second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots. Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}. For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{grid}-based graphical objects on a single plot (these include graphics produced using \pkg{lattice} (hence, \pkg{pdp}) and \pkg{ggplot2}). This is demonstrated in multiple examples throughout this paper.

Currently supported models are described in Table~\ref{tab:models}. In these cases, the user does not need to supply a prediction function (more on this in Section~\ref{sec:prediction}) or a value for the \code{type} argument (i.e., \code{"regression"} or \code{"classification"}). In other situations, the user may need to specify one or both of these arguments. This allows \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002}.% and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.

\begin{table}[!htbp]
  \begin{tabular}{p{4cm}ll}
    \toprule
      Type of model & R package & Object class \\
      \midrule
      Decision tree             & \CRANpkg{C50} \citep{C50-pkg} & \code{"C5.0"} \\
                                & \pkg{party}    & \code{"BinaryTree"} \\
                                & \pkg{partykit} & \code{"party"} \\
                                & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
      Bagged decision trees     & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
                                & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \\ & & \code{"regbagg"} \\
      Boosted decision trees    & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
                                & \pkg{gbm}      & \code{"gbm"} \\
                                & \CRANpkg{xgboost} & \code{"xgb.Booster"} \\
      Cubist                    & \CRANpkg{Cubist} \citep{Cubist-pkg} & \code{"cubist"} \\
      Discriminant analysis     & \CRANpkg{MASS} \citep{venables-modern-2002} & \code{"lda"}, \code{"qda"} \\
      Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
      Linear model              & \pkg{stats}    & \code{"lm"} \\
      Nonlinear least squares   & \pkg{stats}    & \code{"nls"} \\
      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
                                & \CRANpkg{mda} \citep{mda-pkg} & \code{"mars"} \\
      Projection pursuit regression & \pkg{stats} & \code{"ppr"} \\
      Random forest             & \pkg{randomForest} & \code{"randomForest"} \\
                                & \pkg{party}        & \code{"RandomForest"} \\
                                & \pkg{partykit} & \code{"cforest"} \\
                                & \CRANpkg{ranger} \citep{ranger-pkg} & \code{"ranger"} \\
      Support vector machine    & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                                & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
      \bottomrule
  \end{tabular}
  \caption{Models specifically supported by the \pkg{pdp} package. \strong{Note:} for some of these cases, the user may still need to supply additional arguments in the call to \code{partial}.}
  \label{tab:models}
\end{table}

The \code{partial} function also supports objects of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}. This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}. An example is given in Section~\ref{sec:xgboost}.

Another important argument to \code{partial} is \code{train}. If \code{train = NULL} (the default), \code{partial} tries to extract the original training data from the fitted model object. For objects that typically store a copy of the training data (e.g., objects of class \code{"BinaryTree"}, \code{"RandomForest"}, and \code{"train"}), this is straightforward. Otherwise, \code{partial} will attempt to extract the call stored in \code{object} (if available) and use that to evaluate the training data in the same environment from which \code{partial} was called. This can cause problems when, for example, the training data have been changed after fitting the model, but before calling \code{partial}. Hence, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}\footnote{For brevity, we ignore this option in most of the examples in this paper.}. If \code{train = NULL} and the training data can not be extracted from the fitted model, the user will be prompted with an informative error message (this will occur, for example, when using \code{partial} with \code{"ksvm"} and \code{"xgb.Booster"} objects):
\begin{example}
Error: The training data could not be extracted from object. Please supply
the raw training data using the `train` argument in the call to `partial`.
\end{example}

For illustration, we will use a corrected version of the Boston housing data analyzed in \citet{harrison-1978-hedonic}; the data are available in the \pkg{pdp} package (see \code{?pdp::boston} for details). We begin by loading the data and fitting a random forest with default tuning parameters and 500 trees:
\begin{example}
data(boston, package = "pdp")  # load the (corrected) Boston housing data
library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # Figure 1
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_rf_vimp}
  \caption{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}
  \label{fig:plotmo_vs_partial}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Single predictor PDPs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Single predictor PDPs}

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor---the keywords here are "single predictor". For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
\begin{example}
library(pdp)  # for partial, plotPartial, and grid.arrange functions
partial(boston.rf, pred.var = "lstat", plot = TRUE)  # Figure 2 (left)
\end{example}
The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.

For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame. This is illustrated in the example below which increases the line width, adds a LOESS smooth, and customizes the $y$-axis label. The result is displayed in the right side of Figure~\ref{fig:pd_lstat}. \strong{Note:} to encourage writing more readable code, the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported whenever \pkg{pdp} is loaded.
\begin{example}
# Figure 2 (right)
boston.rf %>%  # the %>% operator is read as "and then"
  partial(pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} based on a random forest. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
  \label{fig:pd_lstat}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Multi-predictor PDPs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multi-predictor PDPs}
\label{sec:multi}

The benefit of using \code{partial} is threefold: (1) it is a flexible, generic function that can be used to obtain different kinds of PDPs for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The \code{grid.arrange} function is used to display three PDPs, which make use of various \code{plotPartial} options\footnote{See Section~\ref{sec:computational} for an example of how to add a label to the colorkey in these types of graphs.}, on the same graph. The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
# Compute partial dependence data for lstat and rm
pd <- partial(boston.rf, pred.var = c("lstat", "rm"))

# Default PDP
pdp1 <- plotPartial(pd)

# Add contour lines and use a different color palette
rwb <- colorRampPalette(c("red", "white", "blue"))
pdp2 <- plotPartial(pd, contour = TRUE, col.regions = rwb)

# 3-D surface
pdp3 <- plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
                    colorkey = TRUE, screen = list(z = -20, x = -60))

# Figure 3
grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
\end{example}
\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a random forest. \textit{Left}: Default plot. \textit{Middle}: With contour lines and a different color palette. \textit{Right}: Using a 3-D surface.}
  \label{fig:pd_lstat_rm}
\end{widefigure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Avoiding extrapolation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Avoiding extrapolation}
\label{sec:extrapolation}

It is not wise to draw conclusions from PDPs in regions outside the area of the training data. Here we describe two ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls. Rug displays are one-dimensional plots added to the axes. Both \code{partial} and \code{plotPartial} have a \code{rug} option that, when set to \code{TRUE}, will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and vertical axes. The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
\begin{example}
# Figure 4 (left)
partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
\end{example}

In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on. When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted. Over interpreting the PDP outside of this region is considered extrapolation and is ill-advised. The right display in Figure~\ref{fig:partial_extrap} was produced using:
\begin{example}
# Figure 4 (right)
partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_extrap}
  \caption{Examples of PDPs with the addition of a rug display (left) and a convex hull (right).}
  \label{fig:partial_extrap}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Addressing computational concerns
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Addressing computational concerns}
\label{sec:computational}

Constructing PDPs can be quite computationally expensive\footnote{The exception is regression trees based on single-variable splits which can make use of the efficient weighted tree traversal method described in \citet{friedman-2001-greedy}, however, only the \pkg{gbm} package seems to make use of this approach; consequently, \pkg{pdp} can also exploit this strategy when used with \pkg{gbm} models (see \code{?partial} for details).} Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; \code{grid.arrange} is used to display all three PDPs on the same graph, side by side. The results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
# Figure 5
grid.arrange(
  partial(boston.rf, "rm", plot = TRUE),
  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
  ncol = 3
)
\end{example}

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_manual}
  \caption{Partial dependence of \code{cmedv} on \code{rm}. \textit{Left}: Default plot. \textit{Middle}: Using a reduced grid size. \textit{Right}: Using a user-specified grid.}
  \label{fig:partial_manual}
\end{widefigure}

The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops. This makes it easy to request progress bars (e.g., \code{progress = "text"}) or run \code{partial} in parallel. In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package. To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].

To illustrate, we will use the Los Angeles ozone pollution data described in \citet{breiman-1985-estimating}. The data contain daily measurements of ozone concentration (\code{ozone}) along with eight meteorological quantities for 330 days in the Los Angeles basin in 1976.\footnote{The data are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.data}. Details, including variable information, are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.info}.} The following code chunk loads the data into R:
\begin{example}
ozone <- read.csv(paste0("http://statweb.stanford.edu/~tibs/ElemStatLearn/",
                         "datasets/LAozone.data"), header = TRUE)
\end{example}

Next, we use the multivariate adaptive regression splines (MARS) algorithm introduced in \citet{friedman-1991-mars} to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
\begin{example}
library(earth)  # for earth function (i.e., MARS algorithm)
ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
summary(ozone.mars)
\end{example}
The MARS model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
\begin{itemize}
  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
\end{itemize}
To understand this interaction, we can use a PDP. However, since the partial dependence between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.

Setting up a parallel backend is rather straightforward. To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
\begin{example}
library(doParallel)  # load the parallel backend
cl <- makeCluster(4)  # use 4 workers
registerDoParallel(cl)  # register the parallel backend
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}. See \code{?plyr::adply} for details.}. This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel. The last three lines of code add a label to the colorkey. The result is displayed in Figure~\ref{fig:partial_par}. \strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
\begin{example}
partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
stopCluster(cl)  # good practice

# Add a label to the colorkey
lattice::trellis.focus("legend", side = "right", clipp.off = TRUE, highlight = FALSE)
grid::grid.text("ozone", x = 0.2, y = 1.05, hjust = 0.5, vjust = 1)
lattice::trellis.unfocus()
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_par}
  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}. Since \code{dpg} is continuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
  \label{fig:partial_par}
\end{figure}

It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display. The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels. If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle. The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}). Hence, it will be more effective to use categorical variables to define the panels in higher dimensional displays when possible.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Classification problems
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Classification problems}
\label{sec:classification}

Traditionally, for classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp. 369---370]{hastie-elements-2009}. Suppose the response is categorical with $K$ levels, then for each class we compute
\begin{equation}
\label{eqn:avg-logit}
f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{k = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
\end{equation}
where $p_k(x)$ is the predicted probability for the $k$-th class. Plotting $f_k(x)$ helps us understand how the log-odds for the $k$-th class depends on different subsets of the predictor variables.

To illustrate, we consider Edgar Anderson's iris data from the \pkg{datasets} package. The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris: setosa, versicolor, and virginica. We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} package (the tuning parameters were determined using 5-fold cross-validation).
\begin{example}
library(e1071)  # for svm function
iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
                cost = 0.25, probability = TRUE)
\end{example}
\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set \code{probability = TRUE} in the call to \code{svm}.

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                 which.class = i, grid.resolution = 101, progress = "text")
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}

# Figure 7
library(ggplot2)
ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = yhat, fill = yhat)) +
  geom_tile() +
  geom_contour(color = "white", alpha = 0.5) +
  scale_fill_distiller(name = "Centered\nlogit", palette = "Spectral") +
  theme_bw() +
  facet_grid(~ Species)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} for the iris data.}
  \label{fig:partial_iris}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% User-defined prediction functions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{User-defined prediction functions}
\label{sec:prediction}

PDPs are essentially just averaged predictions; this is apparent from step 1. (c) in Algorithm~\ref{alg:pdp}. Consequently, as pointed out by \citet{goldstein-peeking-2015}, strong heterogeneity can conceal the complexity of the modeled relationship between the response and predictors of interest. This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE plot procedure.

With \code{partial} it is possible to replace the mean in step 1. (c) of Algorithm~\ref{alg:pdp} with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale. It is even possible to obtain ICE curves. This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0). This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}. The supplied prediction function must return either a single prediction or a vector of predictions. Returning the mean of all the predictions will result in the traditional PDP. Returning a vector of predictions (i.e., one for each observation) will result in a set of ICE curves. The examples below illustrate.

Using the \code{pred.fun} argument, it is possible to obtain PDPs for classification problems on the probability scale. We just need to write a function that computes the predicted class probability of interest averaged across all observations. The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:classification} to extract the average predicted probability of belonging to the \code{Setosa} class.
\begin{example}
pred.prob <- function(object, newdata) {  # see ?predict.svm
  pred <- predict(object, newdata, probability = TRUE)
  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
  mean(prob.setosa)
}
\end{example}
Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The following chunk of code uses \code{pred.prob} to obtain PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale. The results are displayed in Figure~\ref{fig:partial_iris_prob}.
\begin{example}
# PDPs for Petal.Width and Petal.Length on the probability scale
pdp.pw <- partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob,
                  plot = TRUE)
pdp.pl <- partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob,
                  plot = TRUE)
pdp.pw.pl <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
                     pred.fun = pred.prob, plot = TRUE)

# Figure 8
grid.arrange(pdp.pw, pdp.pl, pdp.pw.pl, ncol = 3)
\end{example}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probability of belonging to the setosa species.}
  \label{fig:partial_iris_prob}
\end{figure}

For regression problems, the default prediction function is essentially
\begin{example}
pred.fun <- function(object, newdata) {
  mean(predict(object, newdata), na.rm = TRUE)
}
\end{example}
This corresponds to step step 1. (c) in Algorithm~\ref{alg:pdp}. Suppose we would like ICE curves instead. To accomplish this we need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata} (i.e., just remove the call to \code{mean} in \code{pred.fun}). The code snippet below illustrates this for the Boston housing example using the predictor \code{rm}. The result is displayed in Figure~\ref{fig:ice_boston}. \strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in the following code chunk, there will be one curve for each observation in \code{boston}.
\begin{example}
# Use partial to obtain ICE curves
pred.ice <- function(object, newdata) predict(object, newdata)
rm.ice <- partial(boston.rf, pred.var = "rm", pred.fun = pred.ice)

# Figure 9
plotPartial(rm.ice, rug = TRUE, train = boston, alpha = 0.3)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{ice_boston}
  \caption{ICE curves depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. Each curve corresponds to a different observation.}
  \label{fig:ice_boston}
\end{figure}

The curves in Figure~\ref{fig:ice_boston} indicate some heterogeneity in the fitted model (i.e., some of the curves depict the opposite relationship). Such heterogeneity can be easier to spot using c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}. Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process the output from \code{partial} to obtain c-ICE curves (similar to the construction of \dfn{raw change scores} \citep[pg. 130]{fitzmaurice-2011-applied} for longitudinal data). This is shown below.
\begin{example}
# Post-process rm.ice to obtain c-ICE curves
library(dplyr)  # for group_by and mutate functions
rm.ice <- rm.ice %>%
  group_by(yhat.id) %>%  # perform next operation within each yhat.id
  mutate(yhat.centered = yhat - first(yhat))  # so each curve starts at yhat = 0
\end{example}

Since the PDP is just the average of the corresponding ICE curves, it is quite simple to display both on the same plot. This is easily accomplished using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together. The code snippet below plots the ICE curves and c-ICE curves, along with their averages, for the predictor \code{rm} in the Boston housing example. The results are displayed in Figure~\ref{fig:ice_cice_boston}.
\begin{example}
# ICE curves with their average
p1 <- ggplot(rm.ice, aes(rm, yhat)) +
  geom_line(aes(group = yhat.id), alpha = 0.2) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)

# c-ICE curves with their average
p2 <- ggplot(rm.ice, aes(rm, yhat.centered)) +
  geom_line(aes(group = yhat.id), alpha = 0.2) +
  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)

# Figure 10
grid.arrange(p1, p2, ncol = 2)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{ice_cice_boston}
  \caption{ICE curves (black curves) and their average (red curve) depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. \textit{Left}: Uncentered (here the red curve is just the traditional PDP). \textit{Right}: Centered.}
  \label{fig:ice_cice_boston}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Using partial with xgboost
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Using partial with the XGBoost library}
\label{sec:xgboost}

To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool. XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable. The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}. It has been shown to be many times faster than the well-known \pkg{gbm} package. However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for constructing PDPs. Fortunately, the \pkg{pdp} package can be used to fill this gap.

For illustration, we return to the Boston housing example. The code chunk below uses \pkg{caret} to tune an \pkg{xgboost} model using 10-fold cross-validation. (After loading \pkg{caret}, use \code{getModelInfo("xgbTree")} for information on tuning \pkg{xgboost} models.) \strong{Warning:} The following code chunk may take a few minutes to run.
\begin{example}
# Tune an XGBoost model using 10-fold cross-validation
library(caret)  # functions related to classification and regression training
set.seed(202)  # for reproducibility
boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                    y = boston$cmedv, method = "xgbTree", metric = "Rsquared",
                    trControl = trainControl(method = "cv", number = 10),
                    tuneLength = 10)
\end{example}
The optimal model had a cross-validated $R^2$ of $0.902$ (use \code{print(boston.xgb\$bestTune)} to view the optimum tuning parameters). The next snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}.
\begin{example}
# PDPs for lstat and rm
pdp.lstat <- partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE)
pdp.rm <- partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE)
pdp.lstat.rm <- partial(boston.xgb, pred.var = c("lstat", "rm"),
                        plot = TRUE, chull = TRUE)

# Figure 11
grid.arrange(pdp.lstat, pdp.rm, pdp.lstat.rm, ncol = 3)
\end{example}
\begin{figure}[!htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_xgb}
  \caption{PDPs for the top two most important variables in the Boston housing data using \pkg{xgboost}. Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
  \label{fig:boston_xgb}
\end{figure}
%\code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}

The \code{train} function creates objects of class \code{"train"}, whereas the \code{xgboost} function creates objects of class \code{"xgb.Booster"}. Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object, there is no need to supply it in the call to \code{partial} in this example. However, this is not the case when using the \pkg{xgboost} package directly. To illustrate, we fit the same model using the \code{xgboost} function with the optimum tuning parameters found previously using \pkg{caret}. %stored in \code{boston.xgb\$bestTune}.%
\begin{example}
library(xgboost)  # for xgboost function
set.seed(203)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
                      colsample_bytree = 0.8, min_child_weight = 1,
                      subsample = 0.9444444)
\end{example}
To use \code{partial} with \code{"xgb.Booster"} objects, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown). (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
\begin{example}
partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
        train = subset(boston, select = -cmedv))
\end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Summary
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}

PDPs can be used to graphically examine the dependence of the response on low cardinality subsets of the features, accounting for the average effect of the other predictors. In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package. We also briefly discussed related approaches available in other R packages. Suggestions to avoid extrapolation and high execution times were discussed and demonstrated via examples.

In terms of future development, \pkg{pdp} can be expanded in a number of ways. For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response. It would also be worthwhile to implement the partial dependence-based $H$-statistic \citep{friedman-2008-predictive} for assessing the strength of interaction between predictors.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Summary
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Acknowledgments}

The author would like to thank two anonymous reviewers and the Editor for their helpful comments
and suggestions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Back matter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliography{greenwell}

\address{Brandon M. Greenwell\\
  Infoscitex, a DCS Company\\
  4027 Colonel Glenn Highway\\
  Suite 210\\
  Dayton, OH 45431-1672\\
  United States of America\\}
\email{greenwell.brandon@gmail.com}

### 2020-02-23 R/autoplot.R

@@ -255,11 +255,11 @@ ggplot_ice_curves <- function(
     }
 
     # Add rug plot to x-axis
-    if (rug) {
+    if (isTRUE(rug)) {
       if (is.null(train)) {
         stop("The training data must be supplied for rug display.")
       } else {
-        x.name <- which(names(train) == names(object)[[1L]])
+        x.name <- which(colnames(train) == names(object)[[1L]])
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))
@@ -311,11 +311,11 @@ ggplot_one_predictor_pdp <- function(
       geom_line(...)
 
     # Add rug plot to x-axis
-    if (rug) {
+    if (isTRUE(rug)) {
       if (is.null(train)) {
         stop("The training data must be supplied for rug display.")
       } else {
-        x.name <- which(names(train) == names(object)[[1L]])
+        x.name <- which(colnames(train) == names(object)[[1L]])
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))
@@ -378,11 +378,11 @@ ggplot_two_predictor_pdp <- function(
       facet_wrap(~ object[[1L]])
 
     # Add rug plot to the x-axis
-    if (rug) {
+    if (isTRUE(rug)) {
       if (is.null(train)) {
         stop("The training data must be supplied for rug display.")
       } else {
-        x.name <- which(names(train) == names(object)[[2L]])
+        x.name <- which(colnames(train) == names(object)[[2L]])
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))
@@ -407,11 +407,11 @@ ggplot_two_predictor_pdp <- function(
       facet_wrap(~ object[[2L]])
 
     # Add rug plot to x-axis
-    if (rug) {
+    if (isTRUE(rug)) {
       if (is.null(train)) {
         stop("The training data must be supplied for rug display.")
       } else {
-        x.name <- which(names(train) == names(object)[[1L]])
+        x.name <- which(colnames(train) == names(object)[[1L]])
         x.rug <- data.frame(as.numeric(
           stats::quantile(train[, x.name, drop = TRUE], probs = 0:10/10,
                           na.rm = TRUE)))

### 2020-02-23 R/getParDepReg.R

@@ -1,102 +1,0 @@
-#' @keywords internal
-getParDepReg <- function(object, pred.var, pred.grid, inv.link, ice, train,
-                         progress, parallel, paropts, ...) {
-
-  # User-supplied inverse link function
-  if (!is.null(inv.link)) {
-
-    # ICE curves with user-specified inverse link function
-    if (ice) {
-
-      # Get predictions
-      plyr::adply(
-        .data = pred.grid,
-        .margins = 1,
-        .progress = progress,
-        .parallel = parallel,
-        .paropts = paropts,
-        .id = NULL,
-        .fun = function(x) {
-          temp <- train
-          temp[, pred.var] <- x
-          pred <- getIcePredRegInvLink(
-            object = object, newdata = temp, inv.link = inv.link, ...
-          )
-          if (is.null(names(pred))) {
-            stats::setNames(pred, paste0("yhat.", 1L:length(pred)))
-          } else {
-            stats::setNames(pred, paste0("yhat.", names(pred)))
-          }
-        })
-
-    # PDP with user-specified inverse link function
-    } else {
-
-      # Get predictions
-      plyr::adply(
-        .data = pred.grid,
-        .margins = 1,
-        .progress = progress,
-        .parallel = parallel,
-        .paropts = paropts,
-        .id = NULL,
-        .fun = function(x) {
-          temp <- train
-          temp[, pred.var] <- x
-          stats::setNames(
-          getParPredRegInvLink(
-            object = object, newdata = temp, inv.link = inv.link, ...), "yhat"
-          )
-        })
-
-    }
-
-  # Default
-  } else {
-
-    # Default Ice curves
-    if (ice) {
-
-      # Get predictions
-      plyr::adply(
-        .data = pred.grid,
-        .margins = 1,
-        .progress = progress,
-        .parallel = parallel,
-        .paropts = paropts,
-        .id = NULL,
-        .fun = function(x) {
-          temp <- train
-          temp[, pred.var] <- x
-          pred <- getIcePredReg(object, newdata = temp, ...)
-          if (is.null(names(pred))) {
-            stats::setNames(pred, paste0("yhat.", 1L:length(pred)))
-          } else {
-            stats::setNames(pred, paste0("yhat.", names(pred)))
-          }
-        }
-      )
-
-    # Default PDP
-    } else {
-
-      # Get predictions
-      plyr::adply(
-        .data = pred.grid,
-        .margins = 1,
[…]

### 2020-02-23 R/getParPredCls.R

@@ -1,335 +1,0 @@
-#' @keywords internal
-getParClsLogit <- function(object, newdata, which.class, ...) {
-  UseMethod("getParClsLogit")
-}
-
-
-#' @keywords internal
-getParClsProb <- function(object, newdata, which.class, ...) {
-  UseMethod("getParClsProb")
-}
-
-
-#' @keywords internal
-getParClsLogit.default <- function(object, newdata, which.class, ...) {
-  pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
-  mean(multiclass_logit(pr, which.class = which.class), na.rm = TRUE)
-}
-
-
-#' @keywords internal
-getParClsProb.default <- function(object, newdata, which.class, ...) {
-  pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
-  mean(pr[, which.class], na.rm = TRUE)
-}
-
-
-#' @keywords internal
-getParClsLogit.BinaryTree <- function(object, newdata, which.class, ...) {
-  pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
-  mean(multiclass_logit(do.call(rbind, pr), which.class = which.class),
-       na.rm = TRUE)
-}
-
-
-#' @keywords internal
-getParClsProb.BinaryTree <- function(object, newdata, which.class, ...) {
-  pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
-  mean(do.call(rbind, pr)[, which.class], na.rm = TRUE)
-}
-
-
-#' @keywords internal
-getParClsLogit.bagging <- function(object, newdata, which.class, ...) {
-  pr <- stats::predict(object, newdata = newdata, ...)$prob
-  mean(multiclass_logit(pr, which.class = which.class), na.action = TRUE)
-}
-
-
-#' @keywords internal
-getParClsProb.bagging <- function(object, newdata, which.class, ...) {
-  pr <- stats::predict(object, newdata = newdata, ...)$prob
-  mean(pr[, which.class], na.rm = TRUE)
-}
-
-
-#' @keywords internal
-getParClsLogit.boosting <- function(object, newdata, which.class, ...) {
-  pr <- stats::predict(object, newdata = newdata, ...)$prob
-  mean(multiclass_logit(pr, which.class = which.class), na.rm = TRUE)
-}
-
-
-#' @keywords internal
-getParClsProb.boosting <- function(object, newdata, which.class, ...) {
[…]

### 2020-02-23 R/get_feature_effects.R

@@ -1,6 +1,94 @@
 #' @keywords internal
+get_feature_effects <- function(
+  object,
+  pred.var,
+  pred.grid,
+  pred.fun,
+  inv.link,
+  ice,
+  task,
+  which.class,
+  logit,
+  train,
+  progress,
+  parallel,
+  paropts,
+  ...
+) {
+
+  # Compute feature effects
+  if (is.null(pred.fun)) {
+
+    # Get predictions
+    plyr::adply(
+      .data = pred.grid,
+      .margins = 1,
+      .progress = progress,
+      .parallel = parallel,
+      .paropts = paropts,
+      .id = NULL,
+      .fun = function(x) {
+        temp <- train
+        temp[, pred.var] <- x
+        pred <- if (task == "regression") {
+          get_predictions(object, newdata = temp, inv.link = inv.link, ...)
+        } else {
+          get_probs(object, newdata = temp, which.class = which.class,
+                    logit = logit, ...)
+        }
+        if (isTRUE(ice)) {
+          if (is.null(names(pred))) {
+            stats::setNames(pred, paste0("yhat.", 1L:length(pred)))
+          } else {
+            stats::setNames(pred, paste0("yhat.", names(pred)))
+          }
+        } else {  # average predictions
+          mean(pred, na.rm = TRUE)
+        }
+      }
+    )
+
+  } else {
+
+    # Get predictions (using user-supplied prediction wrapper)
+    plyr::adply(
+      .data = pred.grid,
+      .margins = 1,
+      .progress = progress,
+      .parallel = parallel,
+      .paropts = paropts,
+      .id = NULL,
+      .fun = function(x) {
+        temp <- train
+        temp[, pred.var] <- x
+        out <- pred.fun(object, newdata = temp)
+        if (length(out) == 1) {
+          stats::setNames(out, "yhat")
+        } else {
+          if (is.null(names(out))) {
+            stats::setNames(out, paste0("yhat.", 1L:length(out)))
+          } else {
+            stats::setNames(out, paste0("yhat.", names(out)))
+          }
+        }
+      }
+    )
+
+  }
+
+}
+
+
+#' @keywords internal
 #' @useDynLib pdp, .registration = TRUE
-getParDepGBM <- function(object, pred.var, pred.grid, which.class, prob, ...) {
+get_feature_effects_gbm <- function(
+  object,
+  pred.var,
+  pred.grid,
+  which.class,
+  prob,
+  ...
+) {
 
   # Extract number of trees
   dots <- list(...)

### 2020-02-23 R/partial.R

@@ -414,7 +414,8 @@ partial.default <- function(
     # Stop and notify user that pred.fun cannot be used when recursive = TRUE
     # with "gbm" objects
     if (!is.null(pred.fun)) {
-      stop("Option `pred.fun` cannot currently be used when `recursive = TRUE`.")
+      stop("Option `pred.fun` cannot currently be used when ",
+           "`recursive = TRUE`.")
     }
 
     # Notify user that progress bars are not avaiable for "gbm" objects when
@@ -426,12 +427,19 @@ partial.default <- function(
     # Stop and notify user that parallel functionality is currently not
     # available for "gbm" objects when recursive = TRUE
     if (parallel) {
-      stop("Option `parallel` cannot currently be used when `recursive = TRUE`.")
+      stop("Option `parallel` cannot currently be used when ",
+           "`recursive = TRUE`.")
     }
 
     # Use Friedman's weighted tree traversal approach
-    pd.df <- getParDepGBM(object, pred.var = pred.var, pred.grid = pred.grid,
-                          which.class = which.class, prob = prob, ...)
+    pd.df <- get_feature_effects_gbm(
+      object = object,
+      pred.var = pred.var,
+      pred.grid = pred.grid,
+      which.class = which.class,
+      prob = prob,
+      ...
+    )
     class(pd.df) <- c("partial", "data.frame")  # assign class labels
     names(pd.df) <- c(pred.var, "yhat")  # rename columns
     rownames(pd.df) <- NULL  # remove row names
@@ -439,23 +447,27 @@ partial.default <- function(
   } else {
 
     # Use brute force approach
-    pd.df <- if (!is.null(pred.fun)) {  # user-supplied prediction function
-      getParDepMan(object, pred.var = pred.var, pred.grid = pred.grid,
-                   pred.fun = pred.fun, train = train, progress = progress,
-                   parallel = parallel, paropts = paropts, ...)
-    } else if (type == "regression") {
-      getParDepReg(object, pred.var = pred.var, pred.grid = pred.grid,
-                   inv.link = inv.link, ice = ice, train = train,
-                   progress = progress, parallel = parallel,
-                   paropts = paropts, ...)
-    } else if (type == "classification") {
-      getParDepCls(object, pred.var = pred.var, pred.grid = pred.grid,
-                   which.class = which.class, prob = prob, ice = ice,
-                   train = train, progress = progress, parallel = parallel,
-                   paropts = paropts, ...)
[…]

### 2020-02-23 R/plotPartial.R

@@ -257,8 +257,8 @@ plot_ice_curves <- function(
           stop("The training data must be supplied for rug display.")
         } else {
           panel.rug(
-            stats::quantile(train[[names(object)[1L]]], probs = 0:10/10,
-                            na.rm = TRUE)
+            stats::quantile(train[, names(object)[1L], drop = TRUE],
+                            probs = 0:10/10, na.rm = TRUE)
           )
         }
       }
@@ -281,12 +281,12 @@ plot_one_predictor_pdp <- function(object, smooth, rug, train = NULL, ...) {
         if (smooth) {
           panel.loess(xx, yy, ...)
         }
-        if (rug) {
+        if (isTRUE(rug)) {
           if (is.null(train)) {
             stop("The training data must be supplied for rug display.")
           } else {
             panel.rug(
-              stats::quantile(train[[names(object)[1L]]],
+              stats::quantile(train[, names(object)[1L], drop = TRUE],
                               probs = 0:10/10, na.rm = TRUE)
             )
           }
@@ -338,12 +338,13 @@ plot_two_predictor_pdp <- function(
         if (smooth) {
           panel.loess(xx, yy, ...)
         }
-        if (rug) {
+        if (isTRUE(rug)) {
           if (is.null(train)) {
             stop("The training data must be supplied for rug display.")
            } else {
              panel.rug(
-               stats::quantile(train[[names(object)[1L]]], probs = 0:10/10,
+               stats::quantile(train[, names(object)[1L], drop = TRUE],
+                               probs = 0:10/10,
                                na.rm = TRUE)
              )
            }
@@ -380,13 +381,15 @@ plot_two_predictor_pdp <- function(
               }
             }
             # Add a rug display
-            if (rug) {
+            if (isTRUE(rug)) {
               panel.rug(
                 x = stats::quantile(
-                  train[[names(object)[1L]]], probs = 0:10/10, na.rm = TRUE
+                  train[, names(object)[1L], drop = TRUE],
+                  probs = 0:10/10, na.rm = TRUE
                 ),
                 y = stats::quantile(
-                  train[[names(object)[2L]]], probs = 0:10/10, na.rm = TRUE
+                  train[, names(object)[2L], drop = TRUE],
+                  probs = 0:10/10, na.rm = TRUE
                 ),
                 col = "black"
               )
@@ -466,13 +469,13 @@ plot_three_predictor_pdp <- function(
         if (smooth) {
           panel.loess(xx, yy, ...)
         }
[…]

### 2020-02-23 R/utils.R

@@ -178,54 +178,3 @@ train_chull <- function(pred.var, pred.grid, train) {
     pred.grid
   }
 }
-
-
-#' Cross join
-#'
-#' Uses \code{\link[data.table]{data.table}} to efficiently perform a cross
-#' join between two data frames; that is, the Cartesian product of the rows
-#' from two data frames.
-#'
-#' @param x A data frame.
-#'
-#' @param y A data frame.
-#'
-#' @return A \code{\link[data.table]{data.table}} containing the Cartesian
-#' product of the rows from \code{x} and \code{y}.
-#'
-#' @keywords internal
-#'
-#' @import data.table
-crossJoin <- function(x, y) {
-  x <- data.table::as.data.table(x)
-  y <- data.table::as.data.table(y)
-  Reduce(
-    f = function(x, y) cbind(x, y[rep(1:.N, each = nrow(x))]),
-    x = list(x, y)
-  )
-}
-
-
-#' @keywords internal
-#'
-#' @import data.table
-getParDepRegInMem <- function(object, pred.var, pred.grid, train, ...) {
-
-  # Perform a cross join between feature grid and trainin data
-  train[, names(pred.grid)] <- NULL  # remove features of interest
-  ice_data <- crossJoin(train, y = pred.grid)  # Cartesian product
-
-  # Convert ice_data to the same class as train if train is not a data frame
-  if (inherits(train, "matrix")) {
-    ice_data <- data.matrix(ice_data)
-  }
-  if (inherits(train, "dgCMatrix")) {
-    ice_data <- methods::as(data.matrix(ice_data), "dgCMatrix")
-  }
-
-  # Score the data, remove extra columns, and aggregate
-  ice_data[["yhat"]] <- getIcePredReg(object, newdata = ice_data, ...)
-  ice_data <- ice_data[, c(names(pred.grid), "yhat"), with = FALSE]
-  ice_data[, lapply(.SD, FUN = mean), by = names(pred.grid), .SDcols = "yhat"]
-
-}

### 2020-02-23 inst/tinytest/test_partial.R

@@ -1,273 +1,287 @@
-if (require(gbm, quietly = TRUE) && require(ggplot2, quietly = TRUE)) {
-
-  # Load the Ames housing data
-  ames <- as.data.frame(AmesHousing::make_ames())
-
-  # Fit a gbm model
-  set.seed(1918)  # for reproducibility
-  fit <- gbm(
-    Sale_Price ~ .,
-    data = ames,
-    distribution = "gaussian",
-    n.trees = 1500,
-    interaction.depth = 6,
-    shrinkage = 0.1,
-    bag.fraction = 1,
-    cv.folds = 5,  # implicitly uses more cores than R CMD Check will allow
-    n.cores = 2  # set to CRAN max
-  )
-  best_iter <- gbm.perf(fit, method = "cv")
-
-  # Partial dependence w/wo setting approx = TRUE
-  pd1 <- partial(fit, pred.var = "Gr_Liv_Area", n.trees = best_iter,
-                 recursive = FALSE, progress = "progress")
-  pd2 <- partial(fit, pred.var = "Gr_Liv_Area", approx = TRUE,
-                 recursive = FALSE, progress = "progress",
-                 n.trees = best_iter)
-
-  # Basic expectation(s)
-  expect_true(inherits(pd1, what = "partial"))
-  expect_true(inherits(pd2, what = "partial"))
-  expect_identical(dim(pd1), target = c(51L, 2L))
-  expect_identical(dim(pd2), target = c(51L, 2L))
-
-  # Display plots side by side
-  grid.arrange(
-    plotPartial(pd1),
-    plotPartial(pd2),
-    nrow = 1
-  )
-
-  # Users can call plotPartial() to construct lattice-based PDPs
-  pdp1_lattice <- plotPartial(pd1, smooth = TRUE, rug = TRUE, train = ames)
-
-  # Users can call autoplot() to construct ggplot2-based PDPs
-  pdp1_ggplot2 <- autoplot(pd1, smooth = TRUE, rug = TRUE, train = ames)
-
-  # Expectation(s)
-  expect_true(inherits(pdp1_lattice, what = "trellis"))
-  expect_true(inherits(pdp1_ggplot2, what = "ggplot"))
-
-  # Users should be able to supply the predictor name(s) or column position(s)
-  # in train
-  expect_identical(
-    current = partial(fit, pred.var = "Gr_Liv_Area", n.trees = best_iter),
[…]

### 2020-02-23 inst/tinytest/test_pkg_e1071.R

@@ -1,3 +1,13 @@
+# Exits
+if (!requireNamespace("e1071", quietly = TRUE)) {
+  exit_file("Package e1071 missing")
+}
+
+# # Load required packages
+# suppressMessages({
+#   library(e1071)
+# })
+
 # Load Friedman benchmark data
 friedman1 <- readRDS("friedman.rds")$friedman1
 friedman2 <- readRDS("friedman.rds")$friedman2

### 2020-02-23 slowtests/slowtests-rjournal.R

@@ -132,7 +132,7 @@ cl <- makeCluster(4)  # use 4 cores
 registerDoParallel(cl)
 partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
         chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
-lattice::trellis.focus("legend", side = "right", clipp.off = TRUE, 
+lattice::trellis.focus("legend", side = "right", clipp.off = TRUE,
                        highlight = FALSE)
 grid::grid.text("ozone", 0.2, 1.05, hjust = 0.5, vjust = 1)
 lattice::trellis.unfocus()
@@ -195,7 +195,7 @@ grid.arrange(
 grid.arrange(
   partial(iris.svm, pred.var = "Petal.Width", prob = TRUE, plot = TRUE),
   partial(iris.svm, pred.var = "Petal.Length", prob = TRUE, plot = TRUE),
-  partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"), prob = TRUE, 
+  partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"), prob = TRUE,
           plot = TRUE),
   ncol = 3
 )
@@ -231,7 +231,7 @@ plotPartial(rm.ice, rug = TRUE, train = boston, alpha = 0.2)
 # Post-process rm.ice to obtain c-ICE curves
 rm.ice <- rm.ice %>%
   dplyr::group_by(yhat.id) %>%
-  dplyr::mutate(yhat.centered = yhat - first(yhat))
+  dplyr::mutate(yhat.centered = yhat - dplyr::first(yhat))
 
 # ICE curves with PDP
 p1 <- ggplot(rm.ice, aes(rm, yhat)) +
@@ -254,7 +254,7 @@ grid.arrange(p1, p2, ncol = 2)
 # p3 <- autoplot(pd.ice, alpha = 0.3)      # ICE curves: ggplot2 version
 # p4 <- autoplot(pd.cice, alpha = 0.3)     # c-ICE curves: ggplot2 version
 # grid.arrange(p1, p2, p3, p4, ncol = 2)
-# 
+#
 # # Try some other things
 # grid.arrange(
 #   autoplot(pd.ice, center = TRUE, alpha = 0.3),
@@ -308,7 +308,7 @@ grid.arrange(pdp.lstat, pdp.rm, pdp.lstat.rm, ncol = 3)
[…]

### 2020-04-13 README.Rmd
minor tweaks

@@
 [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/pdp)](https://cran.r-project.org/package=pdp)
+[![R build status](https://github.com/bgreenwell/pdp/workflows/R-CMD-check/badge.svg)](https://github.com/bgreenwell/pdp/actions)
 [![Build Status](https://travis-ci.org/bgreenwell/pdp.svg?branch=master)](https://travis-ci.org/bgreenwell/pdp)

### 2020-04-14 README.Rmd
rm old badges

--- 2019-10-28 README.Rmd
@@
 [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/pdp)](https://cran.r-project.org/package=pdp)
-[![Build Status](https://travis-ci.org/bgreenwell/pdp.svg?branch=master)](https://travis-ci.org/bgreenwell/pdp)
-[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/bgreenwell/pdp?branch=master&svg=true)](https://ci.appveyor.com/project/bgreenwell/pdp)
+[![R build status](https://github.com/bgreenwell/pdp/workflows/R-CMD-check/badge.svg)](https://github.com/bgreenwell/pdp/actions)
 [![Codecov test coverage](https://codecov.io/gh/bgreenwell/pdp/branch/master/graph/badge.svg)](https://codecov.io/gh/bgreenwell/pdp?branch=master)
-[![Downloads](http://cranlogs.r-pkg.org/badges/pdp)](http://cranlogs.r-pkg.org/badges/pdp)
 [![Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/pdp)](http://cranlogs.r-pkg.org/badges/grand-total/pdp)

### 2020-12-15 README.Rmd
fix tests

@@
 [![R build status](https://github.com/bgreenwell/pdp/workflows/R-CMD-check/badge.svg)](https://github.com/bgreenwell/pdp/actions)
+[![R-CMD-check](https://github.com/bgreenwell/pdp/workflows/R-CMD-check/badge.svg)](https://github.com/bgreenwell/pdp/actions)
 [![Codecov test coverage](https://codecov.io/gh/bgreenwell/pdp/branch/master/graph/badge.svg)](https://codecov.io/gh/bgreenwell/pdp?branch=master)

### 2020-12-15 inst/tinytest/test_exemplar.R

@@ -8,11 +8,7 @@ df1 <- data.frame(
 )
 
 # Data frame that pdp::exemplar() should return
-df2 <- data.frame(
-  x = 3L,
-  y = factor("c", levels = letters[1L:3L]),
-  z = Sys.Date() + 2
-)
+df2 <- data.frame(x = 3L, y = "c", z = Sys.Date() + 2)
 
 # Expectation(s)
 df3 <- exemplar(df1)

### 2021-08-13 rjournal/Untitled/greenwell.tex
rm yhat.id when pred.fun returns a singleton

--- 2020-02-23 rjournal/greenwell.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Front matter
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+% !TeX root = RJwrapper.tex
 \title{pdp: An R Package for Constructing Partial Dependence Plots}
-\author{by Brandon M. Greenwell}
+\author{by Author One}
 
@@
 
-\abstract{
-Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality).
-Unfortunately, it can be challenging to understand the results of such models and explain them to management.
-Partial dependence plots offer a simple solution.
-Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be more easily understood.
-These plots are especially useful in explaining the output from black box models.
-In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
+\abstract{%
+Complex nonparametric models---like neural networks, random forests, and
+support vector machines---are more common than ever in predictive
+analytics, especially when dealing with large observational databases
+that don't adhere to the strict assumptions imposed by traditional
+statistical techniques (e.g., multiple linear regression which assumes
+linearity, homoscedasticity, and normality). Unfortunately, it can be
+challenging to understand the results of such models and explain them to
+management. Partial dependence plots offer a simple solution. Partial
+dependence plots are low-dimensional graphical renderings of the
+prediction function \(\widehat{f}\left(\boldsymbol{x}\right)\) so that
+the relationship between the outcome and predictors of interest can be
+more easily understood. These plots are especially useful in explaining
+the output from black box models. In this paper, we introduce \pkg{pdp},
+a general R package for constructing partial dependence plots.
 }
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Introduction
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-\label{sec:intro}
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
 
-\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data.
-One of their goals was to find a housing value equation using data on median home values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see \citet[Table IV]{harrison-1978-hedonic} for a description of each variable.
-The data violate many classical assumptions like linearity, normality, and constant variance.
-Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$).
-Part of the payoff for there time and efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
+\textbf{Note:} This is an updated version of the original R Journal
+article.
+
+\citet{harrison-1978-hedonic} were among the first to analyze the
+well-known Boston housing data. One of their goals was to find a housing
+value equation using data on median home values from \(n = 506\) census
+tracts in the suburbs of Boston from the 1970 census; see
+\citet[Table IV]{harrison-1978-hedonic} for a description of each
+variable. The data violate many classical assumptions like linearity,
+normality, and constant variance. Nonetheless,
+\citeauthor{harrison-1978-hedonic}---using a combination of
+transformations, significance testing, and grid searches---were able to
+find a reasonable fitting model (\(R^2 = 0.81\)). Part of the payoff for
+there time and efforts was an interpretable prediction equation which is
+reproduced in Equation\textasciitilde{}\eqref{eqn:boston}.
 \begin{equation}
@@
 
-Nowadays, many supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy.
-(We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation since these algorithms typically do not produce simple prediction formulas like Equation~\eqref{eqn:boston}.
-These models can still provide insight into the data, but it is not in the form of simple equations.
-For example, quantifying predictor importance has become an essential task in the analysis of "big data", and many supervised learning algorithms, like tree-based methods, can naturally assign variable importance scores to all of the predictors in the training data.
+Nowadays, many supervised learning algorithms can fit the data
+automatically in seconds---typically with higher accuracy. (We will
+revisit the Boston housing data in
+Section\textasciitilde{}\ref{sec:boston}.) The downfall, however, is
+some loss of interpretation since these algorithms typically do not
+produce simple prediction formulas like
+Equation\textasciitilde{}\eqref{eqn:boston}. These models can still
+provide insight into the data, but it is not in the form of simple
+equations. For example, quantifying predictor importance has become an
+essential task in the analysis of ``big data'', and many supervised
+learning algorithms, like tree-based methods, can naturally assign
+variable importance scores to all of the predictors in the training
+data.
 
-While determining predictor importance is a crucial task in any supervised learning problem, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response.
-This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2001-greedy} for details.
-PDPs help visualize the relationship between a subset of the features (typically 1-3) and the response while accounting for the average effect of the other predictors in the model.
-They are particularly effective with black box models like random forests and support vector machines.
+While determining predictor importance is a crucial task in any
+supervised learning problem, ranking variables is only part of the story
+and once a subset of ``important'' features is identified it is often
+necessary to assess the relationship between them (or subset thereof)
+and the response. This can be done in many ways, but in machine learning
+it is often accomplished by constructing \dfn{partial dependence plots}
+(PDPs); see \citet{friedman-2001-greedy} for details. PDPs help
+visualize the relationship between a subset of the features (typically
+1-3) and the response while accounting for the average effect of the
+other predictors in the model. They are particularly effective with
+black box models like random forests and support vector machines.
 
-Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_p\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$.
-If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_c = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the "partial dependence" of the response on $\boldsymbol{z}_s$ is defined as
-\begin{equation}
+Let \(\boldsymbol{x} = \left\{x_1, x_2, \dots, x_p\right\}\) represent
+the predictors in a model whose prediction function is
+\(\widehat{f}\left(\boldsymbol{x}\right)\). If we partition
+\(\boldsymbol{x}\) into an interest set, \(\boldsymbol{z}_s\), and its
+compliment,
+\(\boldsymbol{z}_c = \boldsymbol{x} \setminus \boldsymbol{z}_s\), then
+the ``partial dependence'' of the response on \(\boldsymbol{z}_s\) is
+defined as \begin{equation}
 \label{eqn:avg_fun}
   f_s\left(\boldsymbol{z}_s\right) = E_{\boldsymbol{z}_c}\left[\widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)\right] = \int \widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)p_{c}\left(\boldsymbol{z}_c\right)d\boldsymbol{z}_c,
-\end{equation}
-where $p_{c}\left(\boldsymbol{z}_c\right)$ is the marginal probability density of $\boldsymbol{z}_c$: $p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s$.
-Equation~\eqref{eqn:avg_fun} can be estimated from a set of training data by
-\begin{equation}
+\end{equation} where \(p_{c}\left(\boldsymbol{z}_c\right)\) is the
+marginal probability density of \(\boldsymbol{z}_c\):
+\(p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s\).
+Equation\textasciitilde{}\eqref{eqn:avg_fun} can be estimated from a set
+of training data by \begin{equation}
 \label{eqn:pdf}
 \bar{f}_s\left(\boldsymbol{z}_s\right) = \frac{1}{n}\sum_{i = 1}^n\widehat{f}\left(\boldsymbol{z}_s,\boldsymbol{z}_{i, c}\right),
-\end{equation}
-where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, we average out the effects of all the other predictors in the model.
+\end{equation} where \(\boldsymbol{z}_{i, c}\)
+\(\left(i = 1, 2, \dots, n\right)\) are the values of
+\(\boldsymbol{z}_c\) that occur in the training sample; that is, we
+average out the effects of all the other predictors in the model.
 
-Constructing a PDP \eqref{eqn:pdf} in practice is rather straightforward.
-To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$.
-The partial dependence of the response on $x_1$ can be constructed as follows:
+Constructing a PDP \eqref{eqn:pdf} in practice is rather
+straightforward. To simplify, let \(\boldsymbol{z}_s = x_1\) be the
+predictor variable of interest with unique values
+\(\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}\). The partial dependence
+of the response on \(x_1\) can be constructed as follows:
 
@@
 \end{algorithm}
-Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records.
-Fortunately, the algorithm can be parallelized quite easily (more on this in Section~\ref{sec:computational}).
-It can also be easily extended to larger subsets of two or more features as well.
 
-Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm} \citep{gbm-pkg}, among others; these are limited in the sense that they only apply to the models fit using the respective package.
-For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"} and the \code{plot} function in \pkg{gbm} only applies to \code{"gbm"} objects.
-While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space.
-Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., generalized additive models and neural networks).
-However, to our knowledge, there is no general package for constructing PDPs in R.
-For example, PDPs for a conditional random forest as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively.
-The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for constructing PDPs that can be applied to several classes of fitted models.
+Algorithm\textasciitilde{}\ref{alg:pdp} can be quite computationally
+intensive since it involves \(k\) passes over the training records.
+Fortunately, the algorithm can be parallelized quite easily (more on
+this in Section\textasciitilde{}\ref{sec:computational}). It can also be
+easily extended to larger subsets of two or more features as well.
 
-The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}.
-According to \citeauthor{plotmo-pkg}, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level).
-These plots allow for up to two variables at a time.
-They are also less accurate than PDPs, but are faster to construct.
-For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs.
-As of \pkg{plotmo} version 3.3.0, there is now support for constructing PDPs, but it is not the default.
-The main difference is that \pkg{plotmo}, rather than applying step 1.
-(a)-(c) in Algorithm~\ref{alg:pdp}, accumulates all the data at once thereby reducing the number of internal calls to \code{predict}.
-The trade-off is a slight increase in speed at the expense of using more memory.
-So, why use the \pkg{pdp} package?
-As will be discussed in the upcoming sections, \pkg{pdp}:
+Limited implementations of Friedman's PDPs are available in packages
+\CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}
+\citep{gbm-pkg}, among others; these are limited in the sense that they
+only apply to the models fit using the respective package. For example,
+the \code{partialPlot} function in \pkg{randomForest} only applies to
+objects of class \code{"randomForest"} and the \code{plot} function in
+\pkg{gbm} only applies to \code{"gbm"} objects. While the
+\pkg{randomForest} implementation will only allow for a single
+predictor, the \pkg{gbm} implementation can deal with any subset of the
+predictor space. Partial dependence functions are not restricted to
+tree-based models; they can be applied to any supervised learning
+algorithm (e.g., generalized additive models and neural networks).
+However, to our knowledge, there is no general package for constructing
+PDPs in R. For example, PDPs for a conditional random forest as
+implemented by the \code{cforest} function in the \CRANpkg{party} and
+\CRANpkg{partykit} packages; see \citet{party-pkg} and
+\citet{partykit-pkg}, respectively. The \CRANpkg{pdp} \citep{pdp-pkg}
+package tries to close this gap by offering a general framework for
+constructing PDPs that can be applied to several classes of fitted
+models.
+
+The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to
+\pkg{pdp}. According to \citeauthor{plotmo-pkg}, \pkg{plotmo} constructs
+``a poor man's partial dependence plot.'' In particular, it plots a
+model's response when varying one or two predictors while holding the
+other predictors in the model constant (continuous features are fixed at
+their median value, while factors are held at their first level). These
+plots allow for up to two variables at a time. They are also less
+accurate than PDPs, but are faster to construct. For additive models
+(i.e., models with no interactions), these plots are identical in shape
+to PDPs. As of \pkg{plotmo} version 3.3.0, there is now support for
+constructing PDPs, but it is not the default. The main difference is
+that \pkg{plotmo}, rather than applying step 1. (a)-(c) in
+Algorithm\textasciitilde{}\ref{alg:pdp}, accumulates all the data at
+once thereby reducing the number of internal calls to \code{predict}.
+The trade-off is a slight increase in speed at the expense of using more
+memory. So, why use the \pkg{pdp} package? As will be discussed in the
+upcoming sections, \pkg{pdp}:
+
 \begin{itemize}
@@
 
-PDPs can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}.
-To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots---available in the \CRANpkg{ICEbox} package.
-ICE plots display the estimated relationship between the response and a predictor of interest for each observation.
-Consequently, the PDP for a predictor of interest can be obtained by averaging the corresponding ICE curves across all observations.
-In Section~\ref{sec:prediction}, it is shown how to obtain ICE curves using the \pkg{pdp} package.
-It is also possible to display the PDP for a single predictor with \pkg{ICEbox}; see \code{?ICEbox::plot.ice} for an example.
-\pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor.
-The ability to construct centered ICE (c-ICE) plots and derivative ICE (d-ICE) plots is also available in \pkg{ICEbox}; c-ICE plots help visualize heterogeneity in the modeled relationship between observations, and d-ICE plots help to explore interaction effects.
+Several additional packages have been developed in recent years that
+provide similar functionality to \pkg{pdp}, such as the \CRANpkg{iml}
+package \citep{iml-pkg}.
 
-Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model.
-For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots.
-\dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provide tabular and graphical displays for the terms in parametric models while holding all other predictors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots.
-However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).
+PDPs can be misleading in the presence of substantial interactions
+\citep{goldstein-peeking-2015}. To overcome this issue
+\citeauthor*{goldstein-peeking-2015} developed the concept of
+\dfn{individual conditional expectation} (ICE) plots---available in the
+\CRANpkg{ICEbox} package. ICE plots display the estimated relationship
+between the response and a predictor of interest for each observation.
+Consequently, the PDP for a predictor of interest can be obtained by
+averaging the corresponding ICE curves across all observations. In
+Section\textasciitilde{}\ref{sec:prediction}, it is shown how to obtain
+ICE curves using the \pkg{pdp} package. It is also possible to display
+the PDP for a single predictor with \pkg{ICEbox}; see
+\code{?ICEbox::plot.ice} for an example. \pkg{ICEbox} only allows for
+one variable at a time (i.e., no multivariate displays), though color
+can be used effectively to display information about an additional
+predictor. The ability to construct centered ICE (c-ICE) plots and
+derivative ICE (d-ICE) plots is also available in \pkg{ICEbox}; c-ICE
+plots help visualize heterogeneity in the modeled relationship between
+observations, and d-ICE plots help to explore interaction effects.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Constructing PDPs in R
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Many other techniques exist for visualizing relationships between the
+predictors and the response based on a fitted model. For example, the
+\CRANpkg{car} package \citep{fox-car-2011} contains many functions for
+constructing \dfn{partial-residual} and \dfn{marginal-model} plots.
+\dfn{Effect displays}, available in the \CRANpkg{effects} package
+\citep{fox-effects-2003}, provide tabular and graphical displays for the
+terms in parametric models while holding all other predictors at some
+constant value---similar in spirit to \pkg{plotmo}'s marginal model
+plots. However, these methods were designed for simpler parametric
+models (e.g., linear and generalized linear models), whereas
+\pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box
+models (although, they can be used for simple parametric models as
+well).
+
 \section{Constructing PDPs in R}
-\label{sec:boston}
 
-The \pkg{pdp} package is useful for constructing PDPs for many classes of fitted models in R.
-PDPs are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest.
-The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}.
-The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}.
-Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}.
-The two most important functions exported by \pkg{pdp} are:
+The \pkg{pdp} package is useful for constructing PDPs for many classes
+of fitted models in R. PDPs are especially useful for visualizing the
+relationships discovered by complex machine learning algorithms such as
+a random forest. The latest stable release is available from
+\href{https://cran.r-project.org/package=pdp}{CRAN}. The development
+version is located on GitHub: \url{https://github.com/bgreenwell/pdp}.
+Bug reports and suggestions are appreciated and should be submitted to
+\url{https://github.com/bgreenwell/pdp/issues}. The two most important
+functions exported by \pkg{pdp} are:
+
 \begin{itemize}
@@
 
-The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the fitted model and predictors are specified using the \code{object} and \code{pred.var} arguments, respectively---these are the only required arguments.
-If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function.
-The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is labeled \code{yhat}\footnote{There is one exception to this.
-When a function supplied via the \code{pred.fun} argument returns multiple predictions, the second to last and last columns will be labeled \code{yhat} and \code{yhat.id}, respectively (see Section~\ref{sec:prediction}).} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
-If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} (with fewer plotting options) and returns the PDP in the form of a \pkg{lattice} plot (i.e., a \code{"trellis"} object).
-\strong{Note:}  it is recommended to call \code{partial} with \code{plot = FALSE} and store the results; this allows for more flexible plotting, and the user will not have to waste time calling \code{partial} again if the default plot is not sufficient.
+The \code{partial} function evaluates the partial dependence
+\eqref{eqn:pdf} from a fitted model over a grid of predictor values; the
+fitted model and predictors are specified using the \code{object} and
+\code{pred.var} arguments, respectively---these are the only required
+arguments. If \code{plot = FALSE} (the default), \code{partial} returns
+an object of class \code{"partial"} which inherits from the class
+\code{"data.frame"}; put another way, by default, \code{partial} returns
+a data frame with an additional class that is recognized by the
+\code{plotPartial} function. The columns of the data frame are labeled
+in the same order as the features supplied to \code{pred.var}, and the
+last column is labeled
+\code{yhat}\footnote{There is one exception to this.
+When a function supplied via the \code{pred.fun} argument returns multiple predictions, the second to last and last columns will be labeled \code{yhat} and \code{yhat.id}, respectively (see Section~\ref{sec:prediction}).}
+and contains the values of the partial dependence function
+\(\bar{f}_s\left(\boldsymbol{z}_s\right)\). If \code{plot = TRUE}, then
+\code{partial} makes an internal call to \code{plotPartial} (with fewer
+plotting options) and returns the PDP in the form of a \pkg{lattice}
+plot (i.e., a \code{"trellis"} object). \strong{Note:} it is recommended
+to call \code{partial} with \code{plot = FALSE} and store the results;
+this allows for more flexible plotting, and the user will not have to
+waste time calling \code{partial} again if the default plot is not
+sufficient.
 
-The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options.
-For example, \code{plotPartial} makes it straight forward to add a LOESS smooth, or produce a 3-D surface instead of a false color level plot (the default).
-Of course, since the default output produced by \code{partial} is still a data frame, the user can easily use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance (see Section~\ref{sec:classification} and Section~\ref{sec:prediction} for examples).
+The \code{plotPartial} function can be used for displaying more advanced
+PDPs; it operates on objects of class \code{"partial"} and has many
+useful plotting options. For example, \code{plotPartial} makes it
+straight forward to add a LOESS smooth, or produce a 3-D surface instead
+of a false color level plot (the default). Of course, since the default
+output produced by \code{partial} is still a data frame, the user can
+easily use any plotting package he/she desires to visualize the
+results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance (see
+Section\textasciitilde{}\ref{sec:classification} and
+Section\textasciitilde{}\ref{sec:prediction} for examples).
 
-\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics.
-\pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}.
-\pkg{grid} graphics behave a little differently than traditional R graphics, and two points are worth making (see \code{?lattice} for more details):
+\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for
+its graphics. \pkg{lattice} itself is built on top of \CRANpkg{grid}
+\citep{grid-pkg}. \pkg{grid} graphics behave a little differently than
+traditional R graphics, and two points are worth making (see
+\code{?lattice} for more details):
+
 \begin{enumerate}
@@
 \end{enumerate}
-A consequence of the second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots.
-Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}.
-For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{grid}-based graphical objects on a single plot (these include graphics produced using \pkg{lattice} (hence, \pkg{pdp}) and \pkg{ggplot2}).
-This is demonstrated in multiple examples throughout this paper.
 
-Currently supported models are described in Table~\ref{tab:models}.
-In these cases, the user does not need to supply a prediction function (more on this in Section~\ref{sec:prediction}) or a value for the \code{type} argument (i.e., \code{"regression"} or \code{"classification"}).
-In other situations, the user may need to specify one or both of these arguments.
-This allows \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002}.% and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+A consequence of the second point is that the \code{par} function cannot
+be used to control the layout of multiple \pkg{lattice} (and hence
+\pkg{pdp}) plots. Simple solutions are available in packages
+\CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra}
+\citep{gridExtra-pkg}. For convenience, \pkg{pdp} imports the
+\code{grid.arrange} function from \pkg{gridExtra} which makes it easy to
+display multiple \pkg{grid}-based graphical objects on a single plot
+(these include graphics produced using \pkg{lattice} (hence, \pkg{pdp})
+and \pkg{ggplot2}). This is demonstrated in multiple examples throughout
+this paper.
+
+Currently supported models are described in
+Table\textasciitilde{}\ref{tab:models}. In these cases, the user does
+not need to supply a prediction function (more on this in
+Section\textasciitilde{}\ref{sec:prediction}) or a value for the
+\code{type} argument (i.e., \code{"regression"} or
+\code{"classification"}). In other situations, the user may need to
+specify one or both of these arguments. This allows \code{partial} to be
+flexible enough to handle many of the model types not listed in
+Table\textasciitilde{}\ref{tab:models}; for example, neural networks
+from the \CRANpkg{nnet} package \citep{venables-modern-2002}.\% and
+projection pursuit regression \citep{friedman-ppr-1981} using the
+\code{ppr} function in the \pkg{stats} package.
 
@@
 
-The \code{partial} function also supports objects of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}.
-This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}.
-An example is given in Section~\ref{sec:xgboost}.
+The \code{partial} function also supports objects of class
+\code{"train"} produced using the \code{train} function from the
+well-known \CRANpkg{caret} package \citep{caret-pkg}. This means that
+\code{partial} can be used with any classification or regression model
+that has been fit using \pkg{caret}'s \code{train} function; see
+\url{http://topepo.github.io/caret/available-models.html} for a current
+list of models supported by \pkg{caret}. An example is given in
+Section\textasciitilde{}\ref{sec:xgboost}.
 
-Another important argument to \code{partial} is \code{train}.
-If \code{train = NULL} (the default), \code{partial} tries to extract the original training data from the fitted model object.
-For objects that typically store a copy of the training data (e.g., objects of class \code{"BinaryTree"}, \code{"RandomForest"}, and \code{"train"}), this is straightforward.
-Otherwise, \code{partial} will attempt to extract the call stored in \code{object} (if available) and use that to evaluate the training data in the same environment from which \code{partial} was called.
-This can cause problems when, for example, the training data have been changed after fitting the model, but before calling \code{partial}.
-Hence, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}\footnote{For brevity, we ignore this option in most of the examples in this paper.}.
-If \code{train = NULL} and the training data can not be extracted from the fitted model, the user will be prompted with an informative error message (this will occur, for example, when using \code{partial} with \code{"ksvm"} and \code{"xgb.Booster"} objects):
+Another important argument to \code{partial} is \code{train}. If
+\code{train = NULL} (the default), \code{partial} tries to extract the
+original training data from the fitted model object. For objects that
+typically store a copy of the training data (e.g., objects of class
+\code{"BinaryTree"}, \code{"RandomForest"}, and \code{"train"}), this is
+straightforward. Otherwise, \code{partial} will attempt to extract the
+call stored in \code{object} (if available) and use that to evaluate the
+training data in the same environment from which \code{partial} was
+called. This can cause problems when, for example, the training data
+have been changed after fitting the model, but before calling
+\code{partial}. Hence, it is good practice to always supply the training
+data via the \code{train} argument in the call to
+\code{partial}\footnote{For brevity, we ignore this option in most of the examples in this paper.}.
+If \code{train = NULL} and the training data can not be extracted from
+the fitted model, the user will be prompted with an informative error
+message (this will occur, for example, when using \code{partial} with
+\code{"ksvm"} and \code{"xgb.Booster"} objects):
+
 \begin{example}
@@
 
-For illustration, we will use a corrected version of the Boston housing data analyzed in \citet{harrison-1978-hedonic}; the data are available in the \pkg{pdp} package (see \code{?pdp::boston} for details).
-We begin by loading the data and fitting a random forest with default tuning parameters and 500 trees:
-\begin{example}
+For illustration, we will use a corrected version of the Boston housing
+data analyzed in \citet{harrison-1978-hedonic}; the data are available
+in the \pkg{pdp} package (see \code{?pdp::boston} for details). We begin
+by loading the data and fitting a random forest with default tuning
+parameters and 500 trees:
+
+\begin{Schunk}
+\begin{Sinput}
+library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
+
 data(boston, package = "pdp")  # load the (corrected) Boston housing data
-library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
+
+# Fit a default random forest to the Boston housing data
 set.seed(101)  # for reproducibility
@@
 varImpPlot(boston.rf)  # Figure 1
-\end{example}
-The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89.
-The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}.
-Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}).
-The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.
+\end{Sinput}
+\begin{figure}[!htb]
 
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{boston_rf_vimp}
-  \caption{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}
-  \label{fig:plotmo_vs_partial}
-\end{figure}
+{\centering \includegraphics[width=1\linewidth]{greenwell_files/figure-latex/boston-vip-1}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Single predictor PDPs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Single predictor PDPs}
-
-As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor---the keywords here are "single predictor".
-For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
-\begin{example}
-partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
-\end{example}
-The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
-\begin{example}
-library(pdp)  # for partial, plotPartial, and grid.arrange functions
-partial(boston.rf, pred.var = "lstat", plot = TRUE)  # Figure 2 (left)
-\end{example}
-The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.
-
-For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame.
-This is illustrated in the example below which increases the line width, adds a LOESS smooth, and customizes the $y$-axis label.
-The result is displayed in the right side of Figure~\ref{fig:pd_lstat}.
-\strong{Note:} to encourage writing more readable code, the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported whenever \pkg{pdp} is loaded.
-\begin{example}
-# Figure 2 (right)
-boston.rf %>%  # the %>% operator is read as "and then"
-  partial(pred.var = "lstat") %>%
-  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
-\end{example}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{pd_lstat}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat} based on a random forest. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
-  \label{fig:pd_lstat}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Multi-predictor PDPs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Multi-predictor PDPs}
-\label{sec:multi}
-
-The benefit of using \code{partial} is threefold: (1) it is a flexible, generic function that can be used to obtain different kinds of PDPs for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
-For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
-The \code{grid.arrange} function is used to display three PDPs, which make use of various \code{plotPartial} options\footnote{See Section~\ref{sec:computational} for an example of how to add a label to the colorkey in these types of graphs.}, on the same graph.
-The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
-\begin{example}
-# Compute partial dependence data for lstat and rm
-pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
-
-# Default PDP
-pdp1 <- plotPartial(pd)
-
-# Add contour lines and use a different color palette
-rwb <- colorRampPalette(c("red", "white", "blue"))
-pdp2 <- plotPartial(pd, contour = TRUE, col.regions = rwb)
-
-# 3-D surface
-pdp3 <- plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
-                    colorkey = TRUE, screen = list(z = -20, x = -60))
-
-# Figure 3
-grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
-\end{example}
-\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.
-
-\begin{widefigure}[htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a random forest.
-\textit{Left}: Default plot.
-\textit{Middle}: With contour lines and a different color palette.
-\textit{Right}: Using a 3-D surface.}
-  \label{fig:pd_lstat_rm}
-\end{widefigure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Avoiding extrapolation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Avoiding extrapolation}
-\label{sec:extrapolation}
-
-It is not wise to draw conclusions from PDPs in regions outside the area of the training data.
-Here we describe two ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
-Rug displays are one-dimensional plots added to the axes.
-Both \code{partial} and \code{plotPartial} have a \code{rug} option that, when set to \code{TRUE}, will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and vertical axes.
-The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
-\begin{example}
-# Figure 4 (left)
-partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
-\end{example}
-
-In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on.
-When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted.
-Over interpreting the PDP outside of this region is considered extrapolation and is ill-advised.
-The right display in Figure~\ref{fig:partial_extrap} was produced using:
-\begin{example}
-# Figure 4 (right)
-partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
-\end{example}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{partial_extrap}
-  \caption{Examples of PDPs with the addition of a rug display (left) and a convex hull (right).}
-  \label{fig:partial_extrap}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Addressing computational concerns
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Addressing computational concerns}
-\label{sec:computational}
-
-Constructing PDPs can be quite computationally expensive\footnote{The exception is regression trees based on single-variable splits which can make use of the efficient weighted tree traversal method described in \citet{friedman-2001-greedy}, however, only the \pkg{gbm} package seems to make use of this approach; consequently, \pkg{pdp} can also exploit this strategy when used with \pkg{gbm} models (see \code{?partial} for details).} Several strategies are available to ease the computational burden in larger problems.
-For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!).
-We could get very reasonable results using a reduced number of points.
-Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}.
-Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument.
-To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; \code{grid.arrange} is used to display all three PDPs on the same graph, side by side.
-The results are displayed in Figure~\ref{fig:partial_manual}.
-\begin{example}
-# Figure 5
-grid.arrange(
-  partial(boston.rf, "rm", plot = TRUE),
-  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
-  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
-  ncol = 3
-)
-\end{example}
-
-\begin{widefigure}[htbp]
-  \centering
-  \includegraphics[width=0.8\linewidth]{partial_manual}
-  \caption{Partial dependence of \code{cmedv} on \code{rm}. \textit{Left}: Default plot. \textit{Middle}: Using a reduced grid size. \textit{Right}: Using a user-specified grid.}
-  \label{fig:partial_manual}
-\end{widefigure}
-
-The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops.
-This makes it easy to request progress bars (e.g., \code{progress = "text"}) or run \code{partial} in parallel.
-In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package.
-To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].
-
-To illustrate, we will use the Los Angeles ozone pollution data described in \citet{breiman-1985-estimating}.
-The data contain daily measurements of ozone concentration (\code{ozone}) along with eight meteorological quantities for 330 days in the Los Angeles basin in 1976.\footnote{The data are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.data}.
-Details, including variable information, are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.info}.} The following code chunk loads the data into R:
-\begin{example}
-ozone <- read.csv(paste0("http://statweb.stanford.edu/~tibs/ElemStatLearn/",
-                         "datasets/LAozone.data"), header = TRUE)
-\end{example}
-
-Next, we use the multivariate adaptive regression splines (MARS) algorithm introduced in \citet{friedman-1991-mars} to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
-\begin{example}
-library(earth)  # for earth function (i.e., MARS algorithm)
-ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
-summary(ozone.mars)
-\end{example}
-The MARS model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
-\begin{itemize}
-  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
-  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
-  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
-\end{itemize}
-To understand this interaction, we can use a PDP. However, since the partial dependence between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.
-
-Setting up a parallel backend is rather straightforward.
-To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
-\begin{example}
-library(doParallel)  # load the parallel backend
-cl <- makeCluster(4)  # use 4 workers
-registerDoParallel(cl)  # register the parallel backend
-\end{example}
-Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}.
-See \code{?plyr::adply} for details.}.
-This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
-The last three lines of code add a label to the colorkey.
-The result is displayed in Figure~\ref{fig:partial_par}.
-\strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
-\begin{example}
-partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
-        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
-stopCluster(cl)  # good practice
-
-# Add a label to the colorkey
-lattice::trellis.focus("legend", side = "right", clipp.off = TRUE, highlight = FALSE)
-grid::grid.text("ozone", x = 0.2, y = 1.05, hjust = 0.5, vjust = 1)
-lattice::trellis.unfocus()
-\end{example}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=0.8\linewidth]{partial_par}
-  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}.
-Since \code{dpg} is continuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
-  \label{fig:partial_par}
-\end{figure}
-
-It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display.
-The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels.
-If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle.
-The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}).
-Hence, it will be more effective to use categorical variables to define the panels in higher dimensional displays when possible.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Classification problems
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Classification problems}
-\label{sec:classification}
-
-Traditionally, for classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp.
-369---370]{hastie-elements-2009}.
-Suppose the response is categorical with $K$ levels, then for each class we compute
-\begin{equation}
-\label{eqn:avg-logit}
-f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{k = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
-\end{equation}
-where $p_k(x)$ is the predicted probability for the $k$-th class. Plotting $f_k(x)$ helps us understand how the log-odds for the $k$-th class depends on different subsets of the predictor variables.
-
-To illustrate, we consider Edgar Anderson's iris data from the \pkg{datasets} package.
-The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris: setosa, versicolor, and virginica.
-We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} package (the tuning parameters were determined using 5-fold cross-validation).
-\begin{example}
-library(e1071)  # for svm function
-iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
-                cost = 0.25, probability = TRUE)
-\end{example}
-\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set \code{probability = TRUE} in the call to \code{svm}.
-
-Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
-\begin{example}
-pd <- NULL
-for (i in 1:3) {
-  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, grid.resolution = 101, progress = "text")
-  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
 }
 
-# Figure 7
-library(ggplot2)
-ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = yhat, fill = yhat)) +
-  geom_tile() +
-  geom_contour(color = "white", alpha = 0.5) +
-  scale_fill_distiller(name = "Centered\nlogit", palette = "Spectral") +
-  theme_bw() +
-  facet_grid(~ Species)
-\end{example}
+\caption[Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees]{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}\label{fig:boston-vip}
+\end{figure}
+\end{Schunk}
 
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
-  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} for the iris data.}
-  \label{fig:partial_iris}
-\end{figure}
+The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) \(R^2\)
+of 0.89. The variable importance scores are displayed in
+Figure\textasciitilde{}\ref{fig:boston-vip}. Both plots indicate that
+the percentage of lower status of the population (\code{lstat}) and the
+average number of rooms per dwelling (\code{rm}) are highly associated
+with the median value of owner-occupied homes (\code{cmedv}). The
+question then arises, ``What is the nature of these associations?'' To
+help answer this, we can look at the partial dependence of \code{cmedv}
+on \code{lstat} and \code{rm}, both individually and together.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% User-defined prediction functions
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{User-defined prediction functions}
-\label{sec:prediction}
+\bibliography{greenwell.bib}
 
-PDPs are essentially just averaged predictions; this is apparent from step 1.
-(c) in Algorithm~\ref{alg:pdp}.
-Consequently, as pointed out by \citet{goldstein-peeking-2015}, strong heterogeneity can conceal the complexity of the modeled relationship between the response and predictors of interest.
-This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE plot procedure.
-
-With \code{partial} it is possible to replace the mean in step 1.
-(c) of Algorithm~\ref{alg:pdp} with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale.
-It is even possible to obtain ICE curves.
-This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0).
-This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}.
-The supplied prediction function must return either a single prediction or a vector of predictions.
-Returning the mean of all the predictions will result in the traditional PDP.
-Returning a vector of predictions (i.e., one for each observation) will result in a set of ICE curves.
-The examples below illustrate.
-
-Using the \code{pred.fun} argument, it is possible to obtain PDPs for classification problems on the probability scale.
-We just need to write a function that computes the predicted class probability of interest averaged across all observations.
-The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:classification} to extract the average predicted probability of belonging to the \code{Setosa} class.
-\begin{example}
-pred.prob <- function(object, newdata) {  # see ?predict.svm
-  pred <- predict(object, newdata, probability = TRUE)
-  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
-  mean(prob.setosa)
+\address{%
+Author One\\
+Affiliation\\%
+line 1\\ line 2\\
+%
+\url{https://journal.r-project.org}\\%
+\textit{ORCiD: \href{https://orcid.org/0000-0002-9079-593X}{0000-0002-9079-593X}}\\%
+\href{mailto:author1@work}{\nolinkurl{author1@work}}%
 }
-\end{example}
-Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}.
-The following chunk of code uses \code{pred.prob} to obtain PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale.
-The results are displayed in Figure~\ref{fig:partial_iris_prob}.
-\begin{example}
-# PDPs for Petal.Width and Petal.Length on the probability scale
-pdp.pw <- partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob,
-                  plot = TRUE)
-pdp.pl <- partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob,
-                  plot = TRUE)
-pdp.pw.pl <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                     pred.fun = pred.prob, plot = TRUE)
-
-# Figure 8
-grid.arrange(pdp.pw, pdp.pl, pdp.pw.pl, ncol = 3)
-\end{example}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
-  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probability of belonging to the setosa species.}
-  \label{fig:partial_iris_prob}
-\end{figure}
-
-For regression problems, the default prediction function is essentially
-\begin{example}
-pred.fun <- function(object, newdata) {
-  mean(predict(object, newdata), na.rm = TRUE)
-}
-\end{example}
-This corresponds to step step 1.
-(c) in Algorithm~\ref{alg:pdp}.
-Suppose we would like ICE curves instead.
-To accomplish this we need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata} (i.e., just remove the call to \code{mean} in \code{pred.fun}).
-The code snippet below illustrates this for the Boston housing example using the predictor \code{rm}.
-The result is displayed in Figure~\ref{fig:ice_boston}.
-\strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in the following code chunk, there will be one curve for each observation in \code{boston}.
-\begin{example}
-# Use partial to obtain ICE curves
-pred.ice <- function(object, newdata) predict(object, newdata)
-rm.ice <- partial(boston.rf, pred.var = "rm", pred.fun = pred.ice)
-
-# Figure 9
-plotPartial(rm.ice, rug = TRUE, train = boston, alpha = 0.3)
-\end{example}
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=0.8\linewidth]{ice_boston}
-  \caption{ICE curves depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. Each curve corresponds to a different observation.}
-  \label{fig:ice_boston}
-\end{figure}
-
-The curves in Figure~\ref{fig:ice_boston} indicate some heterogeneity in the fitted model (i.e., some of the curves depict the opposite relationship).
-Such heterogeneity can be easier to spot using c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}.
-Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process the output from \code{partial} to obtain c-ICE curves (similar to the construction of \dfn{raw change scores} \citep[pg.
-130]{fitzmaurice-2011-applied} for longitudinal data).
-This is shown below.
-\begin{example}
-# Post-process rm.ice to obtain c-ICE curves
-library(dplyr)  # for group_by and mutate functions
-rm.ice <- rm.ice %>%
-  group_by(yhat.id) %>%  # perform next operation within each yhat.id
-  mutate(yhat.centered = yhat - first(yhat))  # so each curve starts at yhat = 0
-\end{example}
-
-Since the PDP is just the average of the corresponding ICE curves, it is quite simple to display both on the same plot.
-This is easily accomplished using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together.
-The code snippet below plots the ICE curves and c-ICE curves, along with their averages, for the predictor \code{rm} in the Boston housing example.
-The results are displayed in Figure~\ref{fig:ice_cice_boston}.
-\begin{example}
-# ICE curves with their average
-p1 <- ggplot(rm.ice, aes(rm, yhat)) +
-  geom_line(aes(group = yhat.id), alpha = 0.2) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
-
-# c-ICE curves with their average
-p2 <- ggplot(rm.ice, aes(rm, yhat.centered)) +
-  geom_line(aes(group = yhat.id), alpha = 0.2) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
-
-# Figure 10
-grid.arrange(p1, p2, ncol = 2)
-\end{example}
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{ice_cice_boston}
-  \caption{ICE curves (black curves) and their average (red curve) depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example.
-\textit{Left}: Uncentered (here the red curve is just the traditional PDP).
-\textit{Right}: Centered.}
-  \label{fig:ice_cice_boston}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Using partial with xgboost
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Using partial with the XGBoost library}
-\label{sec:xgboost}
-
-To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool.
-XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable.
-The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}.
-It has been shown to be many times faster than the well-known \pkg{gbm} package.
-However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for constructing PDPs.
-Fortunately, the \pkg{pdp} package can be used to fill this gap.
-
-For illustration, we return to the Boston housing example.
-The code chunk below uses \pkg{caret} to tune an \pkg{xgboost} model using 10-fold cross-validation.
-(After loading \pkg{caret}, use \code{getModelInfo("xgbTree")} for information on tuning \pkg{xgboost} models.) \strong{Warning:} The following code chunk may take a few minutes to run.
-\begin{example}
-# Tune an XGBoost model using 10-fold cross-validation
-library(caret)  # functions related to classification and regression training
-set.seed(202)  # for reproducibility
-boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
-                    y = boston$cmedv, method = "xgbTree", metric = "Rsquared",
-                    trControl = trainControl(method = "cv", number = 10),
-                    tuneLength = 10)
-\end{example}
-The optimal model had a cross-validated $R^2$ of $0.902$ (use \code{print(boston.xgb\$bestTune)} to view the optimum tuning parameters).
-The next snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together.
-The results are displayed in Figure~\ref{fig:boston_xgb}.
-\begin{example}
-# PDPs for lstat and rm
-pdp.lstat <- partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE)
-pdp.rm <- partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE)
-pdp.lstat.rm <- partial(boston.xgb, pred.var = c("lstat", "rm"),
-                        plot = TRUE, chull = TRUE)
-
-# Figure 11
-grid.arrange(pdp.lstat, pdp.rm, pdp.lstat.rm, ncol = 3)
-\end{example}
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{boston_xgb}
-  \caption{PDPs for the top two most important variables in the Boston housing data using \pkg{xgboost}.
-Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
-  \label{fig:boston_xgb}
-\end{figure}
-%\code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}
-
-The \code{train} function creates objects of class \code{"train"}, whereas the \code{xgboost} function creates objects of class \code{"xgb.Booster"}.
-Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object, there is no need to supply it in the call to \code{partial} in this example.
-However, this is not the case when using the \pkg{xgboost} package directly.
-To illustrate, we fit the same model using the \code{xgboost} function with the optimum tuning parameters found previously using \pkg{caret}.
-%stored in \code{boston.xgb\$bestTune}.%
-\begin{example}
-library(xgboost)  # for xgboost function
-set.seed(203)  # for reproducibility
-boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
-                      label = boston$cmedv, objective = "reg:linear",
-                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
-                      colsample_bytree = 0.8, min_child_weight = 1,
-                      subsample = 0.9444444)
-\end{example}
-To use \code{partial} with \code{"xgb.Booster"} objects, we need to supply the original training data (minus the response) in the call to \code{partial}.
-The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown).
-(Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
-\begin{example}
-partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
-        train = subset(boston, select = -cmedv))
-\end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Summary
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Summary}
-
-PDPs can be used to graphically examine the dependence of the response on low cardinality subsets of the features, accounting for the average effect of the other predictors.
-In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package.
-We also briefly discussed related approaches available in other R packages.
-Suggestions to avoid extrapolation and high execution times were discussed and demonstrated via examples.
-
-In terms of future development, \pkg{pdp} can be expanded in a number of ways.
-For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.
-It would also be worthwhile to implement the partial dependence-based $H$-statistic \citep{friedman-2008-predictive} for assessing the strength of interaction between predictors.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Summary
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgments}
-
-The author would like to thank two anonymous reviewers and the Editor for their helpful comments
-and suggestions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Back matter
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliography{greenwell}
-
-\address{Brandon M. Greenwell\\
-  Infoscitex, a DCS Company\\
-  4027 Colonel Glenn Highway\\
-  Suite 210\\
-  Dayton, OH 45431-1672\\
-  United States of America\\}
-\email{greenwell.brandon@gmail.com}

### 2021-08-13 R/autoplot.R

@@ -1,5 +1,3 @@
-.data <- NULL
-
 #' Plotting Partial Dependence Functions
 #'
 #' Plots partial dependence functions (i.e., marginal effects) using
@@ -76,6 +74,8 @@
 #' @importFrom ggplot2 scale_fill_distiller scale_fill_viridis_c stat_summary
 #'
 #' @importFrom ggplot2 theme_bw xlab ylab
+#'
+#' @importFrom rlang .data
 #'
 #' @rdname autoplot.partial
 #'

### 2021-08-13 R/pardep.R

@@ -125,10 +125,14 @@ pardep <- function(object, pred.var, pred.grid, pred.fun, inv.link, ice, task,
       preds
     }
     len <- length(yhat) / nrow(pred.grid)  # FIXME: Is there a more robust way?
-    grid.id <- rep(seq_len(nrow(pred.grid)), each = len)
-    yhat.id <- rep(seq_len(len), times = nrow(pred.grid))
-    res <- data.frame(pred.grid[grid.id, ], "yhat" = yhat, "yhat.id" = yhat.id)
-    names(res) <- c(names(pred.grid), "yhat", "yhat.id")
+    if (len == 1L) {  # no need for `yhat.id` when `pred.fun()` returns a singleton
+      res <- cbind(pred.grid, "yhat" = yhat)
+    } else {  # multiple predictions per call to `pred.fun()`
+      grid.id <- rep(seq_len(nrow(pred.grid)), each = len)
+      yhat.id <- rep(seq_len(len), times = nrow(pred.grid))
+      res <- data.frame(pred.grid[grid.id, ], "yhat" = yhat, "yhat.id" = yhat.id)
+      names(res) <- c(names(pred.grid), "yhat", "yhat.id")
+    }
 
   }
 

### 2021-08-13 R/plotPartial.R

@@ -228,7 +228,7 @@ plot_ice_curves <- function(object, plot.pdp, center, pdp.col, pdp.lwd, pdp.lty,
     stats::as.formula(paste("yhat ~", names(object)[1L])), data = object,
     groups = object$yhat.id, type = plot.type, ...,
     panel = function(xx, yy, ...) {
-      panel.xyplot(xx, yy, col = "black", ...)
+      panel.xyplot(xx, yy, col = 1, ...)
       if (plot.pdp) {
         pd <- average_ice_curves(object)
         panel.xyplot(
@@ -240,10 +240,9 @@ plot_ice_curves <- function(object, plot.pdp, center, pdp.col, pdp.lwd, pdp.lty,
         if (is.null(train)) {
           stop("The training data must be supplied for rug display.")
         } else {
-          panel.rug(
-            stats::quantile(train[, names(object)[1L], drop = TRUE],
-                            probs = 0:10/10, na.rm = TRUE)
-          )
+          panel.rug(stats::quantile(train[, names(object)[1L], drop = TRUE],
+                                    probs = 0:10/10, na.rm = TRUE),
+                    col = 1)
         }
       }
 })
@@ -261,7 +260,7 @@ plot_one_predictor_pdp <- function(object, smooth, rug, train = NULL, ...) {
     xyplot(
       stats::as.formula(paste("yhat ~", names(object)[1L])), data = object,
       type = "l", ..., panel = function(xx, yy, ...) {
-        panel.xyplot(xx, yy, col = "black", ...)
+        panel.xyplot(xx, yy, col = 1, ...)
         if (smooth) {
           panel.loess(xx, yy, ...)
         }
@@ -269,10 +268,9 @@ plot_one_predictor_pdp <- function(object, smooth, rug, train = NULL, ...) {
           if (is.null(train)) {
             stop("The training data must be supplied for rug display.")
           } else {
-            panel.rug(
-              stats::quantile(train[, names(object)[1L], drop = TRUE],
-                              probs = 0:10/10, na.rm = TRUE)
-            )
+            panel.rug(stats::quantile(train[, names(object)[1L], drop = TRUE],
+                                      probs = 0:10/10, na.rm = TRUE),
+                      col = 1)
           }
         }
     })
@@ -318,7 +316,7 @@ plot_two_predictor_pdp <- function(
     # Draw a paneled line plot
     xyplot(
[…]

### 2021-08-13 R/utils.R

@@ -177,9 +177,9 @@ in.out <- function(bnd, x) {
   ind <- is.na(rowSums(bnd))
   bnd[ind, ] <- lowLim
   n <- nrow(bnd)
-  um <-.C(in_out, bx = as.double(bnd[, 1]), by = as.double(bnd[, 2]),
-          break.code = as.double(lowLim), x = as.double(x[, 1]),
-          y = as.double(x[, 2]), inside = as.integer(x[, 2] * 0),
-          nb = as.integer(n), n = as.integer(nrow(x)))
+  um <- .C(in_out, bx = as.double(bnd[, 1]), by = as.double(bnd[, 2]),
+           break.code = as.double(lowLim), x = as.double(x[, 1]),
+           y = as.double(x[, 2]), inside = as.integer(x[, 2] * 0),
+           nb = as.integer(n), n = as.integer(nrow(x)))
   as.logical(um$inside)
 }

### 2021-08-13 slowtests/slowtests-rjournal.R

@@ -238,12 +238,12 @@ rm.ice <- rm.ice %>%
 # ICE curves with PDP
 p1 <- ggplot(rm.ice, aes(rm, yhat)) +
   geom_line(aes(group = yhat.id), alpha = 0.2) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
+  stat_summary(fun = mean, geom = "line", col = "red", size = 1)
 
 # c-ICE curves with centered PDP
 p2 <- ggplot(rm.ice, aes(rm, yhat.centered)) +
   geom_line(aes(group = yhat.id), alpha = 0.2) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
+  stat_summary(fun = mean, geom = "line", col = "red", size = 1)
 
 # Figure 10
 grid.arrange(p1, p2, ncol = 2)

### 2021-08-13 inst/tinytest/test_partial.R
update tests

@@ -138,19 +138,19 @@ pd_pfun3 <- partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1,
                     pred.fun = pfun3, recursive = FALSE)  # no need for n.trees
 pd_pfun4 <- partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1,
                     pred.fun = pfun4, recursive = FALSE)  # no need for n.trees
-# expect_true(inherits(pd_pfun1, what = "partial"))
+expect_true(inherits(pd_pfun1, what = "partial"))
 expect_true(inherits(pd_pfun3, what = "ice"))
 expect_true(inherits(pd_pfun4, what = "ice"))
-# expect_identical(dim(pd_pfun1), target = c(2L, 2L))
+expect_identical(dim(pd_pfun1), target = c(2L, 2L))
 expect_identical(dim(pd_pfun3), target = c(as.integer(2 * nrow(ames)), 3L))
 expect_identical(dim(pd_pfun4), target = c(as.integer(2 * nrow(ames)), 3L))
 # expect_identical(pd_pfun4$yhat.id[1L], target = "row_1")
 
-# expect_identical(
-#   current = pd_pfun1,
-#   target = partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1,
-#                    recursive = FALSE, n.trees = best_iter)
-# )
+expect_identical(
+  current = pd_pfun1,
+  target = partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1,
+                   recursive = FALSE, n.trees = best_iter)
+)
 expect_error(  # cannot use `pred.fun` when `recursive = TRUE`
   partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1, pred.fun = pfun1)
 )

### 2021-08-13 slowtests/slowtests-in_memory.R

@@ -1,105 +1,0 @@
-# Load required packages
-library(ggplot2)
-library(pdp)
-
-# Set up data set(s) for training
-trn <- AmesHousing::make_ames()
-X <- subset(ames, select = -Sale_Price)
-y <- ames$Sale_Price
-
-# Helper functions
-par_dep <- function(object, ...) {
-  partial(object, pred.var = "lstat", ...)
-}
-timeit <- function(expr, digits = 3) {
-  secs <- system.time(expr)["elapsed"]
-  paste(round(secs, digits = digits), "seconds")
-}
-
-
-# Cross join benchmarks --------------------------------------------------------
-
-# Load required packages
-library(microbenchmark)
-
-# Run benchmark
-x <- data.frame(x = 1:51)
-y <- data.frame(y = rnorm(500), z = rnorm(500))
-mb <- microbenchmark(
-  merge(x, y, by = NULL),
-  pdp:::crossJoin(x, y),
-  times = 100L
-)
-autoplot(mb)
-
-
-# Package: Cubist --------------------------------------------------------------
-# library(Cubist)
-# set.seed(1550)
-# fit_Cubist <- cubist(X, y = y, committees = 10)
-# t1 <- timeit(pd1 <- partial(fit_Cubist, pred.var = "Gr_Liv_Area", gr = 10))
-# t2 <- timeit(pd2 <- partial(fit_Cubist, pred.var = "Gr_Liv_Area", gr = 51, in.memory = TRUE))
-# grid.arrange(
-#   autoplot(pd1) + ggtitle(t1),
-#   autoplot(pd2) + ggtitle(t2),
-#   nrow = 1
-# )
-
-# Package: earth ---------------------------------------------------------------
-library(earth)
-fit_earth <- earth(Sale_Price ~ ., data = trn, degree = 3)
-t1 <- timeit(pd1 <- par_dep(fit_earth, gr = 1000))
-t2 <- timeit(pd2 <- par_dep(fit_earth, in.memory = TRUE, gr = 1000))
-grid.arrange(
-  autoplot(pd1) + ggtitle(t1),
-  autoplot(pd2) + ggtitle(t2),
-  nrow = 1
-)
-
-
-# Package: randomForest --------------------------------------------------------
-library(randomForest)
-set.seed(1507)
[…]

### 2021-08-13 rjournal/Untitled/greenwell.Rmd

--- 2020-02-23 rjournal/greenwell.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Front matter
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{pdp: An R Package for Constructing Partial Dependence Plots}
-\author{by Brandon M. Greenwell}
-
-\maketitle
-
-\abstract{
+---
+title: "pdp: An R Package for Constructing Partial Dependence Plots"
+author:
+  # see ?rjournal_article for more information
+  - name: Author One
+    affiliation: Affiliation
+    address:
+    - line 1
+    - line 2
+    url: https://journal.r-project.org
+    orcid: 0000-0002-9079-593X
+    email:  author1@work
+abstract: >
   Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality).
@@
 In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
-}
+preamble: |
+  \usepackage[utf8]{inputenc}
+  \usepackage[T1]{fontenc}
+  \usepackage{RJournal}
+  \usepackage{amsmath,amssymb,array}
+  \usepackage{booktabs}
+  \usepackage{algorithm}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Introduction
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-\label{sec:intro}
+# per R journal requirement, the bib filename should be the same as the output 
+# tex file. Don't forget to rename the bib file and change this example value.
+bibliography: greenwell.bib
+
+output: rticles::rjournal_article
+---
+
+```{r setup, include=FALSE}
+library(magrittr)
+library(pdp)
+
+# Set global knitr chunk options
+knitr::opts_chunk$set(
+  cache = TRUE,
+  fig.width = 6,
+  fig.asp = 0.618,
+  out.width = "100%",
+  fig.align = "center",
+  fig.pos = "!htb",
+  message = FALSE,
+  warning = FALSE
+)
+```
+
+## Introduction
+
+**Note:** This is an updated version of the original R Journal article.
 
@@
 
+Several additional packages have been developed in recent years that provide similar functionality to \pkg{pdp}, such as the \CRANpkg{iml} package \citep{iml-pkg}.
+
 PDPs can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Constructing PDPs in R
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+<!----------------------------------------------------------------------------->
 \section{Constructing PDPs in R}
-\label{sec:boston}
+<!----------------------------------------------------------------------------->
 
@@
 Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}.
-For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{grid}-based graphical objects on a single plot (these include graphics produced using \pkg{lattice} (hence, \pkg{pdp}) and \pkg{ggplot2}).
+For example, the \code{grid.arrange} function from \pkg{gridExtra} makes it easy to display multiple \pkg{grid}-based graphical objects on a single plot (these include graphics produced using \pkg{lattice} (hence, \pkg{pdp}) and \pkg{ggplot2}).
 This is demonstrated in multiple examples throughout this paper.
@@
 We begin by loading the data and fitting a random forest with default tuning parameters and 500 trees:
-\begin{example}
+
+```{r boston-vip, fig.cap="Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees."}
+library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
+
 data(boston, package = "pdp")  # load the (corrected) Boston housing data
-library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
+
+# Fit a default random forest to the Boston housing data
 set.seed(101)  # for reproducibility
@@
 varImpPlot(boston.rf)  # Figure 1
-\end{example}
+```
+
 The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89.
-The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}.
+The variable importance scores are displayed in Figure~\ref{fig:boston-vip}.
 Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}).
@@
 
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{boston_rf_vimp}
-  \caption{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}
-  \label{fig:plotmo_vs_partial}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Single predictor PDPs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+<!----------------------------------------------------------------------------->
 \subsection{Single predictor PDPs}
+<!----------------------------------------------------------------------------->
 
@@
 For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
-\begin{example}
+
+```{r boston-lstat-pd, eval=FALSE}
 partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
-\end{example}
+```
+
 The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
-\begin{example}
+
+```{r boston-lstat-pd-plot, eval=FALSE}
 library(pdp)  # for partial, plotPartial, and grid.arrange functions
-partial(boston.rf, pred.var = "lstat", plot = TRUE)  # Figure 2 (left)
-\end{example}
-The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.
+partial(boston.rf, pred.var = "lstat", plot = TRUE)
+```
 
-For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame.
+The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package by default to produce all of its displays.
+
+```{r boston-lstat-pd-plot-ggplot2, eval=FALSE}
+partial(boston.rf, pred.var = "lstat", plot = TRUE, plot.engine = "ggplot2")
+```
+
+For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial()} (for \pkg{lattice} graphics) or \code{autoplot()} (for \pkg{ggplot2} graphics) functions on the resulting data frame.
 This is illustrated in the example below which increases the line width, adds a LOESS smooth, and customizes the $y$-axis label.
 The result is displayed in the right side of Figure~\ref{fig:pd_lstat}.
-\strong{Note:} to encourage writing more readable code, the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported whenever \pkg{pdp} is loaded.
-\begin{example}
-# Figure 2 (right)
+```{r boston-lstat-pd-plotPartial, eval=FALSE}
+library(magrittr)
+
 boston.rf %>%  # the %>% operator is read as "and then"
@@
   plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
-\end{example}
+```
 
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{pd_lstat}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat} based on a random forest. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
-  \label{fig:pd_lstat}
-\end{figure}
+```{r pd-lstat, fig.cap="Partial dependence of \\code{cmedv} on \code{lstat} based on a random forest.
+\\textit{Left}: Default plot.
+\\textit{Right}: Customized plot obtained using the \\code{plotPartial()} function."}
+p1 <- partial(boston.rf, pred.var = "lstat", plot = TRUE, plot.engine = "ggplot2")
+p2 <- boston.rf %>%  # the %>% operator is read as "and then"
+  partial(pred.var = "lstat") %>%
+  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
+gridExtra::grid.arrange(p1, p2, nrow = 1)
+```
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Multi-predictor PDPs
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+```{r boston-lstat-pdps, fig.cap="Partial dependence of \code{cmedv} on \code{lstat} based on a random forest using both \\pkg{lattice}- and \pkg{ggplot2}-based graphics with various options."}
+library(ggplot2)
+library(gridExtra)
+
+p1 <- partial(boston.rf, pred.var = "lstat", plot = TRUE, plot.engine = "ggplot2")
+p2 <- boston.rf %>%  # the %>% operator is read as "and then"
+  partial(pred.var = "lstat") %>%
+  autoplot(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
+
+# Display plots in a grid
+grid.arrange(p1, p2, nrow = 1)
+```
+
+<!----------------------------------------------------------------------------->
 \subsection{Multi-predictor PDPs}
-\label{sec:multi}
+<!----------------------------------------------------------------------------->
 
@@
 \end{widefigure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Avoiding extrapolation
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Avoiding extrapolation}
-\label{sec:extrapolation}
-
-It is not wise to draw conclusions from PDPs in regions outside the area of the training data.
-Here we describe two ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
-Rug displays are one-dimensional plots added to the axes.
-Both \code{partial} and \code{plotPartial} have a \code{rug} option that, when set to \code{TRUE}, will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and vertical axes.
-The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
-\begin{example}
-# Figure 4 (left)
-partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
-\end{example}
-
-In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on.
-When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted.
-Over interpreting the PDP outside of this region is considered extrapolation and is ill-advised.
-The right display in Figure~\ref{fig:partial_extrap} was produced using:
-\begin{example}
-# Figure 4 (right)
-partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
-\end{example}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{partial_extrap}
-  \caption{Examples of PDPs with the addition of a rug display (left) and a convex hull (right).}
-  \label{fig:partial_extrap}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Addressing computational concerns
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Addressing computational concerns}
-\label{sec:computational}
-
-Constructing PDPs can be quite computationally expensive\footnote{The exception is regression trees based on single-variable splits which can make use of the efficient weighted tree traversal method described in \citet{friedman-2001-greedy}, however, only the \pkg{gbm} package seems to make use of this approach; consequently, \pkg{pdp} can also exploit this strategy when used with \pkg{gbm} models (see \code{?partial} for details).} Several strategies are available to ease the computational burden in larger problems.
-For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!).
-We could get very reasonable results using a reduced number of points.
-Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}.
-Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument.
-To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; \code{grid.arrange} is used to display all three PDPs on the same graph, side by side.
-The results are displayed in Figure~\ref{fig:partial_manual}.
-\begin{example}
-# Figure 5
-grid.arrange(
-  partial(boston.rf, "rm", plot = TRUE),
-  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
-  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
-  ncol = 3
-)
-\end{example}
-
-\begin{widefigure}[htbp]
-  \centering
-  \includegraphics[width=0.8\linewidth]{partial_manual}
-  \caption{Partial dependence of \code{cmedv} on \code{rm}. \textit{Left}: Default plot. \textit{Middle}: Using a reduced grid size. \textit{Right}: Using a user-specified grid.}
-  \label{fig:partial_manual}
-\end{widefigure}
-
-The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops.
-This makes it easy to request progress bars (e.g., \code{progress = "text"}) or run \code{partial} in parallel.
-In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package.
-To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].
-
-To illustrate, we will use the Los Angeles ozone pollution data described in \citet{breiman-1985-estimating}.
-The data contain daily measurements of ozone concentration (\code{ozone}) along with eight meteorological quantities for 330 days in the Los Angeles basin in 1976.\footnote{The data are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.data}.
-Details, including variable information, are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.info}.} The following code chunk loads the data into R:
-\begin{example}
-ozone <- read.csv(paste0("http://statweb.stanford.edu/~tibs/ElemStatLearn/",
-                         "datasets/LAozone.data"), header = TRUE)
-\end{example}
-
-Next, we use the multivariate adaptive regression splines (MARS) algorithm introduced in \citet{friedman-1991-mars} to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
-\begin{example}
-library(earth)  # for earth function (i.e., MARS algorithm)
-ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
-summary(ozone.mars)
-\end{example}
-The MARS model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
-\begin{itemize}
-  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
-  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
-  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
-\end{itemize}
-To understand this interaction, we can use a PDP. However, since the partial dependence between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.
-
-Setting up a parallel backend is rather straightforward.
-To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
-\begin{example}
-library(doParallel)  # load the parallel backend
-cl <- makeCluster(4)  # use 4 workers
-registerDoParallel(cl)  # register the parallel backend
-\end{example}
-Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}.
-See \code{?plyr::adply} for details.}.
-This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
-The last three lines of code add a label to the colorkey.
-The result is displayed in Figure~\ref{fig:partial_par}.
-\strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
-\begin{example}
-partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
-        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
-stopCluster(cl)  # good practice
-
-# Add a label to the colorkey
-lattice::trellis.focus("legend", side = "right", clipp.off = TRUE, highlight = FALSE)
-grid::grid.text("ozone", x = 0.2, y = 1.05, hjust = 0.5, vjust = 1)
-lattice::trellis.unfocus()
-\end{example}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=0.8\linewidth]{partial_par}
-  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}.
-Since \code{dpg} is continuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
-  \label{fig:partial_par}
-\end{figure}
-
-It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display.
-The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels.
-If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle.
-The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}).
-Hence, it will be more effective to use categorical variables to define the panels in higher dimensional displays when possible.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Classification problems
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Classification problems}
-\label{sec:classification}
-
-Traditionally, for classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp.
-369---370]{hastie-elements-2009}.
-Suppose the response is categorical with $K$ levels, then for each class we compute
-\begin{equation}
-\label{eqn:avg-logit}
-f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{k = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
-\end{equation}
-where $p_k(x)$ is the predicted probability for the $k$-th class. Plotting $f_k(x)$ helps us understand how the log-odds for the $k$-th class depends on different subsets of the predictor variables.
-
-To illustrate, we consider Edgar Anderson's iris data from the \pkg{datasets} package.
-The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris: setosa, versicolor, and virginica.
-We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} package (the tuning parameters were determined using 5-fold cross-validation).
-\begin{example}
-library(e1071)  # for svm function
-iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
-                cost = 0.25, probability = TRUE)
-\end{example}
-\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set \code{probability = TRUE} in the call to \code{svm}.
-
-Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
-\begin{example}
-pd <- NULL
-for (i in 1:3) {
-  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, grid.resolution = 101, progress = "text")
-  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
-}
-
-# Figure 7
-library(ggplot2)
-ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = yhat, fill = yhat)) +
-  geom_tile() +
-  geom_contour(color = "white", alpha = 0.5) +
-  scale_fill_distiller(name = "Centered\nlogit", palette = "Spectral") +
-  theme_bw() +
-  facet_grid(~ Species)
-\end{example}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
-  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} for the iris data.}
-  \label{fig:partial_iris}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% User-defined prediction functions
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{User-defined prediction functions}
-\label{sec:prediction}
-
-PDPs are essentially just averaged predictions; this is apparent from step 1.
-(c) in Algorithm~\ref{alg:pdp}.
-Consequently, as pointed out by \citet{goldstein-peeking-2015}, strong heterogeneity can conceal the complexity of the modeled relationship between the response and predictors of interest.
-This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE plot procedure.
-
-With \code{partial} it is possible to replace the mean in step 1.
-(c) of Algorithm~\ref{alg:pdp} with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale.
-It is even possible to obtain ICE curves.
-This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0).
-This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}.
-The supplied prediction function must return either a single prediction or a vector of predictions.
-Returning the mean of all the predictions will result in the traditional PDP.
-Returning a vector of predictions (i.e., one for each observation) will result in a set of ICE curves.
-The examples below illustrate.
-
-Using the \code{pred.fun} argument, it is possible to obtain PDPs for classification problems on the probability scale.
-We just need to write a function that computes the predicted class probability of interest averaged across all observations.
-The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:classification} to extract the average predicted probability of belonging to the \code{Setosa} class.
-\begin{example}
-pred.prob <- function(object, newdata) {  # see ?predict.svm
-  pred <- predict(object, newdata, probability = TRUE)
-  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
-  mean(prob.setosa)
-}
-\end{example}
-Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}.
-The following chunk of code uses \code{pred.prob} to obtain PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale.
-The results are displayed in Figure~\ref{fig:partial_iris_prob}.
-\begin{example}
-# PDPs for Petal.Width and Petal.Length on the probability scale
-pdp.pw <- partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob,
-                  plot = TRUE)
-pdp.pl <- partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob,
-                  plot = TRUE)
-pdp.pw.pl <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                     pred.fun = pred.prob, plot = TRUE)
-
-# Figure 8
-grid.arrange(pdp.pw, pdp.pl, pdp.pw.pl, ncol = 3)
-\end{example}
-
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
-  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probability of belonging to the setosa species.}
-  \label{fig:partial_iris_prob}
-\end{figure}
-
-For regression problems, the default prediction function is essentially
-\begin{example}
-pred.fun <- function(object, newdata) {
-  mean(predict(object, newdata), na.rm = TRUE)
-}
-\end{example}
-This corresponds to step step 1.
-(c) in Algorithm~\ref{alg:pdp}.
-Suppose we would like ICE curves instead.
-To accomplish this we need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata} (i.e., just remove the call to \code{mean} in \code{pred.fun}).
-The code snippet below illustrates this for the Boston housing example using the predictor \code{rm}.
-The result is displayed in Figure~\ref{fig:ice_boston}.
-\strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in the following code chunk, there will be one curve for each observation in \code{boston}.
-\begin{example}
-# Use partial to obtain ICE curves
-pred.ice <- function(object, newdata) predict(object, newdata)
-rm.ice <- partial(boston.rf, pred.var = "rm", pred.fun = pred.ice)
-
-# Figure 9
-plotPartial(rm.ice, rug = TRUE, train = boston, alpha = 0.3)
-\end{example}
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=0.8\linewidth]{ice_boston}
-  \caption{ICE curves depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. Each curve corresponds to a different observation.}
-  \label{fig:ice_boston}
-\end{figure}
-
-The curves in Figure~\ref{fig:ice_boston} indicate some heterogeneity in the fitted model (i.e., some of the curves depict the opposite relationship).
-Such heterogeneity can be easier to spot using c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}.
-Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process the output from \code{partial} to obtain c-ICE curves (similar to the construction of \dfn{raw change scores} \citep[pg.
-130]{fitzmaurice-2011-applied} for longitudinal data).
-This is shown below.
-\begin{example}
-# Post-process rm.ice to obtain c-ICE curves
-library(dplyr)  # for group_by and mutate functions
-rm.ice <- rm.ice %>%
-  group_by(yhat.id) %>%  # perform next operation within each yhat.id
-  mutate(yhat.centered = yhat - first(yhat))  # so each curve starts at yhat = 0
-\end{example}
-
-Since the PDP is just the average of the corresponding ICE curves, it is quite simple to display both on the same plot.
-This is easily accomplished using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together.
-The code snippet below plots the ICE curves and c-ICE curves, along with their averages, for the predictor \code{rm} in the Boston housing example.
-The results are displayed in Figure~\ref{fig:ice_cice_boston}.
-\begin{example}
-# ICE curves with their average
-p1 <- ggplot(rm.ice, aes(rm, yhat)) +
-  geom_line(aes(group = yhat.id), alpha = 0.2) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
-
-# c-ICE curves with their average
-p2 <- ggplot(rm.ice, aes(rm, yhat.centered)) +
-  geom_line(aes(group = yhat.id), alpha = 0.2) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
-
-# Figure 10
-grid.arrange(p1, p2, ncol = 2)
-\end{example}
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{ice_cice_boston}
-  \caption{ICE curves (black curves) and their average (red curve) depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example.
-\textit{Left}: Uncentered (here the red curve is just the traditional PDP).
-\textit{Right}: Centered.}
-  \label{fig:ice_cice_boston}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Using partial with xgboost
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Using partial with the XGBoost library}
-\label{sec:xgboost}
-
-To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool.
-XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable.
-The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}.
-It has been shown to be many times faster than the well-known \pkg{gbm} package.
-However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for constructing PDPs.
-Fortunately, the \pkg{pdp} package can be used to fill this gap.
-
-For illustration, we return to the Boston housing example.
-The code chunk below uses \pkg{caret} to tune an \pkg{xgboost} model using 10-fold cross-validation.
-(After loading \pkg{caret}, use \code{getModelInfo("xgbTree")} for information on tuning \pkg{xgboost} models.) \strong{Warning:} The following code chunk may take a few minutes to run.
-\begin{example}
-# Tune an XGBoost model using 10-fold cross-validation
-library(caret)  # functions related to classification and regression training
-set.seed(202)  # for reproducibility
-boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
-                    y = boston$cmedv, method = "xgbTree", metric = "Rsquared",
-                    trControl = trainControl(method = "cv", number = 10),
-                    tuneLength = 10)
-\end{example}
-The optimal model had a cross-validated $R^2$ of $0.902$ (use \code{print(boston.xgb\$bestTune)} to view the optimum tuning parameters).
-The next snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together.
-The results are displayed in Figure~\ref{fig:boston_xgb}.
-\begin{example}
-# PDPs for lstat and rm
-pdp.lstat <- partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE)
-pdp.rm <- partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE)
-pdp.lstat.rm <- partial(boston.xgb, pred.var = c("lstat", "rm"),
-                        plot = TRUE, chull = TRUE)
-
-# Figure 11
-grid.arrange(pdp.lstat, pdp.rm, pdp.lstat.rm, ncol = 3)
-\end{example}
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{boston_xgb}
-  \caption{PDPs for the top two most important variables in the Boston housing data using \pkg{xgboost}.
-Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
-  \label{fig:boston_xgb}
-\end{figure}
-%\code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}
-
-The \code{train} function creates objects of class \code{"train"}, whereas the \code{xgboost} function creates objects of class \code{"xgb.Booster"}.
-Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object, there is no need to supply it in the call to \code{partial} in this example.
-However, this is not the case when using the \pkg{xgboost} package directly.
-To illustrate, we fit the same model using the \code{xgboost} function with the optimum tuning parameters found previously using \pkg{caret}.
-%stored in \code{boston.xgb\$bestTune}.%
-\begin{example}
-library(xgboost)  # for xgboost function
-set.seed(203)  # for reproducibility
-boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
-                      label = boston$cmedv, objective = "reg:linear",
-                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
-                      colsample_bytree = 0.8, min_child_weight = 1,
-                      subsample = 0.9444444)
-\end{example}
-To use \code{partial} with \code{"xgb.Booster"} objects, we need to supply the original training data (minus the response) in the call to \code{partial}.
-The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown).
-(Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
-\begin{example}
-partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
-        train = subset(boston, select = -cmedv))
-\end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Summary
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Summary}
-
-PDPs can be used to graphically examine the dependence of the response on low cardinality subsets of the features, accounting for the average effect of the other predictors.
-In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package.
-We also briefly discussed related approaches available in other R packages.
-Suggestions to avoid extrapolation and high execution times were discussed and demonstrated via examples.
-
-In terms of future development, \pkg{pdp} can be expanded in a number of ways.
-For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.
-It would also be worthwhile to implement the partial dependence-based $H$-statistic \citep{friedman-2008-predictive} for assessing the strength of interaction between predictors.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Summary
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgments}
-
-The author would like to thank two anonymous reviewers and the Editor for their helpful comments
-and suggestions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Back matter
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliography{greenwell}
-
-\address{Brandon M. Greenwell\\
-  Infoscitex, a DCS Company\\
-  4027 Colonel Glenn Highway\\
-  Suite 210\\
-  Dayton, OH 45431-1672\\
-  United States of America\\}
-\email{greenwell.brandon@gmail.com}

### 2021-09-08 R/pardep.R
keep prediction names when available

@@ -129,7 +129,11 @@ pardep <- function(object, pred.var, pred.grid, pred.fun, inv.link, ice, task,
       res <- cbind(pred.grid, "yhat" = yhat)
     } else {  # multiple predictions per call to `pred.fun()`
       grid.id <- rep(seq_len(nrow(pred.grid)), each = len)
-      yhat.id <- rep(seq_len(len), times = nrow(pred.grid))
+      yhat.id <- if (is.null(names(yhat))) {
+        rep(seq_len(len), times = nrow(pred.grid))
+      } else {
+        names(yhat)
+      }
       res <- data.frame(pred.grid[grid.id, ], "yhat" = yhat, "yhat.id" = yhat.id)
       names(res) <- c(names(pred.grid), "yhat", "yhat.id")
     }

### 2022-05-10 R/autoplot.R
prepping for CRAN release

@@ -71,7 +71,7 @@
 #'
 #' @importFrom ggplot2 geom_point geom_rug geom_smooth geom_tile ggplot ggtitle
 #'
-#' @importFrom ggplot2 scale_fill_distiller scale_fill_viridis_c stat_summary
+#' @importFrom ggplot2 scale_fill_viridis_c stat_summary
 #'
 #' @importFrom ggplot2 theme_bw xlab ylab
 #'

### 2022-05-10 R/exemplar.R

@@ -5,7 +5,9 @@
 #' corresponding median value and non-numeric columns are replaced with their
 #' most frequent value.
 #'
-#' @param object A data frame.
+#' @param object A data frame, matrix, or
+#' \code{\link[Matrix:dgCMatrix-class]{dgCMatrix}} (the latter two are
+#' supported by \code{\link[xgboost]{xgboost}}).
 #'
 #' @return A data frame with the same number of columns as \code{object} and a
 #' single row.
@@ -50,6 +52,7 @@ exemplar.matrix <- function(object) {
   res <- ceiling(apply(object, MARGIN = 2, FUN = stats::median))
   data.matrix(t(res))
 }
+
 
 #' @rdname exemplar
 #'

### 2022-05-10 R/plotPartial.R

@@ -436,7 +436,7 @@ plot_three_predictor_pdp <- function(
       )
     }
 
-    # Produce a paneled lineplot
+    # Produce a paneled line plot
     xyplot(
       form, data = object, type = "p", ...,
       panel = function(xx, yy, ...) {

### 2022-05-10 inst/tinytest/test_pkg_gbm.R

@@ -14,6 +14,7 @@ if (require(gbm, quietly = TRUE)) {
   fit1 <- gbm(
     formula = y ~ .,
     data = friedman1,
+    distribution = "gaussian",
     shrinkage = 0.01,
     n.trees = 500,
     interaction.depth = 2
@@ -34,9 +35,9 @@ if (require(gbm, quietly = TRUE)) {
     n.trees = 500,
     interaction.depth = 2
   )
-  iter1 <- gbm.perf(fit1, plot.it = FALSE, method = "OOB")
-  iter2 <- gbm.perf(fit2, plot.it = FALSE, method = "OOB")
-  iter3 <- gbm.perf(fit3, plot.it = FALSE, method = "OOB")
+  iter1 <- suppressMessages(gbm.perf(fit1, plot.it = FALSE, method = "OOB"))
+  iter2 <- suppressMessages(gbm.perf(fit2, plot.it = FALSE, method = "OOB"))
+  iter3 <- suppressMessages(gbm.perf(fit3, plot.it = FALSE, method = "OOB"))
 
   # Compute partial dependence, ICE, etc. for x.3
   pd1 <- partial(fit1, pred.var = "x.3", n.trees = iter1)

### 2022-05-10 rjournal/Untitled/greenwell.R

@@ -1,24 +1,0 @@
-## ----setup, include=FALSE-----------------------------------------------------
-# Set global knitr chunk options
-knitr::opts_chunk$set(
-  cache = TRUE,
-  fig.width = 6,
-  fig.asp = 0.618,
-  out.width = "100%",
-  fig.align = "center",
-  fig.pos = "!htb",
-  message = FALSE,
-  warning = FALSE
-)
-
-
-## ----boston-vip, fig.cap="Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees."----
-library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
-
-data(boston, package = "pdp")  # load the (corrected) Boston housing data
-
-# Fit a default random forest to the Boston housing data
-set.seed(101)  # for reproducibility
-boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
-varImpPlot(boston.rf)  # Figure 1
-

### 2022-05-10 slowtests/slowtests-xgboost.R

@@ -60,8 +60,8 @@ parDepPlot <- function(object, train, ...) {
     theme_light()
 }
 
-# Try all nine combnations (should all look exactly the same!)
-grid.arrange(
+# Try all nine combinations (should all look exactly the same!)
+gridExtra::grid.arrange(
   parDepPlot(bst.matrix, train = X),
   parDepPlot(bst.matrix, train = X.matrix),
   parDepPlot(bst.matrix, train = X.dgCMatrix),

### 2022-05-10 README.Rmd
new release :bookmark:

--- 2020-04-14 README.Rmd
@@
   md_document:
-    variant: markdown_github
+    variant: gfm
 ---
@@
 
-[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/pdp)](https://cran.r-project.org/package=pdp)
-[![R build status](https://github.com/bgreenwell/pdp/workflows/R-CMD-check/badge.svg)](https://github.com/bgreenwell/pdp/actions)
-[![Codecov test coverage](https://codecov.io/gh/bgreenwell/pdp/branch/master/graph/badge.svg)](https://codecov.io/gh/bgreenwell/pdp?branch=master)
-[![Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/pdp)](http://cranlogs.r-pkg.org/badges/grand-total/pdp)
-[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
+<!-- badges: start -->
+[![CRAN status](https://www.r-pkg.org/badges/version/investr)](https://CRAN.R-project.org/package=investr)
+[![R-CMD-check](https://github.com/bgreenwell/pdp/workflows/R-CMD-check/badge.svg)](https://github.com/bgreenwell/pdp/actions)
+[![Codecov test coverage](https://codecov.io/gh/bgreenwell/pdp/branch/master/graph/badge.svg)](https://app.codecov.io/gh/bgreenwell/pdp?branch=master)
+[![Total Downloads](https://cranlogs.r-pkg.org/badges/grand-total/pdp)](https://cranlogs.r-pkg.org/badges/grand-total/pdp)
+<!-- badges: end -->
 

### 2022-05-27 R/pardep.R
make progress arg logical; fix URLs

@@ -7,8 +7,8 @@ pardep <- function(object, pred.var, pred.grid, pred.fun, inv.link, ice, task,
                    which.class, logit, train, progress, parallel, paropts, ...) {
 
   # Disable progress bar for parallel execution
-  if (progress != "none" && isTRUE(parallel)) {
-    progress <- "none"
+  if (isTRUE(progress) && isTRUE(parallel)) {
+    progress <- FALSE
     warning("progress bars are disabled whenever `parallel = TRUE`.",
             call. = FALSE, immediate. = TRUE)
   }
@@ -25,7 +25,7 @@ pardep <- function(object, pred.var, pred.grid, pred.fun, inv.link, ice, task,
 
     # Partial dependence: regression case
     if (task == "regression" && isFALSE(ice)) {
-      if (progress != "none") {
+      if (isTRUE(progress)) {
         pb <- utils::txtProgressBar(min = 0, max = nrow(pred.grid), style = 3)
       }
       yhat <- foreach(i = seq_len(nrow(pred.grid)), .combine = "c",
@@ -34,7 +34,7 @@ pardep <- function(object, pred.var, pred.grid, pred.fun, inv.link, ice, task,
         temp[, pred.var] <- pred.grid[i, pred.var]
         preds <- mean(get_predictions(object, newdata = temp,
                                       inv.link = inv.link, ...), na.rm = TRUE)
-        if (progress != "none") {
+        if (isTRUE(progress)) {
           utils::setTxtProgressBar(pb, value = i)
         }
         preds
@@ -44,7 +44,7 @@ pardep <- function(object, pred.var, pred.grid, pred.fun, inv.link, ice, task,
 
     # Partial dependence: classification case
     if (task == "classification" && isFALSE(ice)) {
-      if (progress != "none") {
+      if (isTRUE(progress)) {
         pb <- utils::txtProgressBar(min = 0, max = nrow(pred.grid), style = 3)
       }
       yhat <- foreach(i = seq_len(nrow(pred.grid)), .combine = "c",
@@ -54,7 +54,7 @@ pardep <- function(object, pred.var, pred.grid, pred.fun, inv.link, ice, task,
         preds <- mean(get_probs(object, newdata = temp,
                                 which.class = which.class, logit = logit, ...),
                       na.rm = TRUE)
-        if (progress != "none") {
+        if (isTRUE(progress)) {
[…]

### 2022-05-27 R/partial.R

@@ -35,15 +35,15 @@
 #' \code{FALSE}. See Goldstein et al. (2014) for details.
 #'
 #' @param approx Logical indicating whether or not to compute a faster, but
-#' approximate, marginal effect plot (similar in spirit to
-#' \code{\link[plotmo]{plotmo}}). If \code{TRUE}, then `partial()` will compute
+#' approximate, marginal effect plot (similar in spirit to the
+#' \strong{plotmo} package). If \code{TRUE}, then \code{partial()} will compute
 #' predictions across the predictors specified in \code{pred.var} while holding
 #' the other predictors constant (a "poor man's partial dependence" function as
 #' Stephen Milborrow, the author of \code{\link[plotmo]{plotmo}}, puts it).
 #' Default is \code{FALSE}. Note this works with \code{ice = TRUE} as well.
 #' WARNING: This option is currently experimental. Use at your own risk. It is
 #' possible (and arguably safer) to do this manually by passing a specific
-#' "exemplar" observation to the train argument and specifying `pred.grid`
+#' "exemplar" observation to the train argument and specifying \code{pred.grid}
 #' manually.
 #'
 #' @param quantiles Logical indicating whether or not to use the sample
@@ -146,10 +146,8 @@
 #' in \code{pred.grid} has the correct class, levels, etc. Default is
 #' \code{TRUE}.
 #'
-#' @param progress Character string giving the name of the progress bar to use
-#' while constructing the partial dependence function. See
-#' \code{\link[plyr]{create_progress_bar}} for details. Default is
-#' \code{"none"}.
+#' @param progress Logical indicating whether or not to display a text-based
+#' progress bar. Default is \code{FALSE}.
 #'
 #' @param parallel Logical indicating whether or not to run \code{partial} in
 #' parallel using a backend provided by the \code{foreach} package. Default is
@@ -233,7 +231,7 @@
 #'
 #' # The partial function allows for multiple predictors
 #' partial(boston.rf, pred.var = c("lstat", "rm"), grid.resolution = 40,
-#'         plot = TRUE, chull = TRUE, progress = "text")
+#'         plot = TRUE, chull = TRUE, progress = TRUE)
 #'
 #' # The plotPartial function offers more flexible plotting
 #' pd <- partial(boston.rf, pred.var = c("lstat", "rm"), grid.resolution = 40)
@@ -264,11 +262,11 @@
 #'
[…]

### 2022-05-27 inst/tinytest/test_partial.R

@@ -40,9 +40,9 @@ best_iter <- gbm::gbm.perf(fit, method = "cv")
 
 # Partial dependence w/wo setting approx = TRUE
 pd1 <- partial(fit, pred.var = "Gr_Liv_Area", n.trees = best_iter,
-               recursive = FALSE, progress = "progress")
+               recursive = FALSE, progress = TRUE)
 pd2 <- partial(fit, pred.var = "Gr_Liv_Area", approx = TRUE,
-               recursive = FALSE, progress = "progress",
+               recursive = FALSE, progress = TRUE,
                n.trees = best_iter)
 
 # Basic expectation(s)
@@ -92,7 +92,7 @@ expect_error(partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid2,
 
 # No progress bars or parallel processing for GBMs when `recursive = TRUE`
 expect_message(partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1,
-                       n.trees = best_iter, progress = "text"))
+                       n.trees = best_iter, progress = TRUE))
 expect_error(partial(fit, pred.var = "Gr_Liv_Area", pred.grid = grid1,
                      n.trees = best_iter, parallel = TRUE))
 

### 2022-05-27 tools/pdp-logo.R

@@ -27,7 +27,7 @@ rescale <- function(x, a, b) {
 
 # Partial dependence of cmedv on lstat and rm
 pd <- partial(rfo, pred.var = c("lstat", "rm"), chull = FALSE,
-              progress = "text", grid.resolution = 100)
+              progress = TRUE, grid.resolution = 100)
 
 # Boundaries of the hexagon
 hex <- data.frame(x = 1.35 * 1 * c(-sqrt(3) / 2, 0, rep(sqrt(3) / 2, 2), 0,

### 2022-05-27 R/partial.R
remove URL to plotmo

@@ -39,7 +39,7 @@
 #' \strong{plotmo} package). If \code{TRUE}, then \code{partial()} will compute
 #' predictions across the predictors specified in \code{pred.var} while holding
 #' the other predictors constant (a "poor man's partial dependence" function as
-#' Stephen Milborrow, the author of \code{\link[plotmo]{plotmo}}, puts it).
+#' Stephen Milborrow, the author of \strong{plotmo} package), puts it).
 #' Default is \code{FALSE}. Note this works with \code{ice = TRUE} as well.
 #' WARNING: This option is currently experimental. Use at your own risk. It is
 #' possible (and arguably safer) to do this manually by passing a specific

### 2022-05-27 slowtests/slowtests-rjournal.R

@@ -29,7 +29,7 @@
 # Load required packages
 library(caret)
 library(doParallel)
-# library(dplyr)  # use dplyr:: instead to avoid conflictions with plyr!
+# library(dplyr)  # use dplyr:: instead to avoid conflicts with plyr!
 library(e1071)
 library(earth)
 library(ggplot2)
@@ -163,7 +163,7 @@ iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
 pd <- NULL
 for (i in 1:3) {
   tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, progress = "text",
+                 which.class = i, progress = TRUE,
                  grid.resolution = 101)
   pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
 }

### 2022-05-27 R/partial.R
remove URL to plotmo

@@ -39,7 +39,7 @@
 #' \strong{plotmo} package). If \code{TRUE}, then \code{partial()} will compute
 #' predictions across the predictors specified in \code{pred.var} while holding
 #' the other predictors constant (a "poor man's partial dependence" function as
-#' Stephen Milborrow, the author of \strong{plotmo} package), puts it).
+#' Stephen Milborrow, the author of \strong{plotmo}, puts it).
 #' Default is \code{FALSE}. Note this works with \code{ice = TRUE} as well.
 #' WARNING: This option is currently experimental. Use at your own risk. It is
 #' possible (and arguably safer) to do this manually by passing a specific
